---
title: "Figure 3: ISC benchmarking results"
author: "Josep Garnica <josep.garnicacaparros@unil.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# Overview

This document generates the benchmarking results figure showing:

1. **Dot plot visualization**: Dot size represents scaled scores (0-100%), color represents ranking
2. **Overall score**: Weighted scores across all tasks
3. **Comprehensive table**: Scores for each method across all tasks and resources

**Tasks evaluated (in order):**

1. Signal degradation (sensitivity to cell type signal disruption)
2. Over-partitioning (sensitivity to artificial cell population splitting)
3. Granularity (robustness to annotation granularity)
4. Complexity (robustness to cellular complexity)
5. Number of samples (robustness to dataset size - samples)
6. Number of cells (robustness to dataset size - cells)
7. Batch effects (robustness to technical differences)
8. Perturbation effects (robustness to biological perturbations)

**Note:** Pseudobulk:Pearson and average.similarity are excluded from plots (but included in tables).

# Setup

```{r setup, warning=FALSE, message=FALSE}
# Load packages
library(dplyr)
library(tibble)
library(BiocParallel)
library(tidyr)
library(ggplot2)
library(stringr)
library(purrr)
library(data.table)
library(openxlsx)
library(patchwork)
library(funkyheatmap)
library(scTypeEval)

# Working directory
# Render from the repository root so all paths are relative and reproducible.
wd <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
knitr::opts_knit$set(root.dir = wd)

# Configuration
CONFIG <- list(
  group = "celltype",  # Options: "celltype" or "ident"
  onlyCell = TRUE,  # For Ncells task, keep only perturbed cell types
  benchmarking_scores = TRUE,
  n_workers = 7
)
```

Benchmarking utils
```{r, eval = T}
source(system.file("benchmarking", "assays_utils.R", package = "scTypeEval"))
source(system.file("benchmarking", "Metrics_benchmarking.R", package = "scTypeEval"))
```


# Helper Functions

```{r helper-functions}
#' Min-Max scaling to [0, 1]
#' @param x Numeric vector
#' @return Scaled vector
minmax_scale <- function(x) {
  x <- as.numeric(x)
  if (all(is.na(x))) return(x)
  rng <- range(x, na.rm = TRUE)
  if (diff(rng) == 0) return(rep(0.5, length(x)))
  (x - rng[1]) / diff(rng)
}

#' Calculate ratio between single and multi-batch consistency
#' @param mb Multi-batch identifier
#' @param df Data frame with consistency measures
#' @return Data frame with consistency ratios
calculate_batch_ratio <- function(mb, df) {
  singles <- str_split(mb, "-", simplify = TRUE) %>% as.vector()
  
  r <- df %>%
    filter(batch %in% c(mb, singles)) %>%
    group_by(dissimilarity_method, consistency.metric,
             celltype, dataset, ident) %>%
    summarise(
      multi_measure = measure[batch == mb],
      single_mean   = mean(measure[batch %in% singles], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    filter(!is.na(multi_measure), !is.na(single_mean)) %>%
    mutate(
      batch = mb,
      drop = single_mean - multi_measure,
      score_celltype = if_else(drop < 0, 0, drop),
      score_celltype = 1 - score_celltype
    )
  return(r)
}

#' Standardize method names
#' @param df Data frame
#' @return Data frame with standardized method names
standardize_methods <- function(df) {
  # Handle both dissimilarity_method and dissimilarity.method column names
  if ("dissimilarity.method" %in% colnames(df) &&
      !"dissimilarity_method" %in% colnames(df)) {
    df <- df %>% rename(dissimilarity_method = dissimilarity.method)
  }
  
    if (!"consistency.metric" %in% colnames(df)) {
    df <- df %>% rename(consistency.metric = method)
  }
  
  df %>%
    mutate(
      consistency.metric = gsub("Orbital\\.medoid", "Nearest.medoid", consistency.metric),
      consistency.metric = gsub("ward\\.PropMatch", "Clustering.agreement", consistency.metric),
      dissimilarity_method = gsub("BestHit", "RecipClassif", dissimilarity_method),
      method_type = paste(consistency.metric, dissimilarity_method, sep = " | ")
    )
}

#' Filter metrics for visualization
#' @param df Data frame
#' @param for_plot Logical, if TRUE removes correlated metrics
#' @return Filtered data frame
filter_metrics <- function(df, for_plot = FALSE) {
  if (for_plot) {
    df <- df %>%
      filter(
        !grepl("Pseudobulk.*Pearson", method_type),
        !grepl("Average.similarity", method_type)
      )
  }
  return(df)
}

#' Compute residual (main) effects for consistency metrics and dissimilarity methods
#'
#' For each task (assay), computes the marginal mean score for each consistency metric
#' and dissimilarity method across all observed metric|dissimilarity combinations,
#' then subtracts the assay grand mean to obtain a centered residual effect.
#'
#' @param task_summary_df Output of the `prepare-viz-data` chunk (must include: method_type, assay, mean_score)
#' @param for_plot If TRUE, applies `filter_metrics(..., for_plot=TRUE)` before computing residuals
#' @return Tibble with per-assay residual effects and ranks
compute_residual_effects <- function(task_summary_df, for_plot = FALSE) {
  df <- task_summary_df
  if (for_plot) {
    df <- filter_metrics(df, for_plot = TRUE)
  }

  df <- df %>%
    mutate(method_type = as.character(method_type)) %>%
    tidyr::separate(
      method_type,
      into = c("consistency_metric", "dissimilarity_method"),
      sep = "[|]",
      remove = FALSE,
      fill = "right"
    )

  assay_grand_mean <- df %>%
    group_by(assay) %>%
    summarise(grand_mean = mean(mean_score, na.rm = TRUE), .groups = "drop")

  metric_effects <- df %>%
    group_by(assay, consistency_metric) %>%
    summarise(
      mean_score = mean(mean_score, na.rm = TRUE),
      n_combos = dplyr::n(),
      .groups = "drop"
    ) %>%
    left_join(assay_grand_mean, by = "assay") %>%
    mutate(
      component = "Consistency metric",
      level = consistency_metric,
      residual_effect = mean_score - grand_mean
    ) %>%
    select(assay, component, level, mean_score, residual_effect, n_combos)

  dissim_effects <- df %>%
    group_by(assay, dissimilarity_method) %>%
    summarise(
      mean_score = mean(mean_score, na.rm = TRUE),
      n_combos = dplyr::n(),
      .groups = "drop"
    ) %>%
    left_join(assay_grand_mean, by = "assay") %>%
    mutate(
      component = "Dissimilarity method",
      level = dissimilarity_method,
      residual_effect = mean_score - grand_mean
    ) %>%
    select(assay, component, level, mean_score, residual_effect, n_combos)

  bind_rows(metric_effects, dissim_effects) %>%
    group_by(assay, component) %>%
    mutate(
      rank_residual = rank(-residual_effect, ties.method = "average"),
      rank_mean = rank(-mean_score, ties.method = "average")
    ) %>%
    ungroup()
}
```

# Data Loading

## 1. Resources Data

```{r load-resources}
message("Loading resources data...")

li.fi <- list.files(file.path(wd, "results/isc/resources/output"),
                    recursive = TRUE,
                    pattern = "[.]rds",
                    full.names = TRUE)

resources_data <- lapply(li.fi, readRDS) %>%
  data.table::rbindlist() %>%
  standardize_methods() %>%
  mutate(
    `Duration (ms)` = duration * 1e3,
    `Peak memory (MB)` = memory_usage_MB
  )

message("✓ Resources data loaded: ", nrow(resources_data), " rows")
```

## 2. Benchmarking Tasks Data

```{r load-benchmarking, warning=F}
message("Loading benchmarking tasks data...")

li.fi <- list.files(file.path(wd, "results/isc/tasks/output"),
                    pattern = "rds$",
                    recursive = TRUE,
                    full.names = TRUE) %>%
  grep("summary", ., value = TRUE, invert = TRUE)

tasks_data <- bplapply(
  li.fi,
  BPPARAM = MulticoreParam(workers = CONFIG$n_workers, progressbar = TRUE),
  function(fi) {
    df0 <- readRDS(fi)
    if (is.null(df0)) {
      message("⚠️ Empty file: ", fi)
      return(NULL)
    }
    
    if (length(unique(df0$rate)) < 2) return(NULL)
    
    dataset <- str_remove(strsplit(fi, "/")[[1]][9], "_\\d+_\\d+$")
    
    # Infer assay type from filename
    assay_type <- case_when(
      grepl("Missclassific", fi, ignore.case = TRUE) ~ "Signal-degradation",
      grepl("NSamples", fi, ignore.case = TRUE) ~ "Number-samples",
      grepl("NCell", fi, ignore.case = TRUE) ~ "Number-cells",
      grepl("Nct", fi, ignore.case = TRUE) ~ "Complexity",
      grepl("mergeCT", fi, ignore.case = TRUE) ~ "Granularity",
      grepl("SplitCelltype", fi, ignore.case = TRUE) ~ "Overpartitioning",
      TRUE ~ "Unknown"
    )
    
    # Filter for specific cell types if needed
    if (assay_type %in% c("Number-cells", "Overpartitioning") && CONFIG$onlyCell) {
      fname <- basename(fi)
      cell_type_raw <- strsplit(fname, split = "_")[[1]][2]
      cell_type <- scTypeEval:::purge_label(cell_type_raw)
      
      if ("celltype" %in% colnames(df0)) {
        df0 <- df0 %>% filter(celltype == cell_type)
      }
    }
    
    if (nrow(df0) < 1) return(NULL)
    
    ident <- unique(df0$original.ident)
    
    # Standardize rates
    if (assay_type == "Number-samples") {
      unique_rates <- sort(unique(as.numeric(df0$rate)), decreasing = TRUE)
      target_values <- c(1, 0.9, 0.7, 0.5)[seq_along(unique_rates)]
      df0$rate <- target_values[match(df0$rate, unique_rates)]
    } else if (!assay_type %in% c("Signal-degradation", "Number-cells", "Overpartitioning")) {
      df0$rate <- case_when(
        assay_type == "Complexity" ~ str_count(as.character(df0$rate), "-") + 1,
        assay_type == "Granularity" ~ as.numeric(df0$rate),
        TRUE ~ as.numeric(df0$rate)
      )
      df0$rate <- minmax_scale(df0$rate)
    }
    
    # Calculate benchmarking scores
    if (CONFIG$benchmarking_scores) {
      summ_type <- case_when(
        grepl("Missclassific", fi, ignore.case = TRUE) ~ "Monotonic",
        grepl("NSamples", fi, ignore.case = TRUE) ~ "Constant",
        grepl("NCell", fi, ignore.case = TRUE) ~ "Constant",
        grepl("Nct", fi, ignore.case = TRUE) ~ "Constant",
        grepl("mergeCT", fi, ignore.case = TRUE) ~ "Constant",
        grepl("SplitCelltype", fi, ignore.case = TRUE) ~ "drop",
        TRUE ~ "Unknown"
      )
      
      g <- if (assay_type == "Granularity") "ident" else CONFIG$group
      
      df0 <- wr.assayPlot(df0,
                          type = summ_type,
                          return.df = TRUE,
                          group = g)
    }
    
    df0$assay <- assay_type
    df0$dataset <- dataset
    df0$ident <- ident
    
    return(df0)
  }
  
) %>%
  rbindlist(use.names = TRUE, fill = TRUE) %>%
  standardize_methods()

message("✓ Tasks data loaded: ", nrow(tasks_data), " rows")

```


## 3. Batch Effect Data

```{r load-batch-effect}
message("Loading batch effect data...")

li.fi <- list.files(file.path(wd, "BatchEffect_singleRun/output"),
                    recursive = TRUE,
                    full.names = TRUE) %>%
  grep("output_281025", ., invert = TRUE, value = TRUE)

mult_dataset <- bplapply(
  li.fi,
  BPPARAM = MulticoreParam(workers = CONFIG$n_workers, progressbar = TRUE),
  function(fi) {
    df <- readRDS(fi)
    if (is.null(df$Consistency)) return(NULL)
    
    df$Consistency %>%
      mutate(rate = stringr::str_count(as.character(batch), "-") + 1)
  }
) %>%
  data.table::rbindlist() 

batch_data <- mult_dataset %>%
  filter(
    !grepl("Healthy|Normal|Tirosh|Jerby|Bassez|Alvarez|nose|airway", batch) &
      rate < 3
  )

message("✓ Batch effect data loaded: ", nrow(batch_data), " rows")
```

## 4. Process Batch Effect Scores

```{r process-batch-effect}
# Identify multi-batches
multi_batches <- unique(batch_data$batch[str_detect(batch_data$batch, "-")])

batch_scores <- lapply(multi_batches, function(x) {
  calculate_batch_ratio(mb = x, df = batch_data)
}) %>%
  data.table::rbindlist() %>%
  standardize_methods() %>%
  mutate(assay = "Batch-effect",
         dataset = batch)

message("✓ Batch effect scores calculated: ", nrow(batch_scores), " rows")
```

## 5. Perturbation Effect Data

```{r process-perturbation}
message("Processing perturbation effect data...")

# Identify perturbation conditions
multi_condition <- mult_dataset %>%
  filter(rate < 3) %>%
  separate(batch, sep = "-", into = c("d1", "d2"), remove = FALSE) %>%
  separate(d1, sep = "_", into = c("source1", "condition1")) %>%
  separate(d2, sep = "_", into = c("source2", "condition2")) %>%
  filter(source1 == source2) %>%
  pull(batch) %>%
  unique() %>%
  .[str_detect(., "-")]

perturbation_scores <- lapply(multi_condition, function(x) {
  calculate_batch_ratio(mb = x, df = mult_dataset)
}) %>%
  data.table::rbindlist() %>%
  standardize_methods() %>%
  mutate(assay = "Perturbation-effect",
         dataset = batch)

message("✓ Perturbation scores calculated: ", nrow(perturbation_scores), " rows")
```



# Data Processing

## Combine All Tasks

```{r combine-tasks}
# Combine all task data
all_tasks <- list(tasks_data, batch_scores, perturbation_scores) %>%
  data.table::rbindlist(use.names = TRUE, fill = TRUE)

# Aggregate by group
all_tasks_agg <- all_tasks %>%
  group_by(dissimilarity_method, consistency.metric, method_type,
           assay, dataset, ident) %>%
  summarise(score = mean(score_celltype, na.rm = TRUE), .groups = "drop")


message("✓ All tasks combined: ", nrow(all_tasks_agg), " rows")
```
Up to here, compared with previous script and results are identical.

## Calculate Overall Scores

```{r calculate-overall-scores}
# Define weight vector for tasks
weight_vector <- c(
  "Signal-degradation" = 0.20,
  "Overpartitioning" = 0.20,
  "Granularity" = 0.10,
  "Complexity" = 0.10,
  "Number-samples" = 0.10,
  "Number-cells" = 0.10,
  "Batch-effect" = 0.10,
  "Perturbation-effect" = 0.10
)

# Verify weights sum to 1
stopifnot(abs(sum(weight_vector) - 1.0) < 0.001)

# Calculate weighted overall scores
overall_scores <- all_tasks_agg %>%
  group_by(method_type, assay) %>%
  summarise(score = mean(score, na.rm = TRUE), .groups = "drop") %>%
  group_by(method_type) %>%
  summarise(
    overall_score = weighted.mean(score, w = weight_vector[assay], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(overall_score)

message("✓ Overall scores calculated for ", nrow(overall_scores), " methods")
```

## Prepare Data for Visualization

```{r prepare-viz-data}
# Create summary statistics for each method-task combination
task_summary <- all_tasks_agg %>%
  group_by(method_type, assay) %>%
  summarise(
    mean_score = mean(score, na.rm = TRUE),
    median_score = median(score, na.rm = TRUE),
    sd_score = sd(score, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  group_by(assay) %>% 
  mutate(scaled_mean = minmax_scale(mean_score)) %>% 
  left_join(
    overall_scores %>%
      select(method_type, overall_score) %>% 
      mutate(overall_rank = rank(-overall_score, ties.method = "average")
  ),
    by = "method_type"
  )

# Add ranking information
task_summary <- task_summary %>%
  group_by(assay) %>%
  mutate(
    rank_in_task = rank(-mean_score, ties.method = "average"),
    percentile = rank_in_task / n() * 100
  ) %>%
  ungroup()
 

# Standardize assay names and order
task_summary <- task_summary %>%
  mutate(
    assay = factor(
      assay,
      levels = c(
        "Signal-degradation",
        "Overpartitioning",
        "Granularity",
        "Complexity",
        "Number-samples",
        "Number-cells",
        "Batch-effect",
        "Perturbation-effect"
      ),
      labels = c(
        "Signal degradation",
        "Over-partitioning",
        "Granularity",
        "Complexity",
        "Num. samples",
        "Num. cells",
        "Batch effect",
        "Perturbation effect"
      )
    )
  )

# Order methods by overall score (best to worst)
method_order <- overall_scores %>%
  arrange(desc(overall_score)) %>%
  pull(method_type)

task_summary <- task_summary %>%
  mutate(method_type = factor(method_type, levels = method_order))

message("✓ Visualization data prepared")
```

## Residual Effects for ICS Metric and Dissimilarity

```{r residual-effects}
# Residual effects computed per task (assay): component marginal mean minus assay grand mean.
# - `residual_effects_all` is computed on all methods (including metrics excluded from plots).
# - `residual_effects_plot` matches the plot universe (excludes Pseudobulk:Pearson and Average.similarity).

residual_effects_all <- compute_residual_effects(task_summary, for_plot = FALSE)
residual_effects_plot <- compute_residual_effects(task_summary, for_plot = TRUE)

message("✓ Residual effects computed: ", nrow(residual_effects_all), " rows (all); ",
  nrow(residual_effects_plot), " rows (plot universe)")
```

# Visualization

## Funkyheatmap: Benchmarking Results

```{r funkyheatmap-plot, fig.width=18, fig.height=10}
# Task scores (wide) - keep absolute values for sizing
plot_data <- task_summary %>%
  filter_metrics(for_plot = TRUE) %>%
  select(method_type, assay, median_score)

task_wide <- plot_data %>%
  tidyr::pivot_wider(names_from = assay, values_from = median_score)

# Extract task columns and reorder to match specification
task_cols <- setdiff(colnames(task_wide), "method_type")

# Define correct task order
task_order <- c(
  "Signal degradation",
  "Over-partitioning",
  "Granularity",
  "Complexity",
  "Num. samples",
  "Num. cells",
  "Batch effect",
  "Perturbation effect"
)

# Reorder task_wide columns to match specification
task_cols_ordered <- task_order[task_order %in% task_cols]
task_wide <- task_wide %>%
  dplyr::select(method_type, all_of(task_cols_ordered))

task_cols <- task_cols_ordered

# Overall scores with absolute values and rank-normalized coloring
overall_plot_data <- overall_scores %>%
  filter_metrics(for_plot = TRUE) %>%
  arrange(desc(overall_score)) %>%
  mutate(
    rank = rank(-overall_score, ties.method = "average"),
    overall_score_color = (rank - 1) / (n() - 1),
    overall_score_display = round(overall_score, 2)
  ) %>%
  select(method_type, overall_score,
         overall_score_color, overall_score_display)

# Resources with rank-normalized colors (no log scale)
resources_summary <- resources_data %>%
  filter_metrics(for_plot = TRUE) %>%
  group_by(method_type) %>%
  summarise(
    median_duration_ms = median(`Duration (ms)`, na.rm = TRUE),
    median_memory_mb = median(`Peak memory (MB)`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # Use absolute values for bar length and color (scaled to [0,1] for plotting)
    median_duration_scaled = minmax_scale(median_duration_ms),
    median_memory_scaled = minmax_scale(median_memory_mb)
  ) %>%
  select(method_type, median_duration_ms, median_memory_mb,
         median_duration_scaled, median_memory_scaled)

# Split method into consistency/dissimilarity
method_info <- overall_scores %>%
  filter_metrics(for_plot = TRUE) %>%
  select(method_type) %>%
  tidyr::separate(method_type,
                  into = c("consistency_metric", "dissimilarity_method"),
                  sep = " \\| ",
                  remove = FALSE,
                  fill = "right")

# Combine to single data frame for funkyheatmap
# Order by descending overall_score
fh_data <- method_info %>%
  left_join(task_wide, by = "method_type") %>%
  left_join(overall_plot_data, by = "method_type") %>%
  left_join(resources_summary, by = "method_type") %>%
  mutate(id = method_type) %>%
  arrange(desc(overall_score))

# For tasks: create ranking columns for color
n_methods <- nrow(fh_data)
for (col in task_cols) {
  scores <- fh_data[[col]]
  # Higher score is better, so best should be rank 0 (green)
  ranks <- rank(-scores, ties.method = "average", na.last = "keep")
  fh_data[[paste0(col, "_color")]] <- ifelse(is.na(scores), NA, (ranks - 1) / (n_methods - 1))
}

# Select columns: metadata, overall (color, value, display), tasks (value), task colors, resources
task_cols_color <- paste0(task_cols, "_color")
fh_data <- fh_data %>%
  dplyr::select(id, consistency_metric, dissimilarity_method, overall_score_color,
                overall_score, overall_score_display, 
                all_of(task_cols), all_of(task_cols_color),
                median_duration_ms, median_memory_mb)

# Recompute scaled resource columns (size/color need to be in [0,1] when scale_column = FALSE)
fh_data <- fh_data %>%
  mutate(
    median_duration_scaled = minmax_scale(median_duration_ms),
    median_memory_scaled = minmax_scale(median_memory_mb)
  )

# Column info for funkyheatmap (use official id_color/id_size/overlay fields)
cinfo_metadata <- tibble::tibble(
  id = c("consistency_metric", "dissimilarity_method", "overall_score", "overall_score_display"),
  id_color = c(NA_character_, NA_character_, "overall_score_color", NA_character_),
  id_size = c(NA_character_, NA_character_, "overall_score", NA_character_),
  group = c("Method", "Method", "Overall", "Overall"),
  name = c("Metric", "Dissimilarity", "", ""),
  geom = c("text", "text", "bar", "text"),
  palette = c(NA_character_, NA_character_, "overall_palette", NA_character_),
  width = c(7, 7, 4, 4),
  overlay = c(FALSE, FALSE, FALSE, TRUE),
  legend = c(FALSE, FALSE, TRUE, FALSE),
  label = c("consistency_metric", "dissimilarity_method", NA_character_, "overall_score_display"),
  size = c(4, 4, NA_real_, 5.5),
  fontface = c("plain", "plain", NA_character_, "bold"),
  hjust = c(0, 0, NA_real_, 0.5),
  vjust = c(0.5, 0.5, NA_real_, 0.5)
)

cinfo_tasks <- tibble::tibble(
  id = task_cols,
  id_color = task_cols_color,
  id_size = task_cols,
  group = rep("Tasks", length(task_cols)),
  name = task_cols,
  geom = rep("circle", length(task_cols)),
  palette = rep("task_palette_rank", length(task_cols)),
  width = rep(1.2, length(task_cols)),
  overlay = rep(FALSE, length(task_cols)),
  legend = rep(FALSE, length(task_cols))
)

cinfo_resources <- tibble::tibble(
  id = c("median_duration_scaled", "median_memory_scaled"),
  id_color = c("median_duration_scaled", "median_memory_scaled"),
  id_size = c("median_duration_scaled", "median_memory_scaled"),
  group = c("Resources", "Resources"),
  name = c("Duration (ms)", "Peak memory (MB)"),
  geom = c("bar", "bar"),
  palette = c("duration_palette", "memory_palette"),
  width = c(3, 3),
  overlay = c(FALSE, FALSE),
  legend = c(FALSE, FALSE),
  draw_outline = c(TRUE, TRUE)
)

cinfo <- dplyr::bind_rows(cinfo_metadata, cinfo_tasks, cinfo_resources)

palettes <- list(
  # 1️⃣ Task rank palette: white → solid blue
  task_palette_rank = grDevices::colorRampPalette(c("#08519c","#FFFFFF"))(50),
  
  # 2️⃣ Overall palette: warm gold → deep orange
  overall_palette = rev(grDevices::colorRampPalette(c("#fffaf0", "#fee391", "#fdae61", "#e6550d", "#a63603"))(9)),
  
  # 3️⃣ Duration palette: soft-to-deep teal/green
  duration_palette = grDevices::colorRampPalette(c("#f7fcf5", "#c7e9c0", "#74c476", "#238b45", "#005a32"))(7),
  
  # 4️⃣ Memory palette: soft lilac → rich purple (clearly distinct from overall)
  memory_palette = grDevices::colorRampPalette(c("#f7f4f9", "#dadaeb", "#9e9ac8", "#6a51a3", "#3f007d"))(7)
)

# Legends: 1 shared 0-1 scale for task rank colors; resources show min-max actual values
duration_range <- range(fh_data$median_duration_ms, na.rm = TRUE)
memory_range <- range(fh_data$median_memory_mb, na.rm = TRUE)

duration_labels <- c(
  as.character(round(duration_range[[1]], 0)), "", "", "",
  as.character(round(duration_range[[2]], 0))
)
memory_labels <- c(
  as.character(round(memory_range[[1]], 0)), "", "", "",
  as.character(round(memory_range[[2]], 0))
)

rank_labels <- as.character(c(1, "", "", "", n_methods))

legends <- list(
  list(
    title = "Intra-task Rank",
    palette = "task_palette_rank",
    geom = "bar",
    labels = rank_labels
  ),
  list(
    title = "Score",
    geom = "circle",
    # Add spacer entries to increase separation between circles
    labels = c("0", "", "", "0.25", "", "", "0.5", "", "", "0.75", "", "", "1"),
    size = c(0, 0, 0, 0.25, 0, 0, 0.5, 0, 0, 0.75, 0, 0, 1),
    color = c(
      "#666666", "#00000000", "#00000000",
      "#666666", "#00000000", "#00000000",
      "#666666", "#00000000", "#00000000",
      "#666666", "#00000000", "#00000000",
      "#666666"
    )
  ),
  # Keep an explicit (disabled) entry so funkyheatmap doesn't auto-add the legend
  list(
    palette = "overall_palette",
    enabled = FALSE
  ),
  list(
    title = "Duration (ms)",
    palette = "duration_palette",
    geom = "bar",
    labels = duration_labels
  ),
  list(
    title = "Memory (MB)",
    palette = "memory_palette",
    geom = "bar",
    labels = memory_labels
  )
)

funky_plot <- funkyheatmap::funky_heatmap(
  fh_data,
  column_info = cinfo,
  palettes = palettes,
  legends = legends,
  position_args = funkyheatmap::position_arguments(
    col_annot_offset = 3.5,
    expand_xmax = 16,
    expand_ymax = 0.8
  ),
  scale_column = FALSE
)

combined_plot <- funky_plot
print(funky_plot)
```

## Residual Effects: Consistency vs Dissimilarity

```{r residual-effects-plot, fig.width=16, fig.height=10}
# Heatmap of residual effects per task. Fill = residual effect; labels show per-task rank (1 = best).

fill_scale <- scale_fill_gradient2(
  low = "#2C7BB6",
  mid = "#F7F7F7",
  high = "#FDAE61",
  midpoint = 0,
  name = "Z-Score\n(effect)"
)

base_theme <- theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 13),
    axis.text.y = element_text(size = 12),
    legend.position = "right"
  )

top_panel_x_theme <- theme(
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank()
)

metric_df <- residual_effects_plot %>%
  filter(component == "Consistency metric")

metric_levels <- metric_df %>%
  group_by(level) %>%
  summarise(avg_resid = mean(residual_effect, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(avg_resid)) %>%
  pull(level)

metric_df <- metric_df %>%
  mutate(level = factor(level, levels = metric_levels))

metric_plot <- ggplot(metric_df, aes(x = assay, y = level, fill = residual_effect)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = as.integer(rank_residual)), size = 3.2, color = "black") +
  fill_scale +
  labs(
    x = NULL,
    y = NULL,
    title = "Consistency metric residual effects",
    subtitle = "Centered main effects (marginal mean minus assay grand mean); labels are ranks within each task"
  ) +
  base_theme +
  top_panel_x_theme

dissim_df <- residual_effects_plot %>%
  filter(component == "Dissimilarity method")

dissim_levels <- dissim_df %>%
  group_by(level) %>%
  summarise(avg_resid = mean(residual_effect, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(avg_resid)) %>%
  pull(level)

dissim_df <- dissim_df %>%
  mutate(level = factor(level, levels = dissim_levels))

dissim_plot <- ggplot(dissim_df, aes(x = assay, y = level, fill = residual_effect)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = as.integer(rank_residual)), size = 3.2, color = "black") +
  fill_scale +
  labs(
    x = NULL,
    y = NULL,
    title = "Dissimilarity method residual effects"
  ) +
  base_theme

residuals_plot <- metric_plot / dissim_plot +
  plot_layout(heights = c(1, 1), guides = "collect")

print(residuals_plot)
```

# Export Results

## Save Figure

```{r save-figure}
# Create output directory
fig_dir <- file.path(
  "~/OneDrive - Université de Genève/CSI/LabPapers/Intersample_Consistency/Figures/plots/Figure3/"
)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

# Save as PDF
pdf_file <- file.path(fig_dir, paste0("Benchmarking_", "_funkyheatmap.pdf"))
ggsave(
  pdf_file,
  plot = combined_plot,
  width = 15,
  height = 10,
  device = "pdf"
)

message("✓ Figure saved: ", pdf_file)

# Save residual-effects plot
residual_pdf_file <- file.path(fig_dir, paste0("Benchmarking_", "residual_effects.pdf"))
ggsave(
  residual_pdf_file,
  plot = residuals_plot,
  width = 8,
  height = 6,
  device = "pdf"
)

message("✓ Residual effects figure saved: ", residual_pdf_file)


```



## Save Table

```{r save-table}
results_table <- task_summary
# Create output directory
table_dir <- "~/OneDrive - Université de Genève/CSI/LabPapers/Intersample_Consistency/tables/"
dir.create(table_dir, showWarnings = FALSE, recursive = TRUE)

# Save as Excel
xlsx_file <- file.path(table_dir, paste0("Benchmarking_results_", CONFIG$group, ".xlsx"))

wb <- createWorkbook()
addWorksheet(wb, "Benchmarking Results")
writeData(wb, "Benchmarking Results", results_table)

# Residual effects worksheet (consistency metrics & dissimilarity methods)
addWorksheet(wb, "Residual Effects")
writeData(
  wb,
  "Residual Effects",
  residual_effects_all %>%
    mutate(assay = as.character(assay)) %>%
    arrange(component, assay, rank_residual)
)

# Format header
headerStyle <- createStyle(
  fontSize = 11,
  fontColour = "white",
  halign = "center",
  valign = "center",
  textDecoration = "bold",
  fgFill = "#4F81BD"
)
addStyle(wb, "Benchmarking Results", headerStyle, rows = 1, cols = 1:ncol(results_table))
addStyle(
  wb,
  "Residual Effects",
  headerStyle,
  rows = 1,
  cols = 1:ncol(residual_effects_all)
)

# Freeze header row
freezePane(wb, "Benchmarking Results", firstRow = TRUE)
freezePane(wb, "Residual Effects", firstRow = TRUE)

# Set column widths
setColWidths(wb, "Benchmarking Results", cols = 1:ncol(results_table), widths = "auto")
setColWidths(wb, "Residual Effects", cols = 1:ncol(residual_effects_all), widths = "auto")

# Save workbook
saveWorkbook(wb, xlsx_file, overwrite = TRUE)

message("✓ Table saved: ", xlsx_file)
```

## Summary Statistics

```{r summary-stats}
cat("\n=== BENCHMARKING SUMMARY ===\n\n")

cat("Configuration:\n")
cat("  Group:", CONFIG$group, "\n")
cat("  Only perturbed cells:", CONFIG$onlyCell, "\n\n")

cat("Data Summary:\n")
cat("  Total methods evaluated:", nrow(overall_scores), "\n")
cat("  Total tasks:", length(unique(task_summary$assay)), "\n")
cat("  Total datasets:", length(unique(all_tasks_agg$dataset)), "\n\n")

cat("Top 5 Methods (Overall Score):\n")
top_methods <- overall_scores %>%
  arrange(desc(overall_score)) %>%
  head(5)
for (i in 1:nrow(top_methods)) {
  cat(sprintf("  %d. %s: %.3f\n", 
              i, 
              top_methods$method_type[i], 
              top_methods$overall_score[i]))
}

cat("\nTask Weights:\n")
for (task in names(weight_vector)) {
  cat(sprintf("  %s: %.1f%%\n", task, weight_vector[task] * 100))
}

cat("\n=== END SUMMARY ===\n")
```

# LinePlots

## Load raw benchmarking output

### Benchmarking Tasks Data

```{r load-benchmarking-raw, warning=F}
message("Loading benchmarking tasks data...")

li.fi <- list.files(file.path(wd, "results/isc/tasks/output"),
                    pattern = "rds$",
                    recursive = TRUE,
                    full.names = TRUE) %>%
  grep("summary", ., value = TRUE, invert = TRUE)

tasks_data_raw <- bplapply(
  li.fi,
  BPPARAM = MulticoreParam(workers = CONFIG$n_workers, progressbar = TRUE),
  function(fi) {
    df0 <- readRDS(fi)
    if (is.null(df0)) {
      message("⚠️ Empty file: ", fi)
      return(NULL)
    }
    
    if (length(unique(df0$rate)) < 2) return(NULL)
    
    dataset_run_dir <- basename(dirname(dirname(fi)))
    dataset <- str_remove(dataset_run_dir, "_[0-9]{6}_[0-9]{8}$")
    
    # Infer assay type from filename
    assay_type <- case_when(
      grepl("Missclassific", fi, ignore.case = TRUE) ~ "Signal-degradation",
      grepl("NSamples", fi, ignore.case = TRUE) ~ "Number-samples",
      grepl("NCell", fi, ignore.case = TRUE) ~ "Number-cells",
      grepl("Nct", fi, ignore.case = TRUE) ~ "Complexity",
      grepl("mergeCT", fi, ignore.case = TRUE) ~ "Granularity",
      grepl("SplitCelltype", fi, ignore.case = TRUE) ~ "Overpartitioning",
      TRUE ~ "Unknown"
    )
    
    # Filter for specific cell types if needed
    if (assay_type %in% c("Number-cells", "Overpartitioning") && CONFIG$onlyCell) {
      fname <- basename(fi)
      cell_type_raw <- strsplit(fname, split = "_")[[1]][2]
      cell_type <- scTypeEval:::purge_label(cell_type_raw)
      
      if ("celltype" %in% colnames(df0)) {
        df0 <- df0 %>% filter(celltype == cell_type)
      }
    }
    
    if (nrow(df0) < 1) return(NULL)
    
    ident <- unique(df0$original.ident)
    
    # Standardize rates
    if (assay_type == "Number-samples") {
      unique_rates <- sort(unique(as.numeric(df0$rate)), decreasing = TRUE)
      target_values <- c(1, 0.9, 0.7, 0.5)[seq_along(unique_rates)]
      df0$rate <- target_values[match(df0$rate, unique_rates)]
    } else if (!assay_type %in% c("Signal-degradation", "Number-cells", "Overpartitioning")) {
      df0$rate <- case_when(
        assay_type == "Complexity" ~ str_count(as.character(df0$rate), "-") + 1,
        assay_type == "Granularity" ~ as.numeric(df0$rate),
        TRUE ~ as.numeric(df0$rate)
      )
      df0$rate <- minmax_scale(df0$rate)
    }
    
    df0$assay <- assay_type
    df0$dataset <- dataset
    df0$ident <- ident
    
    return(df0)
  }
  
) %>%
  rbindlist(use.names = TRUE, fill = TRUE) %>%
  standardize_methods()

message("✓ Raw tasks data loaded: ", nrow(tasks_data_raw), " rows")

```



## auxiliary plots

```{r}
# line plots is for all tasks except for over-partitioning, batch effect, and perturbation effect which need to use barplot

lineplot <- function(df,
                     qs = c(0.05, 0.25, 0.75, 0.95),
                     smooth = TRUE,
                     span = 0.7,
                     title = "",
                     subtitle = ""){
  summary_df <- df %>%
    group_by(rate) %>%
    summarise(
      median = median(measure),
      q1 = quantile(measure, qs[1]),
      q2 = quantile(measure, qs[2]),
      q3 = quantile(measure, qs[3]),
      q4 = quantile(measure, qs[4]),
      .groups = "drop"
    )
  
  # smooth values per facet
  smooth_df <- summary_df
  if (smooth && nrow(summary_df) >= 4) {
    smooth_df <- smooth_df %>%
      mutate(
        median = as.numeric(predict(loess(median ~ rate, span = span))),
        q1 = as.numeric(predict(loess(q1 ~ rate, span = span))),
        q2 = as.numeric(predict(loess(q2 ~ rate, span = span))),
        q3 = as.numeric(predict(loess(q3 ~ rate, span = span))),
        q4 = as.numeric(predict(loess(q4 ~ rate, span = span)))
      )
  }
  
  fpl <- ggplot(smooth_df, aes(x = rate, y = median, group = 1)) +
    geom_line(size = 1.2, color = "black") +
    geom_ribbon(aes(ymin = q1, ymax = q4), alpha = 0.2, fill = "steelblue") +
    geom_ribbon(aes(ymin = q2, ymax = q3), alpha = 0.4, fill = "steelblue") +
      labs(title = title, subtitle = subtitle) +
    ggpubr::theme_classic2() +
      theme(title = element_text(size = 15, hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(size = 14, hjust = 0.5))

  return(fpl)
}

boxplot <- function(df,
                    title = "",
                    subtitle = "",
                    rate_levels = NULL){
  if ("rate_label" %in% colnames(df)) {
    if (!is.null(rate_levels)) {
      df <- df %>%
        mutate(rate_plot = factor(rate_label, levels = rate_levels))
    } else {
      df <- df %>%
        mutate(rate_plot = factor(rate_label, levels = unique(rate_label)))
    }
  } else {
    df <- df %>%
      mutate(rate_plot = factor(rate, levels = sort(unique(rate))))
  }

  pl <- ggplot(df, aes(x = rate_plot, y = measure,
                       color = rate_plot)) +
    geom_boxplot(width = 0.55,
                 outlier.shape = NA,
                 alpha = 0.35,
                 show.legend = F) +
    geom_jitter(
      shape = 21,
      size = 1,
      height = 0,
      width = 0.12,
      alpha = 0.3,
      show.legend = F
    ) +
    scale_color_manual(values = c("grey15", "#3B6FB6")) +
    scale_x_discrete(drop = FALSE) +
    labs(title = title, subtitle = subtitle) +
    ggpubr::theme_classic2() +
    theme(title = element_text(size = 15, hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(size = 14, hjust = 0.5))

  return(pl)
}
```

## Do plots

One plot for each of the best-worst-assay chunk output (total of 16, 8 tasks)

```{r panel-perturbation-data}

# Ensure method_type is present in tasks_data_raw (if not already added)
if (!"method_type" %in% colnames(tasks_data_raw)) {
  tasks_data_raw <- tasks_data_raw %>%
    mutate(method_type = paste(consistency.metric, dissimilarity_method, sep = " | "))
}

# Assemble raw data for the panel (tasks + batch + perturbation)
panel_tasks <- tasks_data_raw

# Raw batch-effect (single vs multi) using mult_dataset
panel_batch <- mult_dataset %>%
  mutate(
    assay = "Batch-effect",
    rate_raw = rate,
    rate = dplyr::if_else(rate_raw <= 1, 0, 1),
    rate_label = dplyr::if_else(rate_raw <= 1, "Single", "Multi")
  ) %>%
  select(-rate_raw) %>%
  standardize_methods()

# Raw perturbation-effect (single conditions vs combined) using mult_dataset
panel_perturb <- mult_dataset %>%
  mutate(
    assay = "Perturbation-effect",
    rate_raw = rate,
    rate = dplyr::if_else(rate_raw <= 1, 0, 1),
    rate_label = dplyr::if_else(rate_raw <= 1, "Single", "Paired")
  ) %>%
  select(-rate_raw) %>%
  standardize_methods()

panel_df <- dplyr::bind_rows(
  panel_tasks,
  panel_batch,
  panel_perturb
)

# Remap plot_data assay names to internal format for lookup
assay_remap <- c(
  "Signal degradation" = "Signal-degradation",
  "Over-partitioning" = "Overpartitioning",
  "Granularity" = "Granularity",
  "Complexity" = "Complexity",
  "Num. samples" = "Number-samples",
  "Num. cells" = "Number-cells",
  "Batch effect" = "Batch-effect",
  "Perturbation effect" = "Perturbation-effect"
)
```

```{r best-worst-methods}
# Method picks for each task in plot_data
best_worst <- plot_data %>%
  group_by(assay) %>%
  summarise(
    best_method = method_type[which.max(median_score)],
    worst_method = method_type[which.min(median_score)],
    .groups = "drop"
  ) %>%
  mutate(assay_internal = assay_remap[assay])
```

### render plots

```{r panel-perturbation-data, fig.width=20, fig.height=10}

make_panel_plot <- function(display_assay_name, which_row = c("best", "worst")) {
  which_row <- match.arg(which_row)
  
  # Look up the internal assay name and best/worst method
  internal_assay <- assay_remap[display_assay_name]
  if (is.na(internal_assay)) {
    return(ggplot() + theme_void() + labs(title = display_assay_name))
  }
  
  row_data <- best_worst %>% filter(assay == display_assay_name)
  if (nrow(row_data) == 0) {
    return(ggplot() + theme_void() + labs(title = display_assay_name))
  }
  
  method <- row_data[[paste0(which_row, "_method")]]
  if (is.na(method) || method == "") {
    return(ggplot() + theme_void() + labs(title = display_assay_name))
  }

  method_title <- gsub(" \\| ", "\n", method)

  df0 <- panel_df %>%
    filter(assay == internal_assay, method_type == method)

  if (nrow(df0) < 1) {
    return(ggplot() + theme_void() + labs(title = display_assay_name))
  }

  is_two_level <- dplyr::n_distinct(df0$rate) <= 2
  if (is_two_level) {
    rate_levels <- NULL
    if (display_assay_name == "Over-partitioning") {
      rate_levels <- c("Split", "Complete")
    } else if (display_assay_name == "Batch effect") {
      rate_levels <- c("Single", "Multi")
    } else if (display_assay_name == "Perturbation effect") {
      rate_levels <- c("Single", "Paired")
    }

    if (!"rate_label" %in% colnames(df0) || all(is.na(df0$rate_label))) {
      rate_vals <- sort(unique(as.numeric(as.character(df0$rate))))
      rate_keys <- as.character(rate_vals)
      if (is.null(rate_levels)) {
        rate_levels <- rate_keys
      }
      df0 <- df0 %>%
        mutate(rate_label = dplyr::recode(as.character(rate),
                                          !!!setNames(rate_levels, rate_keys)))
    }

    plot_obj <- boxplot(df0,
                        title = display_assay_name,
                        subtitle = method_title,
                        rate_levels = rate_levels)
  } else {
    # coerce rate to numeric for line plots
    df0_numeric <- df0 %>% mutate(rate = as.numeric(as.character(rate)))
    if (any(is.na(df0_numeric$rate))) {
      df0_numeric <- df0_numeric %>% filter(!is.na(rate))
    }

    if (nrow(df0_numeric) < 1) {
      return(ggplot() + theme_void() + labs(title = display_assay_name))
    }

    plot_obj <- lineplot(df0_numeric,
                         title = display_assay_name,
                         subtitle = method_title)
  }

  show_axis <- which_row == "worst" && display_assay_name == task_order[[1]]
  if (show_axis) {
    plot_obj <- plot_obj +
      labs(x = "Perturbation rate",
           y = "Inter-Sample Consistency score") +
      theme(axis.title = element_text(size = 10, face = "bold"))
  } else {
    plot_obj <- plot_obj +
      theme(axis.title = element_blank())
  }

  return(plot_obj)
}

plots_best <- lapply(task_order, make_panel_plot, which_row = "best")
plots_worst <- lapply(task_order, make_panel_plot, which_row = "worst")

panel_plot <- (wrap_plots(plots_best, ncol = 8) /
                wrap_plots(plots_worst, ncol = 8))

print(panel_plot)

# Export the panel plot
panel_pdf_file <- file.path(fig_dir, paste0("Benchmarking_", "perturbation_panel.pdf"))
ggsave(
  panel_pdf_file,
  plot = panel_plot,
  width = 20,
  height = 6,
  device = "pdf"
)

message("✓ Perturbation panel plot saved: ", panel_pdf_file)
```




