---
title: "Synthetic gene expression perturbation to evaluate ISC metrics"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# 1. Introduction
Change GEX of some cell types in different ways to evaluate how ISC capture this.

- Do multiple degrees (4) of gex perturbation, only on 2000 hvg to be evaluated by ISC later.

- The cells to be changed should be aprox half of cells within a cell type to be studied, either:
   * Half cells of each sample
   * All cells of half of the samples
   * Variant proportion of cells across samples

- Then evaluate these changes in 2 situations:
   * Same label in unchanged and changed cells
   * Different label in unchanged and changed cells


# 2. load depencies & set paths
```{r, message=F, warning=F}
library(dplyr)
library(ggplot2)
library(tidyr)
library(BiocParallel)
library(Seurat)
library(scTypeEval)
source("utils.R")

global = "2label_silhouette | Pseudobulk:Cosine"
local = "silhouette | recip_classif:Match"
```

## exported figures directory
```{r}
fig_dir <- "figs"

dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```


# 3. Load dataset

Use a simple dataset (PBMC) with enough samples and simplify it to different cell types.

```{r}
path <- "~/Documents/Projects/Consistency_metrics_benchmark/datasets/rds_datasets/Mitchel_Cambridge_Healthy.rds"
cambridge_healthy <- load_single_cell_object(path)
```

## visualization
```{r}
seu <- CreateSeuratObject(counts = cambridge_healthy$counts,
                          meta.data = cambridge_healthy$metadata)
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu)
seu <- RunUMAP(seu, dims = 1:20)

DimPlot(seu,
        group.by = "OriginalAnnotationLevel1",
        label = T
      )
DimPlot(seu,
        group.by = "sample",
        label = T
      )
```

## Basal ISC
```{r}
data("black_list")
bl <- list(black_list$TCR,
           black_list$Immunoglobulins,
           black_list$Ygenes) |>
       unlist()
sc <- wrapper_scTypeEval(count_matrix = cambridge_healthy$counts,
                         metadata = cambridge_healthy$metadata,
                         ident = "OriginalAnnotationLevel1",
                         sample = "sample",
                         dissimilarity_method = c("Pseudobulk:Cosine", "recip_classif:Match"),
                         black_list = bl,
                         ncores = 6,
                         verbose = F)
```

```{r}
cons_basal <- get_consistency_wide(sc)
```


Based on basal ISC we picked the cell type with highest ISC.

We will perturbated the cell type with most ISC basal (B_cell)

```{r}
keep <- c("B_cell", "CD14", "NK_16hi",
          "CD4", "DCs")
to_p <- "B_cell"
ref_p <- "DCs" # cell type ref to add noise
```


## subset

```{r}
metadata <- cambridge_healthy$metadata %>% 
   filter(OriginalAnnotationLevel1 %in% keep)
counts <- cambridge_healthy$counts[,rownames(metadata)]
```

### Seurat umap
```{r}
seu <- CreateSeuratObject(counts = counts,
                          meta.data = metadata)
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu)
seu <- RunUMAP(seu, dims = 1:20)

DimPlot(seu,
        group.by = "OriginalAnnotationLevel1",
        label = T
      )

```


### scType pca

```{r}
set.seed(22)
sc <- wrapper_scTypeEval(count_matrix = counts,
                         metadata = metadata,
                         ident = "OriginalAnnotationLevel1",
                         sample = "sample",
                         dissimilarity_method = c("Pseudobulk:Cosine", "recip_classif:Match"),
                         black_list = bl,
                         ncores = 6,
                         verbose = F)
hvg <- sc@gene_lists$HVG
```

```{r}
plot_pca(sc)

(cons_basal_subset <- get_consistency_wide(sc))
```
## test perturbation
```{r}

ss <- unique(metadata$sample)[1:6]
metadata_perturb <- metadata %>%
   mutate(pertub_annot = ifelse(OriginalAnnotationLevel1 == to_p & sample %in% ss,
                                "perturbed",
                                as.character(OriginalAnnotationLevel1))
          )

cells_perturb <- metadata_perturb %>% 
   filter(pertub_annot == "perturbed") %>% 
   rownames()

cells_reference <- metadata_perturb %>% 
   filter(pertub_annot == ref_p) %>% 
   rownames()

# keep only hvg
counts_hvg <- counts[hvg, rownames(metadata_perturb)]
   
counts_perturb <- perturb_cells(counts_hvg,
                                cells_perturb = cells_perturb,
                                cells_reference = cells_reference,
                                noise = 1)

metadata_perturb <- metadata_perturb %>% 
   filter(pertub_annot != ref_p)
counts_perturb <- counts_perturb[, rownames(metadata_perturb)]

```

#### consistency
```{r}
set.seed(22)
id <- "OriginalAnnotationLevel1"
sc <- wrapper_scTypeEval(count_matrix = counts_perturb,
                         metadata = metadata_perturb,
                         ident = id,
                         sample = "sample",
                         gene_list = list("hvg" = hvg),
                         dissimilarity_method = c("Pseudobulk:Cosine", "recip_classif:Match"),
                         black_list = bl,
                         ncores = 6,
                         verbose = F)

plot_pca(sc)

(cons_basal_test <- get_consistency_wide(sc))
```


# Loop all settings

```{r}
metadata <- cambridge_healthy$metadata %>% 
   filter(OriginalAnnotationLevel1 %in% keep)
counts <- cambridge_healthy$counts[,rownames(metadata)]
```


```{r, warning=F}
noise_rates <- c(0, 0.025, 0.05, 0.075, 0.1, 0.25, 0.5, 2, 3, 30, 100)
#noise_rates <- c(0.075, 0.25, 30, 100)
label_rates <- c("same" = "ident",
                 "different" = "pertub_annot")
perturbation_type <- c("per_sample", "same_sample", "different_sample")
to_p <- "B_cell"
ref_p <- "DCs" # cell type ref to add noise
#results <- list()

for(per in perturbation_type){
   md_perturbed <- select_perturbed_cells(metadata,
                                          ident = "OriginalAnnotationLevel1",
                                          sample = "sample",
                                          perturb_celltype = to_p,
                                          type = per,
                                          seed = 22)
   cells_perturb <- md_perturbed %>% 
      filter(pertub_annot == "perturbed") %>% 
      rownames()
   
   cells_reference <- md_perturbed %>% 
      filter(pertub_annot == ref_p) %>% 
      rownames()
   
   # keep all genes
   counts_hvg <- counts[, rownames(md_perturbed)]
   
   for(n in noise_rates){
      counts_perturb <- perturb_cells(counts_hvg,
                                      cells_perturb = cells_perturb,
                                      cells_reference = cells_reference,
                                      noise = n)
      
      md_perturbed <- md_perturbed %>% 
         filter(pertub_annot != ref_p)
      counts_perturb <- counts_perturb[, rownames(md_perturbed)]
      
      
      for(lab in names(label_rates)){
         
         tit <- paste(per, n, lab)
         cat("## Computing ISC for ", tit, "\n")
         
         set.seed(22)
         sc <- wrapper_scTypeEval(count_matrix = counts_perturb,
                                  metadata = md_perturbed,
                                  ident = label_rates[[lab]],
                                  sample = "sample",
                                  dissimilarity_method = c("Pseudobulk:Cosine",
                                                           "recip_classif:Match"),
                                  black_list = bl,
                                  ncores = 7,
                                  verbose = F)
         
         pca <- plot_pca(sc)
         rmm <- plot_heatmap(sc,
                             dissimilarity_slot = "recip_classif:Match",
                             sort_consistency = "silhouette")
         
         cons <- get_consistency_wide(sc, wide = F) %>% 
            mutate(perturbation_type = per,
                   noise = n,
                   label = lab)
         
         res <- list(plot_pca = pca,
                     rmm = rmm,
                     consistency = cons)
         
         results[[tit]] <- res
         
      }
   }
}
```

## join dataframe
```{r}
allres <- lapply(results, function(x){x$consistency}) %>% 
   data.table::rbindlist() %>% 
   mutate(isc = paste(method_type, label, sep = "_"))
```

## line plots

```{r line_plot}
cols <- c(
  "global"      = "#1B9E77",
  "local" = "#D95F02"
)

df <- allres %>% 
   filter(celltype == scTypeEval:::purge_label(to_p)) %>% 
   mutate(noise = factor(noise),
          perturbation_type = factor(perturbation_type,
                                     levels = c("per_sample","same_sample","different_sample"),
                                     labels = c("Different samples", "50% each sample", "25% | 75% of each sample")))

pl <-  df %>% 
   ggplot(aes(noise, measure, color = method_type)) +
   geom_point(aes(shape = label)) +
   geom_line(aes(group = isc,
                 linetype = label)) +
   scale_color_manual(values = cols) +
   facet_wrap(~perturbation_type, ncol = 1) +
   labs(y = "ISC",
        x = "Perturbation degree") +
   ggpubr::theme_classic2()
```

```{r}
figpath <- file.path(fig_dir, "lineplots.pdf")

pdf(figpath, width = 7, height = 5)
pl
dev.off()
```



```{r}
allpca <- lapply(names(results),
                 function(x){results[[x]]$plot_pca$pseudobulk +
                       ggpubr::theme_classic2() +
                       theme(title = element_blank(),
                             axis.ticks = element_blank(),
                             axis.title = element_blank(),
                             axis.text = element_blank())})
names(allpca) <- names(results)

allpca_combi <- ggpubr::ggarrange(plotlist = allpca)

allph <- lapply(names(results),
                 function(x){results[[x]]$rmm +
                       theme(title = element_blank(),
                             axis.text.x = element_blank())})
names(allph) <- names(results)

allph_combi <- ggpubr::ggarrange(plotlist = allph)
```

```{r}
figpath_pca <- file.path(fig_dir, "pca")
dir.create(figpath_pca)
figpath_rmm <- file.path(fig_dir, "rmm")
dir.create(figpath_rmm)

for(a in names(allpca)){
   path <- file.path(figpath_pca, paste0(a, ".pdf"))
   pdf(path, width = 2, height = 2)
   print(allpca[[a]])
   dev.off()
   
   path <- file.path(figpath_rmm, paste0(a, ".pdf"))
   pdf(path, width = 2, height = 1.2)
   print(allph[[a]])
   dev.off()
}
```












