---
title: "Synthetic gene expression perturbation to evaluate ISC metrics"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# 1. Introduction
Change GEX of some cell types in different ways to evaluate how ISC capture this.

- Do multiple degrees (4) of gex perturbation, only on 2000 hvg to be evaluated by ISC later.

- The cells to be changed should be aprox half of cells within a cell type to be studied, either:
   * Half cells of each sample
   * All cells of half of the samples
   * Variant proportion of cells across samples

- Then evaluate these changes in 2 situations:
   * Same label in unchanged and changed cells
   * Different label in unchanged and changed cells


# 2. load depencies & set paths
```{r, message=F, warning=F}
library(dplyr)
library(ggplot2)
library(tidyr)
library(BiocParallel)
library(Seurat)
library(scTypeEval)

global = "2label_silhouette | Pseudobulk:Cosine"
local = "silhouette | recip_classif:Match"
```

## exported figures directory
```{r}
fig_dir <- "figs"

dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```


# 3. Load dataset

Use a simple dataset (PBMC) with enough samples and simplify it to different cell types.

```{r}
path <- "~/Documents/Projects/Consistency_metrics_benchmark/datasets/rds_datasets/Mitchel_Cambridge_Healthy.rds"
cambridge_healthy <- load_single_cell_object(path)
```

## visualization
```{r}
seu <- CreateSeuratObject(counts = cambridge_healthy$counts,
                          meta.data = cambridge_healthy$metadata)
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu)
seu <- RunUMAP(seu, dims = 1:20)

DimPlot(seu,
        group.by = "OriginalAnnotationLevel1",
        label = T
      )
DimPlot(seu,
        group.by = "sample",
        label = T
      )
```

## Basal ISC
```{r}
data("black_list")
bl <- list(black_list$TCR,
           black_list$Immunoglobulins,
           black_list$Ygenes) |>
       unlist()
sc <- wrapper_scTypeEval(count_matrix = cambridge_healthy$counts,
                         metadata = cambridge_healthy$metadata,
                         ident = "OriginalAnnotationLevel1",
                         sample = "sample",
                         dissimilarity_method = c("Pseudobulk:Cosine", "recip_classif:Match"),
                         black_list = bl,
                         ncores = 6)
```

```{r}
consis <- get_consistency(sc,
                          consistency_metric = c("silhouette", "2label_silhouette"))
consis <- consis %>% 
   mutate(method_type = paste(consistency_metric,
                              dissimilarity_method,
                              sep = " | ")) %>% 
   filter(method_type %in% c(global, local)) %>% 
   mutate(method_type = factor(method_type,
                               levels = c(global, local),
                               labels = c("global", "local")))
cons_basal <- consis %>% 
   select(method_type, measure, celltype) %>% 
   pivot_wider(names_from = "method_type",
               values_from = "measure") %>% 
   mutate(product = global * local)
```


Based on basal ISC we picked the cell type with highest ISC.

We will perturbated the cell type with most ISC basal (B_cell)

```{r}
keep <- c("B_cell", "CD14", "NK_16hi", "CD4")
```


## subset

```{r}
metadata <- cambridge_healthy$metadata %>% 
   filter(OriginalAnnotationLevel1 %in% keep)
counts <- cambridge_healthy$counts[,rownames(metadata)]
```

### Seurat umap
```{r}
seu <- CreateSeuratObject(counts = counts,
                          meta.data = metadata)
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu)
seu <- RunUMAP(seu, dims = 1:20)

DimPlot(seu,
        group.by = "OriginalAnnotationLevel1",
        label = T
      )

```


### scType pca

```{r}
sc <- wrapper_scTypeEval(count_matrix = counts,
                         metadata = metadata,
                         ident = "OriginalAnnotationLevel1",
                         sample = "sample",
                         dissimilarity_method = c("Pseudobulk:Cosine", "recip_classif:Match"),
                         black_list = bl,
                         ncores = 6,
                         verbose = F)
```

```{r}
plot_pca(sc)
```

```{r}

```







