---
title: Consistency metrics on label transfer set up
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

```{r libraries}
require(ggplot2)
require(dplyr)
require(tidyr)
require(ggpubr)

# Custom theme
theme_set(theme_bw(base_size = 12))
```

```{r exporting_dir}
fig_dir <- file.path(
  "~/Documents/Projects/ISC_benchmark_reproducibility/label_transfer_task/figs"
)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

Load dataf from results (within dataset splitting) and results_between (prediciton between different datasets but same annotation)
```{r load_data}
paths <- 
  c("intra-dataset" = "results/aggregated/label_transfer_consistency_detailed.csv",
    "inter-dataset" = "results_between/aggregated/label_transfer_between_consistency_detailed.csv")
res <- lapply(paths, read.csv)
# Preview
lapply(res, head)
```

# Generate ranking-based color palette
```{r create_color_palette}
# Calculate rankings to determine color palette
# Step 1: Calculate mean product per dataset and classifier (across cell types)
# Step 2: Calculate geometric mean across datasets
rankings_for_palette <- cons_combined %>%
  filter(!is.na(f1) & !is.na(product) & classifier != "true_labels") %>%
  group_by(dataset, classifier) %>%
  summarise(
    mean_product_per_dataset = mean(product, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(classifier) %>%
  summarise(
    geom_mean_product = exp(mean(log(mean_product_per_dataset), na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(rank = rank(-geom_mean_product, ties.method = "average")) %>%
  arrange(rank)

# Create a unified, elegant, color-blind friendly palette
# Integrating highly distinguishable top 3 best and top 3 worst colors
n_classifiers <- nrow(rankings_for_palette)

# Define the 6 distinct colors for top 3 best and top 3 worst
colors_top3_best <- c("#0173B2", "#029E73", "#B19CD9")  # Deep blue, forest green, malva/violet
colors_top3_worst <- c("#DE8F05", "#FFD700", "#882255")  # Orange, gold, purple-red

if (n_classifiers <= 6) {
  # If 6 or fewer classifiers, use the 6 distinct colors
  gradient_colors <- c(colors_top3_best, colors_top3_worst)[1:n_classifiers]
} else {
  # For more than 6 classifiers, create a palette that:
  # - Uses top 3 best colors for the best performers
  # - Interpolates intermediate colors
  # - Uses top 3 worst colors for the worst performers
  
  # Generate intermediate colors using color interpolation
  # From best (blue) to worst (red)
  intermediate_palette <- colorRampPalette(c("#0173B2", "#E0E0E0", "#882255"))(n_classifiers - 6)
  
  # Combine: top 3 best + intermediate + top 3 worst
  gradient_colors <- c(
    colors_top3_best,
    intermediate_palette[1:(n_classifiers - 6)],
    colors_top3_worst
  )
}

# Create named palette
classifier_colors <- setNames(gradient_colors, rankings_for_palette$classifier)

# Identify top 6 classifiers
top_3_best <- rankings_for_palette$classifier[1:3]
top_3_worst <- tail(rankings_for_palette$classifier, 3)
top_6_classifiers <- c(top_3_best, top_3_worst)

# For f1_vs_product_mean plot: highlight top 6, gray out the rest
classifier_colors_highlight <- classifier_colors
classifier_colors_highlight[!names(classifier_colors_highlight) %in% top_6_classifiers] <- "#808080"

print("Unified color palette by ranking:")
print(classifier_colors)
print("\nTop 3 best classifiers:")
print(top_3_best)
print("\nTop 3 worst classifiers:")
print(top_3_worst)


```



# Show mean F1 vs mean product
Add in the plot R2 and p-val
```{r f1_vs_product_mean}
# Calculate mean F1 and product per dataset and classifier
mean_metrics <- cons_combined %>%
  filter(!is.na(f1) & !is.na(product)) %>%
  filter(classifier != "true_labels") %>%
  mutate(classifier = factor(classifier,
                             levels = c(rankings_for_palette$classifier))) %>% 
  group_by(dataset, classifier) %>%
  summarise(
    mean_f1 = mean(f1, na.rm = TRUE),
    mean_product = mean(product, na.rm = TRUE),
    .groups = "drop"
  )

# Add highlight status to data (for filtering in legend)
mean_metrics <- mean_metrics %>%
  mutate(is_highlighted = classifier %in% top_6_classifiers)

# Calculate correlation
cor_test_mean <- cor.test(mean_metrics$mean_f1, mean_metrics$mean_product)
r2_mean <- cor_test_mean$estimate^2
pval_mean <- cor_test_mean$p.value

# Create plot
p2 <- ggplot(mean_metrics, aes(x = mean_f1, y = mean_product)) +
  # First layer: gray dots for non-highlighted classifiers (drawn first, so in back)
  geom_point(data = mean_metrics %>% filter(!is_highlighted), 
             aes(color = classifier), alpha = 0.4, size = 1.5) +
  # Second layer: colored dots for highlighted classifiers (drawn last, so in front)
  geom_point(data = mean_metrics %>% filter(is_highlighted), 
             aes(color = classifier), alpha = 0.7, size = 1.5) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed", se = TRUE) +
  scale_color_manual(values = classifier_colors_highlight) +
  labs(
    #title = "Mean F1 vs. Mean Consistency Product per Dataset",
    subtitle = sprintf("R² = %.3f, p-value = %.2e", r2_mean, pval_mean),
    x = "Mean F1 Score",
    y = "Mean ISC Predictions",
    color = "Classifier"
  ) +
  ggpubr::theme_classic2() +
  theme(
    legend.position = "right",
    legend.title = element_blank()
  )

# Custom legend: only show highlighted classifiers
# Get unique highlighted classifiers from the data
highlighted_classifiers <- unique(mean_metrics$classifier[mean_metrics$is_highlighted])
p2 <- p2 + 
  scale_color_manual(
    values = classifier_colors_highlight,
    breaks = highlighted_classifiers
  )

print(p2)

p2.1 <- p2 +
  facet_wrap(~classifier)
print(p2.1)

```

## Export plots

```{r}
# Save as PDF
pdf_file <- file.path(fig_dir, paste0("LabelProjection_", "mean_scatter.pdf"))
ggsave(
  pdf_file,
  plot = p2,
  width = 4,
  height = 3,
  device = "pdf"
)
```


# Compare ranking of classifiers either using F1 or product and how they agree in the ranking

```{r ranking_comparison}
# Calculate overall rankings
# Step 1: Calculate mean metrics per dataset and classifier (across cell types)
# Step 2: Calculate geometric mean across datasets
rankings <- cons_combined %>%
  filter(!is.na(f1) & !is.na(product) & classifier != "true_labels") %>%
  group_by(dataset, classifier) %>%
  summarise(
    mean_f1_per_dataset = mean(f1, na.rm = TRUE),
    mean_product_per_dataset = mean(product, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(classifier) %>%
  summarise(
    geom_mean_f1 = exp(mean(log(mean_f1_per_dataset), na.rm = TRUE)),
    geom_mean_product = exp(mean(log(mean_product_per_dataset), na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    rank_f1 = rank(-geom_mean_f1, ties.method = "average"),
    rank_product = rank(-geom_mean_product, ties.method = "average"),
    rank_difference = abs(rank_f1 - rank_product)
  ) %>%
  mutate(classifier = factor(classifier,
                             levels = c(rankings_for_palette$classifier))) %>%
  arrange(rank_product)

print(rankings)

# Spearman correlation between rankings
rank_cor <- cor.test(rankings$rank_f1, rankings$rank_product, method = "spearman")

cat(sprintf("\nSpearman correlation between F1 and Product rankings: %.3f (p = %.2e)\n", 
            rank_cor$estimate, rank_cor$p.value))

# Visualize ranking comparison
rankings_long <- rankings %>%
  select(classifier, rank_f1, rank_product) %>%
  pivot_longer(cols = c(rank_f1, rank_product), names_to = "metric", values_to = "rank") %>%
  mutate(metric = recode(metric, 
                         rank_f1 = "F1 Rank", 
                         rank_product = "Product Rank"))

p5 <- ggplot(rankings_long, aes(x = metric, y = rank, group = classifier, color = classifier)) +
  geom_line(size = 1.2, alpha = 0.7) +
  geom_point(size = 3) +
  scale_y_reverse() +
  scale_color_manual(values = classifier_colors) +
  labs(
    title = "Classifier Ranking: F1 vs. Consistency Score",
    subtitle = sprintf("Spearman ρ = %.3f (p = %.2e)", rank_cor$estimate, rank_cor$p.value),
    x = "",
    y = "Rank (1 = Best)",
    color = "Classifier"
  ) +
  theme(legend.position = "right")

print(p5)

# Ranking agreement table
p6 <- ggplot(rankings, aes(x = rank_product, y = rank_f1)) +
  geom_point(aes(color = classifier),
             size = 4,) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  ggrepel::geom_label_repel(aes(label = classifier,
                                color = classifier),
                           size = 5) +
  scale_color_manual(values = classifier_colors) +
  scale_x_reverse(breaks = 1:nrow(rankings)) +
  scale_y_reverse(breaks = 1:nrow(rankings)) +
  labs(
    title = "Ranking Agreement: ISC vs F1",
    x = "ISC Rank",
    y = "F1 Rank",
    color = "Classifier"
  ) +
  coord_fixed() +
  ggpubr::theme_classic2() +
  theme(legend.position = "none") 

print(p6)
```

## ranking mean
Per-dataset ranking of classifiers based on product score, then average rank across datasets

```{r ranking_per_dataset}
# Parameter: use geometric mean (TRUE) or arithmetic mean (FALSE)
use_geom_mean <- TRUE

# Step 1: Calculate mean metrics per dataset and classifier (across cell types)
dataset_metrics_mean <- cons_combined %>%
  filter(!is.na(f1) & !is.na(product) & classifier != "true_labels") %>%
  group_by(dataset, classifier) %>%
  summarise(
    mean_f1 = mean(f1, na.rm = TRUE),
    mean_product = mean(product, na.rm = TRUE),
    .groups = "drop"
  )

# Step 2: Compute ranking of classifiers per dataset for both metrics
dataset_rankings <- dataset_metrics_mean %>%
  group_by(dataset) %>%
  mutate(
    rank_f1 = rank(-mean_f1, ties.method = "average"),
    rank_product = rank(-mean_product, ties.method = "average")
  ) %>%
  ungroup()

# Step 3: Calculate mean rank per classifier across datasets
# Use either arithmetic mean or geometric mean based on parameter
mean_rank_per_classifier <- dataset_rankings %>%
  group_by(classifier) %>%
  summarise(
    mean_rank_f1 = if (use_geom_mean) exp(mean(log(rank_f1), na.rm = TRUE)) else mean(rank_f1, na.rm = TRUE),
    mean_rank_product = if (use_geom_mean) exp(mean(log(rank_product), na.rm = TRUE)) else mean(rank_product, na.rm = TRUE),
    sd_rank_f1 = sd(rank_f1, na.rm = TRUE),
    sd_rank_product = sd(rank_product, na.rm = TRUE),
    n_datasets = n(),
    .groups = "drop"
  ) %>%
  mutate(classifier = factor(classifier,
                             levels = c(rankings_for_palette$classifier))) %>%
  arrange(mean_rank_product)

mean_type <- if (use_geom_mean) "Geometric Mean" else "Arithmetic Mean"
cat(sprintf("\nMean rank per classifier across datasets (%s):\n", mean_type))
print(mean_rank_per_classifier)

# Calculate correlation between F1 rank and product rank
rank_cor_dataset <- cor.test(mean_rank_per_classifier$mean_rank_f1, mean_rank_per_classifier$mean_rank_product, method = "spearman")

cat(sprintf("\nSpearman correlation between F1 and Product rankings (per-dataset): %.3f (p = %.2e)\n", 
            rank_cor_dataset$estimate, rank_cor_dataset$p.value))

# Visualize ranking comparison
rankings_long_dataset <- mean_rank_per_classifier %>%
  select(classifier, mean_rank_f1, mean_rank_product) %>%
  pivot_longer(cols = c(mean_rank_f1, mean_rank_product), names_to = "metric", values_to = "rank") %>%
  mutate(metric = recode(metric, 
                         mean_rank_f1 = "F1 Rank", 
                         mean_rank_product = "Product Rank"))

p_dataset_rank <- ggplot(rankings_long_dataset, aes(x = metric, y = rank, group = classifier, color = classifier)) +
  geom_line(size = 1.2, alpha = 0.7) +
  geom_point(size = 3) +
  scale_y_reverse() +
  scale_color_manual(values = classifier_colors) +
  labs(
    title = "Classifier Ranking per Dataset: F1 vs. Consistency Score",
    subtitle = sprintf("Spearman ρ = %.3f (p = %.2e) [%s]", rank_cor_dataset$estimate, rank_cor_dataset$p.value, mean_type),
    x = "",
    y = "Mean Rank (1 = Best)",
    color = "Classifier"
  ) +
  theme(legend.position = "right")

print(p_dataset_rank)

# Ranking agreement table per dataset
p_dataset_rank_agreement <- ggplot(mean_rank_per_classifier, aes(x = mean_rank_product, y = mean_rank_f1)) +
  geom_point(aes(color = classifier),
             size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  ggrepel::geom_label_repel(aes(label = classifier,
                                color = classifier),
                           size = 5) +
  scale_color_manual(values = classifier_colors) +
  scale_x_reverse() +
  scale_y_reverse() +
  labs(
    title = sprintf("Ranking Agreement per Dataset: ISC vs F1 [%s]", mean_type),
    x = "Mean ISC Rank",
    y = "Mean F1 Rank",
    color = "Classifier"
  ) +
  coord_fixed() +
  ggpubr::theme_classic2() +
  theme(legend.position = "none") 

print(p_dataset_rank_agreement)
```


## Export plots
```{r}
# Save as PDF
pdf_file <- file.path(fig_dir, paste0("LabelProjection_", "ranking.pdf"))
ggsave(
  pdf_file,
  plot = p6,
  width = 5,
  height = 5,
  device = "pdf"
)

```





# Reference consistency


Make a geom_point plot showing the mean of F1 in predictions for each consistency in reference, adding error bar plots to show the variability of the different predictions.

```{r reference_consistency}
# filter for only references

# Summarise prediction F1 per dataset, classifier, and cell type
pred_summary <- cons_combined %>%
  mutate(classifier = factor(classifier,
                             levels = c(rankings_for_palette$classifier))) %>% 
  filter(classifier != "true_labels",
         !is.na(f1),
         classifier %in% top_6_classifiers) %>% 
  select(cell_type, dataset, f1, classifier)

# Join reference consistency with prediction F1
joined <- pred_summary %>%
  left_join(cons_ref, by = c("dataset", "cell_type")) %>%
  filter(!is.na(product)) %>% 
  group_by(dataset, cell_type)

# Correlation per classifier (mean F1 vs reference product)
cor_table <- joined %>%
  group_by(classifier) %>%
  summarise(
    r = cor(f1, product, use = "complete.obs"),
    r2 = r^2,
    p_value = cor.test(f1, product)$p.value,
    n_pairs = sum(!is.na(f1) & !is.na(product)),
    .groups = "drop"
  ) %>%
  ungroup() %>% 
  # Apply FDR correction for multiple testing
  mutate(
    p_value_adjusted = p.adjust(p_value, method = "BH"),
    label = sprintf("R² = %.3f\np = %.2e", r2, p_value)
  )

cat("\nCorrelation (F1 vs reference consistency product) per classifier:\n")
print(cor_table)

# Plot: F1 vs reference consistency product with facets and correlation stats
p_ref <- joined %>% 
  ggplot(aes(x = product, y = f1)) +
  geom_point(aes(color = classifier), alpha = 0.7, size = 2, show.legend = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  scale_color_manual(values = classifier_colors) +
  geom_label(data = cor_table, 
            aes(x = Inf, y = -Inf, label = label),
            fill = "white",
            alpha = 0.4,
            hjust = 1.1, vjust = -0.5, size = 3.5, 
            inherit.aes = FALSE) +
  labs(
    title = "Train ISC vs Prediction F1, per cell type",
    x = "ISC Train",
    y = "F1 (predictions)"
  ) +
  ggpubr::theme_classic2() +
  theme(legend.position = "right") +
  facet_wrap(~classifier)

print(p_ref)

```

## Export plots
```{r}
# Save as PDF
pdf_file <- file.path(fig_dir, paste0("LabelProjection_", "TrainISC.pdf"))
ggsave(
  pdf_file,
  plot = p_ref,
  width = 7,
  height = 5,
  device = "pdf"
)

```

