---
title: "Figure 1: Cell type expression profiles are frequently inconsistent across samples"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# Introduction
Cell type expression profiles are frequently inconsistent across samples

This notebook uses processed files in data_processing/ to render Figure 1 and Extended Data Figure 1 from the manuscript.


# load depencies
```{r, message=F, warning=F}
library(dplyr)
library(ggplot2)
library(tidyr)
library(BiocParallel)
library(pheatmap)
library(RColorBrewer)
library(Seurat)

```

## exported figures directory
```{r}
fig_dir <- "exported_figs/1_Figure"

dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```


## Choose dataset

We choose JoaI KUL5, and Barbry_Leroy2020 from processed datasets. These should have been previously processed using `data_processing` pipeline.

```{r}
# path to output processed results
ds_path <- "../data/processed/"
# Figure 1
ds_files <- list.files(path = ds_path,
                       pattern = "KUL5|Leroy.2020",
                       full.names = T)
# Extended Data Figure 1
ds_files <- c(ds_files, "Yost_Stressed_annotation.rds")
```


## Specificy annotation
Specificy the annotation labels to use in this analysis

```{r}
idents <- c("JoaI" = "cell.type",
            "Barbry_Leroy" = "cell_type",
            "Barbry_Leroy" = "cell_type",
            "Yost" = "celltype"
            )

names(ds_files) <- names(idents)
```

## Load datasets

```{r}
pds <- bplapply(
   unique(names(idents)),
   BPPARAM = MulticoreParam(workers = 3, progressbar = TRUE),
   function(i) {
      
      ident <- idents[[i]]
      
      objects <- lapply(
         ds_files[names(ds_files) == i],
         load_single_cell_object,
         split = FALSE
      )
      
      if (length(objects) == 1) {
         
         object <- objects[[1]]
         rcounts <- Seurat::GetAssayData(object,
                                         assay = "RNA",
                                         layer = "counts")
         counts <- as(rcounts, "dgCMatrix")
         metadata <- as.data.frame(object@meta.data)
         object <- list(
            counts = counts,
            metadata = metadata
         )
         
      } else {
         # merge objects
         object <- merge(objects[[1]], objects[-1])
         rcounts <- Seurat::GetAssayData(object,
                                         assay = "RNA",
                                         layer = "counts")
         counts <- as(rcounts, "dgCMatrix")
         metadata <- as.data.frame(object@meta.data)
         object <- list(
            counts = counts,
            metadata = metadata
         )
      }
      
      sub_meta <- object[["metadata"]] %>%
         dplyr::filter(!is.na(.data[[ident]]))
      
      sub_count_matrix <- object[["counts"]][, rownames(sub_meta), drop = FALSE]
      
      list(
         counts = sub_count_matrix,
         metadata = sub_meta
      )
   }
)

names(pds) <- unique(names(idents))
```

Adapt cell type names of Deprez 2019, as they are too long for plotting

```{r}
library(dplyr)

# Original cell type vector
cell_types <- pds$Barbry_Leroy$metadata$cell_type

# Shortened, biologically meaningful names
short_names <- c(
  "vein endothelial cell" = "VeinEC",
  "club cell" = "Club",
  "multi-ciliated epithelial cell" = "Ciliated",
  "respiratory basal cell" = "Basal",
  "mucus secreting cell" = "Mucus",
  "CD1c-positive myeloid dendritic cell" = "CD1c_mDC",
  "serous secreting cell" = "Serous",
  "ionocyte" = "Ionocyte",
  "nasal mucosa goblet cell" = "NasalGoblet",
  "brush cell of tracheobronchial tree" = "Brush",
  "respiratory hillock cell" = "Hillock",
  "acinar cell" = "Acinar",
  "tracheobronchial serous cell" = "TracheoSerous",
  "elicited macrophage" = "ElicitedMac",
  "CD8-positive, alpha-beta T cell" = "CD8_T",
  "pulmonary artery endothelial cell" = "PA_EC",
  "ciliated columnar cell of tracheobronchial tree" = "ColumnarCiliated",
  "CD4-positive, alpha-beta T cell" = "CD4_T",
  "B cell" = "B",
  "endothelial cell of lymphatic vessel" = "LymphEC",
  "non-classical monocyte" = "ncMono",
  "classical monocyte" = "cMono",
  "dendritic cell" = "DC",
  "bronchus fibroblast of lung" = "BronchFib",
  "natural killer cell" = "NK",
  "plasmacytoid dendritic cell" = "pDC",
  "plasma cell" = "Plasma",
  "tracheobronchial smooth muscle cell" = "SMC",
  "alveolar macrophage" = "AlvMac",
  "lung macrophage" = "LungMac",
  "tracheobronchial goblet cell" = "TracheoGoblet",
  "mast cell" = "Mast",
  "epithelial cell of lower respiratory tract" = "LRT_Epi",
  "stromal cell" = "Stroma",
  "pulmonary alveolar type 2 cell" = "AT2",
  "capillary endothelial cell" = "CapEC",
  "pulmonary alveolar type 1 cell" = "AT1",
  "lung neuroendocrine cell" = "NEC",
  "bronchial goblet cell" = "BronchGoblet",
  "T cell" = "T",
  "conventional dendritic cell" = "cDC",
  "epithelial cell of alveolus of lung" = "AlvEpi",
  "alveolar adventitial fibroblast" = "AlvFib",
  "myofibroblast cell" = "Myofib",
  "hematopoietic stem cell" = "HSC"
)

# Recode the cell types
pds$Barbry_Leroy$metadata$cell_type <- recode(cell_types, !!!short_names)

# Check results
unique(pds$Barbry_Leroy$metadata$cell_type)
```


## Process datasets
We will run the standard [`scTypeEval`](10.18129/B9.bioc.scTypeEval) for reciprocal classification Match and Pseudobulk Cosine dissimilarities.

1) **Pseudobulk PCA**: To focus on inter-sample rather than intra-sample variation, we constructed pseudobulk cell type gene expression profiles by aggregating cells of the same reported cell type and same sample of origin, yielding one transcriptomic profile for each combination of cell type and sample. Principal components analysis (PCA) of these pseudobulk profiles revealed heterogeneous patterns across datasets

2) **Reciprocal Match Map (RMM) are based on recip_classif:Match**: We applied a reciprocal best-hit classification approach. Briefly, for each pair of samples within a dataset, a cell type classifier was trained on cell type profiles from one sample and used to predict cell type labels in the other sample. The two samples were swapped and the procedure repeated. A cell type was considered consistent between a given sample pair only if it constituted a mutual best match under this reciprocal classification scheme; otherwise, it was considered inconsistent. 





```{r}
#black list from scTypeEval package
data(black_list)
bl <- list(black_list$TCR, black_list$Immunoglobulins, black_list$Ygenes) %>% unlist()
# start list of scTypeEval objects
sceval <- list() 
# Define list of dissimilarirites
method.diss <- c("recip_classif:Match",
                 "Pseudobulk:Cosine")

for(i in names(idents)){
   cat("## Computing ", i, " ##\n")
   ident <- idents[[i]]
   
   #create sceval object
   sc <- create_scTypeEval(matrix = pds[[i]][["counts"]],
                           metadata = pds[[i]][["metadata"]],
                           black_list = bl)
   # Filter object and normalize counts
   sc <- run_processing_data(sc,
                            ident = ident,
                            sample = "sample",
                            min_samples = 4,
                            verbose = F)
   # Obtain gene space to compute dissimilarities (HVG)
   sc <- run_hvg(sc,
                 ngenes = 2000,
                 ncores = 6,
                 verbose = F)
   
   set.seed(22)
   # Obtain PCs
   sc <- run_pca(sc,
                 verbose = F)
   
   for(mdiss in method.diss){
      # Obtain dissimilarities
      sc <- run_dissimilarity(sc,
                              method = mdiss,
                              ncores = 8,
                              verbose = F)
   }
   
   sceval[[i]] <- sc
}
rm(sc)
```


Get colors to plot the different cell types

```{r}
annotation_colors <- c(
   "#E41A1C", # red
   "#377EB8", # blue
   "#4DAF4A", # green
   "#984EA3", # purple
   "#FF7F00", # orange
   "#ADFF2F", # chartreuse (yellow-green)
   "#A65628", # brown
   "#F781BF", # light pink
   "#999999", # gray
   "#1B9E77", # dark green-turquoise
   "#D95F02", # dark orange
   "#7570B3", # violet
   "#FFD700", # gold
   "#000080", # navy blue
   "#9932CC", # dark orchid
   "#FF1493", # deep magenta
   "#FF69B4",  # hot pink
   "brown4"
)

```


## RMM

RMM summarizes pairwise comparisons of annotated cell types across all sample pairs within each dataset. For each pair of samples, classifiers were trained in both directions on pseudobulk profiles. Each tile represents the comparison of cell types between two samples and is encoded as follows: black, reciprocal best match in both directions; grey, one-directional match only, consistent with absence or non-recovery of the corresponding cell type in the reciprocal sample; white, no match. Symmetrically, each row and column represent a cell type from one sample.


```{r, fig.height=10, fig.width=10}
dissimilarity.slot = "recip_classif:Match"
sort.consistency = "silhouette"
sort.similarity = "Pseudobulk:Cosine"
low.color = "black"
high.color = "white"
hclust.method = "ward.D2"
verbose = TRUE


ph <- lapply(sceval, function(scTypeEval){
   # check if dissimilarity object are present in object
   diss.assays <- check_dissimilarity_assays(scTypeEval, slot = dissimilarity.slot)
   
   # if not indicated just order cell types based on alphabetical
   if(is.null(sort.similarity) && is.null(sort.consistency)){
      if(verbose){message("No ordering (sort.similarity or sort.consistency) indicated,
                          sorting cell type by alphabetical order.")}
   }
   
   # Compute similarity of cell types if indicated
   if(!is.null(sort.similarity)){
      # get dissimilarity matrix
      diss.assays.cluster <- check_dissimilarity_assays(scTypeEval, slot = sort.similarity)
      assay_cluster <- scTypeEval@dissimilarity[[diss.assays.cluster]]
      d_mat_cluster <- as.matrix(assay_cluster@dissimilarity)
      # get similarities between cell types
      if(verbose){message("Computing cell type similarity based on ", diss.assays.cluster, ".")}
      ident_levels <- levels(assay_cluster@ident[[1]])
      mat <- matrix(0, nrow = length(ident_levels),
                    ncol = length(ident_levels),
                    dimnames = list(ident_levels, ident_levels))
      for (i in ident_levels) {
         for (j in ident_levels) {
            sub <- d_mat_cluster[assay_cluster@ident[[1]] == i,
                                 assay_cluster@ident[[1]] == j,
                                 drop = FALSE]
            mat[i, j] <- mean(sub)
         }
      }
      hc <- stats::hclust(as.dist(mat),
                          method = hclust.method)
      ident_order <- hc$labels[hc$order]
   } else {
      ident_order <- NULL
   }
   
   pls <- lapply(diss.assays,
                 function(da){
                    assay <- scTypeEval@dissimilarity[[da]]
                    title <- paste(da, names(assay@ident), sep = " - ")
                    
                    d_mat_plot <- as.matrix(assay@dissimilarity)
                    
                    annot_ident <- assay@ident[[1]]
                    
                    #  order by consistency if indicated
                    if (!is.null(sort.consistency)) {
                       
                       if(verbose){message("Computing consistency metric for ", sort.consistency, ".\n")}
                       
                       consis <- get_consistency(scTypeEval,
                                                 dissimilarity_slot = da,
                                                 consistency_metric = sort.consistency,
                                                 verbose = FALSE) |>
                          dplyr::filter(!is.na(measure)) |>
                          dplyr::arrange(measure) |>
                          dplyr::pull(measure, name = celltype)
                       
                       if(verbose){message("Consistency computed.")}
                       
                       ident_order <- names(consis)  
                    } 
                    
                    if(is.null(ident_order)){
                       # order alphabetically
                       ident_order <- sort(levels(annot_ident))
                    }
                    
                    # reorder cells within cell type: first by ident order, then cluster within ident
                    if(!is.null(sort.similarity)){
                       ordering <- unlist(lapply(ident_order, function(iden) {
                          cells <- names(annot_ident)[annot_ident == iden]
                          if (length(cells) > 2) {
                             hc <- stats::hclust(as.dist(d_mat_cluster[cells, cells]),
                                                 method = hclust.method)
                             cells[hc$order]
                          } else {
                             cells
                          }
                       }))
                    } else {
                       ordering <- unlist(lapply(ident_order, function(iden) {
                          cells <- names(annot_ident)[annot_ident == iden] |>
                             sort()
                          return(cells)
                       }))
                    }
                    
                    # reorder matrix + annotation
                    d_mat_plot <- d_mat_plot[ordering, ordering, drop = FALSE]
                    
                    # convert matrix to tidy df
                    df <- as.data.frame(as.table(d_mat_plot))
                    colnames(df) <- c("row", "col", "value")
                    
                    # ident boundaries + label positions
                    counts <- as.numeric(table(annot_ident)[ident_order])
                    sep_lines <- cumsum(counts)
                    ident_labels <- ident_order
                    
                    # midpoints for axis ticks
                    x_ticks <- sep_lines - counts/2 + 0.5
                    y_ticks <- length(ordering) - (sep_lines - counts/2 + 0.5) + 1
                    
                    # Create color mapping for cell types
                    unique_celltypes <- sort(unique(ident_order))
                    n_types <- length(unique_celltypes)
                    celltype_colors <- setNames(annotation_colors[1:n_types], unique_celltypes)
                    
                    # Format labels: replace dots with spaces and wrap long names
                    # Show labels for mentioned cell types plus some spacing from others
                    ident_labels_formatted <- setNames(
                       sapply(seq_along(ident_order), function(idx) {
                          label <- ident_order[idx]
                          # Replace dots with spaces
                          label_fmt <- gsub("\\.", " ", label)
                          
                          # do not do it if few cell types
                          if(n_types <= 30){
                             return(label_fmt)
                          }
                          
                          d <- ifelse(n_types > 14, 3, 2)
                          
                          # Include if in incl_celltypes OR every 2nd cell type for spacing
                          if (label %in% incl_celltypes || idx %% d == 1) {
                             # Wrap long names (if longer than 20 chars, add linebreak)
                             if (nchar(label_fmt) > 20) {
                                label_fmt <- gsub("^(.{1,20})\\s", "\\1\n", label_fmt)
                             }
                             return(label_fmt)
                          } else {
                             return("")
                          }
                       }, USE.NAMES = FALSE),
                       ident_order
                    )
                    
                    # Get colors for each y-axis label
                    y_label_colors <- celltype_colors[ident_order]
                    
                    # Create mapping from breaks (Sample_CellType) to labels (cell types)
                    x_breaks <- ordering[x_ticks]
                    y_breaks <- rev(ordering)[y_ticks]
                    
                    # Extract cell type from break names (format: "Sample_CellType")
                    x_celltypes <- sapply(x_breaks, function(name) {
                       parts <- strsplit(name, "_")[[1]]
                       if (length(parts) > 1) {
                          paste(parts[-1], collapse = "_")
                       } else {
                          name
                       }
                    })
                    
                    y_celltypes <- sapply(y_breaks, function(name) {
                       parts <- strsplit(name, "_")[[1]]
                       if (length(parts) > 1) {
                          paste(parts[-1], collapse = "_")
                       } else {
                          name
                       }
                    })
                    
                    # Map to formatted labels
                    x_labels_to_use <- setNames(ident_labels_formatted[x_celltypes], x_breaks)
                    y_labels_to_use <- setNames(ident_labels_formatted[y_celltypes], y_breaks)
                    
                    # build plot
                    p <- ggplot2::ggplot(df, ggplot2::aes(x = col, y = row, fill = value)) +
                       ggplot2::geom_tile() +
                       ggplot2::scale_fill_gradient(low = low.color,
                                                    high = high.color) +
                       ggplot2::scale_x_discrete(
                          breaks = x_breaks,
                          labels = x_labels_to_use,
                          expand = c(0, 0)
                       ) +
                       ggplot2::scale_y_discrete(
                          breaks = y_breaks,
                          labels = y_labels_to_use,
                          expand = c(0, 0)
                       ) +
                       ggplot2::theme_minimal(base_size = 12) +
                       ggplot2::theme(
                          axis.text.x = ggplot2::element_blank(),
                          axis.text.y = ggplot2::element_text(hjust = 1,
                                                              color = celltype_colors[y_celltypes],
                                                              face = "bold",
                                                              size = 28),
                          axis.ticks = ggplot2::element_blank(),
                          axis.title = ggplot2::element_blank(),
                          legend.position = "none",
                          panel.grid = ggplot2::element_blank()
                       ) +
                       ggplot2::geom_hline(yintercept = sep_lines + 0.5,
                                           color = "grey70", linewidth = 0.4) +
                       ggplot2::geom_vline(xintercept = sep_lines + 0.5,
                                           color = "grey70", linewidth = 0.4) 
                    
                    return(p)
                 })
   
   names(pls) <- diss.assays
   
   if(length(pls) == 1){
      pls <- pls[[1]]
   }
   
   return(pls)
})

ph
```



### export heatmaps
```{r}
for(a in names(ph)){
   fpath <- file.path(fdir, paste0(a, "_RMM", ".pdf"))
   pdf(fpath, width = 11, height = 9)
   print(ph[[a]])
   dev.off()
}
```

special case for Barbry, too long

```{r, eval = F}
a = "Barbry_Leroy"
fpath <- file.path(fdir, paste0(a, "_RMM", ".pdf"))
pdf(fpath, width = 13, height = 9)
print(ph[[a]])
dev.off()
```



## auxiliar functions scatter plot

```{r}
highlight_consis <- function(assay,
                             ndim = 30,
                             group_column = "ident"){
   df <- assay@embeddings[1:ndim, ] |>
      t() |>
      as.data.frame() |>
      dplyr::mutate(ident = assay@ident[[1]])
   # Extract the first 'ndim' PCs and the group column
   pc_columns <- names(df)[1:ndim]
   
   # Calculate centroids using the first 'ndim' PCs
   cnt <- df %>%
      group_by(.data[[group_column]]) %>%
      summarize(across(all_of(pc_columns), mean), .groups = "drop")
   
   # Merge centroids to compute distances to own centroid
   df <- df %>%
      left_join(cnt, by = group_column, suffix = c("", "_centroid"))
   
   # Compute Euclidean distance to own centroid
   df <- df %>%
      rowwise() %>%
      mutate(
         dist_to_own_centroid = sqrt(sum((c_across(all_of(pc_columns)) - c_across(ends_with("_centroid")))^2))
      ) %>%
      ungroup()
   
   # Compute distances to other centroids
   r <- df %>%
      rowwise() %>%
      mutate(
         closest_other_dist = min(
            sapply(
               cnt[[group_column]][cnt[[group_column]] != .data[[group_column]]], 
               function(other_group) {
                  other_centroid <- cnt[cnt[[group_column]] == other_group, pc_columns]
                  sqrt(sum((c_across(all_of(pc_columns)) - unlist(other_centroid))^2))
               }
            ),
            na.rm = TRUE
         ),
         is_closer_to_own = dist_to_own_centroid < closest_other_dist,
         fils = ifelse(is_closer_to_own, "consistent", "inconsistent")
      ) %>%
      ungroup() %>% 
      pull(fils)
  
  return(r)
}
```


```{r scatter-helper}
helper.plot.scatter <- function(df,
                                show.legend = FALSE,
                                label = TRUE,
                                dot.size = 5,
                                label.size = 8) {
   # Replace dots with spaces in the ident column
   df_plot <- df |>
      dplyr::mutate(ident_formatted = gsub("\\.", " ", .data[[names(df)[3]]]))
   
   if(is.null(df[["Consistency"]])) {
   pl <- ggplot2::ggplot() +
      ggplot2::geom_point(
         data = df_plot,
         ggplot2::aes(
            x = .data[[names(df)[1]]],  # First column
            y = .data[[names(df)[2]]],  # Second column
            color = .data[[names(df)[3]]]  # Third column (original for coloring)
         ),
         size = dot.size,
         show.legend = show.legend) +
      ggpubr::theme_classic2()
   } else {
      pl <- ggplot2::ggplot() +
         ggplot2::geom_point(
            data = df_plot,
            ggplot2::aes(
               x = .data[[names(df)[1]]],  # First column
               y = .data[[names(df)[2]]],  # Second column
               color = .data[[names(df)[3]]],  # Third column (original for coloring)
               fill = .data[["Consistency"]] # filling for consistency coloring
            ),
            size = dot.size,
            show.legend = show.legend,
            shape = 21) +
         scale_fill_manual(values = c("white", "black")) +
         ggpubr::theme_classic2()
   }
   
   if(label){
      # Calculate centroids and format labels
      cnt <- df_plot |>
         dplyr::group_by(.data[[names(df)[3]]]) |>
         dplyr::summarize(
            x = mean(.data[[names(df)[1]]]),
            y = mean(.data[[names(df)[2]]]),
            ident_formatted = dplyr::first(.data[['ident_formatted']]),
            ident_raw = dplyr::first(.data[[names(df)[3]]])
         )

      cnt_incl <- cnt |>
         dplyr::filter(ident_raw %in% incl_celltypes)

      cnt_other <- cnt |>
         dplyr::filter(!ident_raw %in% incl_celltypes)

      # Filter other labels that are too close to included labels
      if (nrow(cnt_incl) > 0 && nrow(cnt_other) > 0) {
         x_range <- range(df_plot[[names(df)[1]]], na.rm = TRUE)
         y_range <- range(df_plot[[names(df)[2]]], na.rm = TRUE)
         dist_thresh <- 0.05 * max(diff(x_range), diff(y_range))

         cnt_other <- cnt_other |>
            dplyr::rowwise() |>
            dplyr::mutate(
               min_dist = min(sqrt((x - cnt_incl$x)^2 + (y - cnt_incl$y)^2))
            ) |>
            dplyr::ungroup() |>
            dplyr::filter(min_dist > dist_thresh) |>
            dplyr::select(-min_dist)
      }
      
      # Always show labels for mentioned cell types
      pl <- pl +
         ggrepel::geom_label_repel(
            data = cnt_incl,
            ggplot2::aes(
               x = x,
               y = y,
               label = .data[['ident_formatted']],
               color = .data[['ident_raw']]
            ),
            alpha = 0.9,
            size = label.size,
            max.overlaps = Inf,
            show.legend = FALSE
         ) +
         # Show other labels when they do not overlap
         ggrepel::geom_label_repel(
            data = cnt_other,
            ggplot2::aes(
               x = x,
               y = y,
               label = .data[['ident_formatted']],
               color = .data[['ident_raw']]
            ),
            alpha = 0.9,
            size = label.size,
            max.overlaps = 4,
            show.legend = FALSE
         ) +
         ggplot2::theme(legend.text = ggplot2::element_text(size = 12, face = "bold"),
                        axis.text = ggplot2::element_text(
                           size = 16),
                        axis.title = ggplot2::element_text(
                           size = 22,
                           face = "bold")
         )
   }
   
   return(pl)
}
```


## PCA

To focus on inter-sample rather than intra-sample variation, we constructed pseudobulk cell type gene expression profiles by aggregating cells of the same reported cell type and same sample of origin, yielding one transcriptomic profile for each combination of cell type and sample. Principal components analysis (PCA) of these pseudobulk profiles revealed heterogeneous patterns across datasets.

```{r}
reduction.slot = "pseudobulk"
label = TRUE
dims = c(1,2)
show.legend = FALSE
dot.size = 3
highlight = FALSE
dim_centroids = 2

pca <- lapply(sceval, function(scTypeEval){
   red.assays <- check_dim_red_assays(scTypeEval, slot = reduction.slot)
   
   pls <- lapply(red.assays,
                 function(da){
                    assay <- scTypeEval@reductions[[da]]
                    ident <- assay@ident[[1]]
                    title <- paste(da, names(assay@ident), sep = " - ")
                    
                    df <- t(assay@embeddings[dims,]) |>
                       as.data.frame() |>
                       dplyr::mutate(ident = ident)
                    
                    if(highlight){
                       df[["Consistency"]] <- highlight_consis(assay,
                                                   ndim = dim_centroids)
                    } else {
                       df[["Consistency"]] <- NULL
                    }
                    
                    # compute variance of PCs
                    vrs <- var_pca(t(assay@embeddings))[dims]
                    vrs <- round(vrs*100, 2)
                    
                    labs <- paste0("PC", dims, " (", vrs, "%)")
                    
                    set.seed(22)
                    pl <- helper.plot.scatter(df,
                                              show.legend = show.legend,
                                              label = label,
                                              dot.size = dot.size) +
                       ggplot2::labs(x = labs[1],
                                     y = labs[2])
                    
                    # Create color mapping for cell types
                    unique_celltypes <- sort(unique(ident))
                    n_types <- length(unique_celltypes)
                    celltype_colors <- setNames(annotation_colors[1:n_types], unique_celltypes)
                    
                    # Format labels: replace dots with spaces and wrap long names
                    label_mapping <- setNames(
                       sapply(unique_celltypes, function(label) {
                          # Replace dots with spaces
                          label_fmt <- gsub("\\.", " ", label)
                          
                          # Wrap long names (if longer than 20 chars, add linebreak)
                          if (nchar(label_fmt) > 20) {
                             label_fmt <- gsub("^(.{1,20})\\s", "\\1\n", label_fmt)
                          }
                          return(label_fmt)
                       }, USE.NAMES = FALSE),
                       unique_celltypes
                    )
                    
                    # Update the plot with colors and formatted labels
                    pl <- pl +
                       ggplot2::scale_color_manual(values = celltype_colors,
                                                   labels = label_mapping)

                    pl <- pl +
                       ggplot2::theme(
                          axis.text = ggplot2::element_blank(),
                          axis.ticks = ggplot2::element_blank(),
                          axis.line = ggplot2::element_blank()
                       )
                    
                    
                    return(pl)
                 })
   
   names(pls) <- red.assays
   
   if(length(pls) == 1){
      pls <- pls[[1]]
   }
   
   return(pls)
})
pca

ggpubr::ggarrange(plotlist = pca, nrow = 1)
```


### export PCA
```{r}
for(a in names(pca)){
   fpath <- file.path(fdir, paste0(a, "_PCA", ".pdf"))
   pdf(fpath, width = 6, height = 6)
   print(pca[[a]])
   dev.off()
}

```



## UMAP

```{r}
umap <- bplapply(sceval,
                 BPPARAM = MulticoreParam(workers = 3, progressbar = T),
                 function(s){
                    
                    u <- lapply(seq_along(s@reductions), function (r){
                       
                       assay <- s@reductions[[r]]
                       ident <- assay@ident[[1]]
                       
                       um <- uwot::umap(t(assay@embeddings),
                                        seed = 22,
                                        # metric = "cosine",
                                        # min_dist = 0.3,
                                        # n_neighbors = 10
                       )
                       
                       df <- um |>
                          as.data.frame() |>
                          dplyr::mutate(ident = ident)
                       
                       
                       pl <- helper.plot.scatter(df,
                                                 show.legend = F,
                                                 label = T,
                                                 dot.size = 1) +
                          ggplot2::labs(x = "UMAP_1",
                                        y = "UMAP_2")
                       
                       # Get the plot data to extract cell type colors
                       pdata <- ggplot_build(pl)$data[[1]]
                       
                       # Create color mapping for cell types
                       unique_celltypes <- sort(unique(df$ident))
                       n_types <- length(unique_celltypes)
                       celltype_colors <- setNames(annotation_colors[1:n_types], unique_celltypes)
                       
                       # Format labels: replace dots with spaces and wrap long names
                       label_mapping <- setNames(
                          sapply(unique_celltypes, function(label) {
                             # Replace dots with spaces
                             label_fmt <- gsub("\\.", " ", label)
                             
                             # Wrap long names (if longer than 20 chars, add linebreak)
                             if (nchar(label_fmt) > 20) {
                                label_fmt <- gsub("^(.{1,20})\\s", "\\1\n", label_fmt)
                             }
                             return(label_fmt)
                          }, USE.NAMES = FALSE),
                          unique_celltypes
                       )
                       
                       # Update the plot with colored and formatted legend
                       pl <- pl +
                          scale_color_manual(values = celltype_colors,
                                           labels = label_mapping)

                       pl <- pl +
                          ggplot2::theme(
                             axis.text = ggplot2::element_blank(),
                             axis.ticks = ggplot2::element_blank(),
                             axis.line = ggplot2::element_blank()
                          )
                       
                       return(pl)
                    })
                    names(u) <- names(s@reductions)
                    return(u[[1]])
                 })

names(umap) <- names(sceval)

umap
```

### export UMAP
```{r}
for(a in names(umap)){
   fpath <- file.path(fdir, paste0(a, "_UMAPsc", ".pdf"))
   pdf(fpath, width = 6, height = 6)
   print(umap[[a]])
   dev.off()
}

```



# Quantify ISC on heatmap

Quantify ratio of consisent vs inconsistent samples

```{r}
quant_isc <- lapply(sceval, function(s){
   diss_matrix <- s@dissimilarity$`RecipClassif:Match`@dissimilarity %>%
      as.matrix() %>% 
      as.data.frame() %>% 
      tibble::rownames_to_column("A") %>% 
      pivot_longer(-A, 
                   names_to = "B",
                   values_to = "match") %>% 
      separate(A, into = c("SampleA", "celltypeA"), sep = "_") %>% 
      separate(B, into = c("SampleB", "celltypeB"), sep = "_") %>% 
      mutate(label_type = ifelse(celltypeA == celltypeB,
                                 "same_label", "different_label"),
             match = ifelse(match == 0,
                            1, 0)
      )
   quant <- diss_matrix %>%
      group_by(celltypeA, label_type) %>% 
      summarize(ISC = sum(match)/n())
   quant %>% 
      filter(label_type == "same_label") %>% 
      select(celltype = celltypeA,
             ISC)
})
```



# Classical internal validation metrics

Compute and compare internal validation metrics with ISC per celltype

* average silhouette score per cell type: Combined cohesion + separation.
* kNN Label Purity (per cell type): Local neighborhood homogeneity.
* Mean Distance to Centroid (per cell type): Pure cohesion metric
* Davies–Bouldin Index (per cell type): Ratio of within-cluster scatter to between-cluster separation.
* ROGUE
* normalized LISI score

## internal validation functions

```{r}
compute_internal_metrics <- function(embeddings,
                                     labels,
                                     samples,
                                     k = 20,
                                     rogue_counts = NULL) {
   # embeddings: cells x dims
   # labels: vector of celltype labels
   # rogue_counts: optional raw counts matrix (genes x cells) for ROGUE
   if (is.null(embeddings) || nrow(embeddings) < 2) {
      return(data.frame())
   }
   labels <- as.factor(labels)
   keep <- !is.na(labels)
   embeddings <- embeddings[keep, , drop = FALSE]
   labels <- droplevels(labels[keep])
   n <- nrow(embeddings)
   if (n < 2) {
      return(data.frame())
   }
   k_use <- max(1, min(k, n - 1))
   celltypes <- levels(labels)
   
   # --- Silhouette (per cell type) ---
   sil_df <- NULL
   if (libraryNamespace("cluster", quietly = TRUE)) {
      d <- dist(embeddings)
      sil <- cluster::silhouette(as.integer(labels), d)
      sil_df <- as.data.frame(sil) %>%
         mutate(celltype = labels[order(attr(d, "Labels"), na.last = NA)])
   }
   
   # --- kNN label purity (per cell type) ---
   knn_purity <- rep(NA_real_, n)
   if (libraryNamespace("RANN", quietly = TRUE)) {
      nn <- RANN::nn2(embeddings, k = k_use + 1)
      idx <- nn$nn.idx[, -1, drop = FALSE]
      nn_labels <- matrix(labels[idx], nrow = n)
      knn_purity <- rowMeans(nn_labels == labels)
   } else {
      # fallback: brute-force distance
      d <- as.matrix(dist(embeddings))
      diag(d) <- Inf
      idx <- apply(d, 1, function(x) order(x)[1:k_use])
      if (k_use == 1) {
         idx <- matrix(idx, ncol = 1)
      }
      nn_labels <- matrix(labels[idx], nrow = n)
      knn_purity <- rowMeans(nn_labels == labels)
   }
   
   # --- Mean distance to centroid (per cell type) ---
   centroids <- lapply(celltypes, function(ct) {
      rowMeans(embeddings[labels == ct, , drop = FALSE])
   })
   names(centroids) <- celltypes
   mean_dist_centroid <- sapply(celltypes, function(ct) {
      idx <- which(labels == ct)
      if (length(idx) < 2) return(NA_real_)
      ctd <- centroids[[ct]]
      mean(sqrt(rowSums((embeddings[idx, , drop = FALSE] -
                            matrix(ctd, nrow = length(idx), ncol = ncol(embeddings), byrow = TRUE))^2)))
   })
   
   # --- Davies–Bouldin (per cell type) ---
   # Si: mean distance to centroid for cluster i
   # Mij: distance between centroids i and j
   Si <- mean_dist_centroid
   cent_mat <- do.call(rbind, centroids)
   if (length(celltypes) > 1) {
      centroid_dist <- as.matrix(dist(cent_mat))
      dbi <- sapply(seq_along(celltypes), function(i) {
         if (is.na(Si[i])) return(NA_real_)
         ratios <- (Si[i] + Si[-i]) / centroid_dist[i, -i]
         max(ratios, na.rm = TRUE)
      })
      names(dbi) <- celltypes
   } else {
      dbi <- setNames(NA_real_, celltypes)
   }
   
   # --- ROGUE (per cell type) ---
   rogue_scores <- setNames(rep(NA_real_, length(celltypes)), celltypes)
   if (!is.null(rogue_counts) && libraryNamespace("ROGUE", quietly = TRUE)) {
      # ROGUE expects genes x cells; labels vector
      try({
         rogue_scores <- ROGUE::rogue(rogue_counts,
                                      labels = labels,
                                      samples = rep("s", length(labels)),
                                      platform = "UMI") %>% 
            unlist()
      }
      , silent = TRUE)
   }
   
   # --- LISI (Local Inverse Simpson's Index, per cell type) ---
   # Measures local label diversity in kNN neighborhood per cell
   # Normalized to [0,1]: (LISI - 1) / (n_categories - 1)
   lisi_scores <- setNames(rep(NA_real_, length(celltypes)), celltypes)
   if (libraryNamespace("scIntegrationMetrics", quietly = TRUE)) {
      try({
         # compute_lisi expects: X (cells x dims), meta_data (df), label_colnames (char vector)
         meta_df <- data.frame(celltype = labels)
         lisi_full <- scIntegrationMetrics::compute_lisi(embeddings,
                                                         meta_df,
                                                         label_colnames = "celltype",
                                                         perplexity = k)
         # Normalize LISI: (raw_lisi - 1) / (n_categories - 1)
         n_celltypes <- length(unique(labels[!is.na(labels)]))
         lisi_normalized <- (lisi_full$celltype - 1) / (n_celltypes - 1)
         # Average normalized LISI per cell type
         lisi_scores <- tapply(1 - lisi_normalized, labels, mean, na.rm = TRUE)[celltypes]
      }, silent = TRUE)
   }
   
   # assemble per-celltype summary
   res <- data.frame(
      celltype = celltypes,
      silhouette = if (!is.null(sil_df)) tapply(sil_df$sil_width, sil_df$celltype, mean, na.rm = TRUE)[celltypes] else NA_real_,
      knn_purity = tapply(knn_purity, labels, mean, na.rm = TRUE)[celltypes],
      mean_dist_centroid = mean_dist_centroid[celltypes],
      davies_bouldin = dbi[celltypes],
      rogue = rogue_scores[celltypes],
      lisi = lisi_scores[celltypes],
      stringsAsFactors = FALSE
   )
   
   return(res)
}

scale_robust <- function(x, q = c(0.05, 0.95)) {
   lo <- quantile(x, q[1], na.rm = TRUE)
   hi <- quantile(x, q[2], na.rm = TRUE)
   x2 <- pmin(pmax(x, lo), hi)
   1 - (x2 - lo) / (hi - lo)
}
```

## Quantify for each sceval objects
Use PCA single-cell object contian in scTypeEval objects
s@reductions$`single-cell`@embeddings

```{r}
internal_metrics <- lapply(names(sceval), function(nm) {
   s <- sceval[[nm]]
   
   # pick embeddings (prefer single-cell)
   emb <- s@reductions[["single-cell"]]@embeddings
   
   labels <- s@reductions[["single-cell"]]@ident[[1]]
   samples <- s@reductions[["single-cell"]]@sample
   hvg <- s@reductions[["single-cell"]]@gene.list
   
   # optional counts for ROGUE (genes x cells)
   rogue_counts <- NULL
   if (!is.null(s@counts)) {
      rogue_counts <- s@counts[hvg, colnames(emb)]
   }
   
   res <- compute_internal_metrics(embeddings = t(emb),
                                   labels = labels,
                                   samples = samples,
                                   k = 30,
                                   rogue_counts = rogue_counts)
   res$dataset <- nm
   res
})

names(internal_metrics) <- names(sceval)
```

## plotting all quantifications
- Use only correlation for LISI
- Label cell type, that can be shown, especially lowly expressed
### dot plot


```{r correlation-plot}
corr_plots <- lapply(names(internal_metrics), function(i) {
   df <- left_join(
      quant_isc[[i]],
      internal_metrics[[i]],
      by = "celltype"
   ) %>% 
      select(celltype, ISC, cLISI = lisi) %>%
      dplyr::mutate(celltype_formatted = gsub("\\.", " ", celltype))
   
   # Create color mapping for cell types
   unique_celltypes <- sort(unique(df$celltype))
   n_types <- length(unique_celltypes)
   celltype_colors <- setNames(annotation_colors[1:n_types], unique_celltypes)
   
   # Format labels: replace dots with spaces and wrap long names
   label_mapping <- setNames(
      sapply(unique_celltypes, function(label) {
         # Replace dots with spaces
         label_fmt <- gsub("\\.", " ", label)
         
         # Wrap long names (if longer than 20 chars, add linebreak)
         if (nchar(label_fmt) > 20) {
            label_fmt <- gsub("^(.{1,20})\\s", "\\1\n", label_fmt)
         }
         return(label_fmt)
      }, USE.NAMES = FALSE),
      unique_celltypes
   )
   
   # Prepare data for labels
   df_labels <- df |>
      dplyr::mutate(
         in_incl = celltype %in% incl_celltypes,
         celltype_fmt_labeled = celltype_formatted
      )
   
   # Filter other labels that are too close to included labels
   df_incl <- df_labels |> dplyr::filter(in_incl)
   df_other <- df_labels |> dplyr::filter(!in_incl)
   
   if (nrow(df_incl) > 0 && nrow(df_other) > 0) {
      dist_thresh <- 0.08  # distance threshold in plot space
      
      df_other <- df_other |>
         dplyr::rowwise() |>
         dplyr::mutate(
            min_dist = min(sqrt((cLISI - df_incl$cLISI)^2 + (ISC - df_incl$ISC)^2))
         ) |>
         dplyr::ungroup() |>
         dplyr::filter(min_dist > dist_thresh) |>
         dplyr::select(-min_dist)
   }
   
   set.seed(22)
   p <- ggplot(df, aes(cLISI, ISC)) +
      geom_jitter(aes(color = celltype),
                  width = 0.05,
                  height = 0.05,
                  size = 5,
                 alpha = 0.6,
                 show.legend = F) +
      # Always show labels for included cell types
      ggrepel::geom_label_repel(
         data = df_incl,
         ggplot2::aes(
            label = celltype_fmt_labeled,
            color = celltype
         ),
         alpha = 0.7,
         size = 8,
         max.overlaps = Inf,
         show.legend = FALSE
      ) +
      # Show other labels when they do not overlap with included
      ggrepel::geom_label_repel(
         data = df_other,
         ggplot2::aes(
            label = celltype_fmt_labeled,
            color = celltype
         ),
         alpha = 0.7,
         size = 8,
         max.overlaps = 8,
         show.legend = FALSE
      ) +
      scale_color_manual(values = celltype_colors,
                        labels = label_mapping) +
      labs(
         x = "cLISI",
         y = "Reciprocal hit proportion",
         color = "Internal validation metric"
      ) +
      scale_x_continuous(
         breaks = c(0, 1),
         labels = c("0", "1"),
         expand = expansion(mult = c(0, 0.1))
      ) +
      scale_y_continuous(
         breaks = c(0, 1),
         labels = c("0", "1"),
         expand = expansion(mult = c(0, 0.1))
      ) +
      coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
      ggpubr::theme_classic2() +
      theme(axis.text = ggplot2::element_text(
                           size = 16),
            axis.ticks = ggplot2::element_blank(),
            axis.title = ggplot2::element_text(
                           size = 22,
                           face = "bold"),
            legend.text = ggplot2::element_text(size = 12, face = "bold")
            )
   
   p
})

names(corr_plots) <- names(internal_metrics)
corr_plots
```


## export quantification plots

```{r}
for(a in names(corr_plots)){
   fpath <- file.path(fdir, paste0(a, "_Quantification_IVM", ".pdf"))
   pdf(fpath, width = 6, height = 6)
   print(corr_plots[[a]])
   dev.off()
}
```















