---
title: "Figure 4: Consistency metrics on supervised cell type classification performance"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# Introduction

This script analyzes the results of the label transfer consistency benchmarking pipeline (`label_transfer_task` pipeline) in a supervised cell type classification performance. 
Exported plots are included in the manuscript as Main Figure 4 and Extended Data Figure 3. 

Building on the benchmarking framework established by Abdelaal et al., 2019 and implemented as part of the Open problems in single-cell analysis (Luecken et al., 2025), we benchmarked 13 classifiers across a diverse collection of datasets (`data_processing/config/specs_datasets.csv`). We implemented two complementary evaluation schemes: (i) intra-dataset, using sample-aware train-test splits to assess generalization across biological replicates, and (ii) inter-dataset, training on one dataset and testing on an independent dataset to emulate cross-study label transfer

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

```{r libraries}
require(ggplot2)
require(dplyr)
require(tidyr)
require(ggpubr)
require(lme4)
source("helpers/supervised_classification.R")
# Custom theme
theme_set(theme_bw(base_size = 12))
```

```{r exporting_dir}
fig_dir <- file.path(
  "~/OneDrive - Université de Genève/CSI/LabPapers/Intersample_Consistency/Figures/plots/Figure4/"
)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

Load data from results (within dataset splitting) and results_between (prediction between different datasets but same annotation) obtained previously in label_transfer_task pipeline.

Add function to collapse replicates
```{r}
iscs <- c("X2label_silhouette...Pseudobulk.Cosine",
          "silhouette...recip_classif.Match",
          "product",
          "SCCAF",
          "MetaNeighbor_Supervised...NA",
          "nsa_cLISI...NA",
          "hotelling",
          "cramer")

labs <- c("ISC Global",
          "ISC Local",
          "ISC Product",
          "SCCAF",
          "MetaNeighbor Supervised",
          "cLISI",
          "scDiagnostics-Hotelling",
          "scDiagnostics-Cramer")
```

Some dataset show very high F1 and isc metrics, making correlation not meaningful

```{r}
excl_datasets <- c("JoaI_CRC.SG1_Normal",
                   "JoaI_CRC.SG2_Tumor",
                   "JoaI_KUL3_Tumor",
                   "JoaI_SMC_Normal")
excl_pat <- paste(excl_datasets, collapse = "|") %>% 
  gsub("[.]", "[.]", .)
```



```{r load_data}
paths <- 
  c(
    "intra_dataset" = file.path("..", "label_transfer_task", "results", "aggregated", "label_transfer_consistency_detailed.csv"),
    "inter_dataset" = file.path("..", "label_transfer_task", "results_between", "aggregated", "label_transfer_between_consistency_detailed.csv")
  )
# paths <- 
#   c(
#     "intra_dataset" = "../label_transfer_task/archive/results/aggregated/label_transfer_consistency_detailed.csv",
#     "inter_dataset" = "../label_transfer_task/archive/results_between/aggregated/label_transfer_between_consistency_detailed.csv")
  
res0 <- lapply(paths, read.csv)

res_raw <- lapply(res0, function(x){
    x$product <- x[["X2label_silhouette...Pseudobulk.Cosine"]] * x[["silhouette...recip_classif.Match"]]
    return(x)
})

res_raw$inter_dataset <- res_raw$inter_dataset %>% 
  separate(dataset_id, sep = "__TO__",
           into = c("dataset_ref", "dataset_test"),
           remove = F)

res <- lapply(res0, function(x){
  x <- x %>% 
    filter(!grepl(excl_pat, dataset_id))
  return(x)
})
res$intra_dataset <- collapse_replicates(res$intra_dataset)
# Preview
lapply(res, head)
lapply(iscs, function(x){
  table(is.na(res$intra_dataset[[x]]))
}) %>% 
  setNames(iscs)
```

# Helper functions

Modular helpers reused by the sections below. Each takes the ISC metric column (`isc_col`) as a parameter so analyses can be looped over all `iscs`. `excl` controls which classifiers (e.g. "Random", "Ensemble") are dropped before computing correlations/rankings.

```{r isc_helpers}
# Set to FALSE to keep "Random" and "Ensemble" classifiers in correlation/ranking analyses
exclude_random_ensemble <- TRUE
excl <- if (exclude_random_ensemble) c("ground_truth", "Random", "Ensemble") else "ground_truth"
```

# Per classifier
Show mean F1 vs mean ISC, looped over every metric in `iscs`.
```{r f1_vs_product_perClassifier}
perClassifier_results <- lapply(iscs, function(isc_col){
  isc_label <- labs[match(isc_col, iscs)]
  lapply(names(res), function(r){
    plot_isc_vs_f1_per_classifier(
      df = res[[r]],
      isc_col = isc_col,
      palette = "f1",
      excl = excl,
      dataset_label = r
    )
  }) %>% setNames(names(res))
}) %>% setNames(iscs)

# Preview for the default metric (product)
lapply(perClassifier_results, function(e){
  lapply(e, function(i){
    print(i[["plot"]])
  })
})
```


# Plot multiple ISC
Overlay correlation lines for every ISC metric in `iscs` on a single plot per dataset.

```{r isc_names}
isc_colors <- c(
  "ISC Global" = "#0173B2",             # blue
  "ISC Local" = "#D55E00",              # orange
  "ISC Product" = "#029E73",            # green
  "SCCAF" = "#CC78BC",                  # purple
  "MetaNeighbor Supervised" = "brown4", # brown
  "cLISI" = "#949494",                  # grey
  "scDiagnostics-Hotelling" = "#56B4E9",# sky blue
  "scDiagnostics-Cramer" = "#CA9161"    # tan
)
```

```{r corr_different_isc, fig.height=10, fig.width=5}
multi_isc_results <- lapply(names(res), function(r){
  plot_multi_isc_corr(res[[r]], iscs = iscs, labs = labs, isc_colors = isc_colors,
                       excl = excl, dataset_label = r)
}) %>% setNames(names(res))

multi_isc_results
```


# Compare ranking of classifiers either using F1 or an ISC metric, looped over `iscs`

```{r ranking_comparison}
ranking_comparison_results <- lapply(iscs, function(isc_col){
  lapply(names(res), function(r){
    plot_ranking_comparison(res[[r]], isc_col = isc_col,
                             excl = excl, dataset_label = r)
  }) %>% setNames(names(res))
}) %>% setNames(iscs)

ranking_comparison_results
```

# ranking mean
Per-dataset ranking of classifiers based on each ISC metric, then average rank across datasets, looped over `iscs`.

```{r ranking_per_dataset}
# Parameter: use geometric mean (TRUE) or arithmetic mean (FALSE)
use_geom_mean <- TRUE

ranking_per_dataset_results <- lapply(iscs, function(isc_col){
  lapply(names(res), function(r){
    plot_ranking_per_dataset(res[[r]], isc_col = isc_col,
                              excl = excl, use_geom_mean = use_geom_mean,
                             dataset_label = r)
  }) %>% setNames(names(res))
}) %>% setNames(iscs)

ranking_per_dataset_results
```


# ISC vs F1 scatter grouped by cell type / classifier, per ISC metric and dataset

For every ISC metric and dataset (intra/inter), plot mean F1 vs mean ISC grouped (colored) either by `cell_type` or by `classifier`, and collect the R² / p-value of each fit into a single summary table.

```{r isc_vs_f1_grouped_all}
group_vars <- list(celltype = c("classifier", "cell_type"),
                  classifier = "classifier")

grouped_results <- lapply(group_vars, function(gv){
  lapply(iscs, function(isc_col){
    isc_label <- labs[match(isc_col, iscs)]
    lapply(names(res), function(r){
      plot_isc_vs_f1_grouped(
        df = res[[r]],
        isc_col = isc_col,
        isc_label = isc_label,
        group_by = gv,
        excl = excl,
        dataset_label = paste0(r, " - ", paste(gv, collapse = ":")),
        palette = "f1"
      )
    }) %>% setNames(names(res))
  }) %>% setNames(iscs)
}) %>% setNames(names(group_vars))

lapply(grouped_results, function(a){
  lapply(a, function(b){
    lapply(b, function(i){
      print(i[["plot"]])
    })
  })
})

```
### explore datasets
```{r, fig.width=20, fig.height=10}
gv = c("classifier", "cell_type")

perdataset <- lapply(seq_along(iscs), function(it){
  isc_col = iscs[it]
  isc_label = labs[it]
  plot_isc_vs_f1_grouped(
    df = res$intra_dataset,
    isc_col = isc_col,
    isc_label = isc_label,
    group_by = gv,
    excl = excl,
    dataset_label = paste0(r, " - ", paste(gv, collapse = ":")),
    palette = "f1"
  )[["plot"]] +
    facet_wrap(~dataset_id)
})
perdataset
```



```{r}

# Combine R2 / p-value stats for every (group_by x isc x dataset) combination
isc_vs_f1_stats_table <- do.call(rbind, lapply(names(group_vars), function(gv){
  do.call(rbind, lapply(iscs, function(isc_col){
    do.call(rbind, lapply(names(res), function(r){
      grouped_results[[gv]][[isc_col]][[r]]$stats
    }))
  }))
})) %>%
  mutate(isc_label = factor(isc_label, levels = labs)) %>%
  arrange(desc(r2), dataset)
rownames(isc_vs_f1_stats_table) <- NULL

isc_vs_f1_stats_table
```


# plot r2 per dataset

```{r plot_r2_per_dataset}
# Per-replicate data (not collapsed) with the product metric, to show variability across replicates
inp <- res
r2_per_dataset <- lapply(names(inp), function(r){
  lapply(iscs, function(isc_col){
    compute_r2_per_dataset_celltype(inp[[r]],
                                    group_cols = c("dataset_id"),
                                    isc_col = isc_col,
                                    excl = excl)
  }) %>%
    bind_rows() %>%
    mutate(dataset_group = r)
}) %>%
  bind_rows() %>%
  mutate(isc_label = factor(labs[match(isc, iscs)], levels = labs))

mr2 <- r2_per_dataset %>% 
  group_by(dataset_group, isc_label) %>% 
  summarize(mean = mean(r2, na.rm = T),
            median = median(r2, na.rm = T))

r2_per_dataset_plots <- lapply(unique(r2_per_dataset$dataset_group),
                               function(dg){
                                 r2_per_dataset %>%
                                   filter(dataset_group == dg) %>% 
                                   ggplot(aes(x = r2,
                                              y = forcats::fct_reorder(isc_label,
                                                                       r2,
                                                                       .fun = median))) +
                                   geom_boxplot(aes(fill = isc_label),
                                                outlier.shape = NA,
                                                alpha = 0.4,
                                                show.legend = F) +
                                   geom_jitter(width = 0,
                                               height = 0.1,
                                               size = 1,
                                               alpha = 0.2) +
                                   labs(
                                     title = dg,
                                     subtitle = "R² per dataset (cell-type level F1 vs ISC in Test)",
                                     y = "", x = expression(R^2)
                                   ) +
                                   ggpubr::theme_classic2()
                               }) %>% setNames(unique(r2_per_dataset$dataset_group))

r2_per_dataset_plots
```

## export plots

```{r}
pdf_file <- file.path(fig_dir, paste0("Test_R2_", "boxplot.pdf"))

pdf(pdf_file, width = 3.5, height = 2.5)
for(a in r2_per_dataset_plots) print(a)
dev.off()
```


Pearson R² can be inflated by a handful of points sitting near the ISC ceiling (values clustering at 0.94-0.95 when the metric is theoretically bounded in [0,1]); those act as high-leverage points on the linear fit. Compare against Spearman (rank-based, insensitive to this) and flag dataset/ISC combos with influential points (Cook's distance > 4/n) to see where Pearson is likely overstating the relationship.

```{r r2_per_dataset_diagnostics}
r2_diagnostics <- r2_per_dataset %>%
  mutate(r2_divergence = r2 - r2_spearman) %>%
  select(dataset_group, dataset_id, isc_label, r2, r2_spearman, r2_divergence, n_influential, n) %>%
  arrange(desc(r2_divergence))

# Flag combos where Pearson is likely inflated: large gap vs Spearman and/or influential points present
r2_diagnostics %>% filter(r2_divergence > 0.15 | n_influential > 0)

r2_vs_spearman_plots <- lapply(unique(r2_per_dataset$dataset_group), function(dg){
  r2_per_dataset %>%
    filter(dataset_group == dg) %>%
    select(dataset_id, isc_label, Pearson = r2, Spearman = r2_spearman) %>%
    pivot_longer(c(Pearson, Spearman), names_to = "method", values_to = "r2") %>%
    ggplot(aes(x = r2, y = forcats::fct_reorder(isc_label, r2, .fun = median), fill = method)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5, position = position_dodge(0.8)) +
    geom_point(position = position_jitterdodge(jitter.width = 0, jitter.height = 0.1, dodge.width = 0.8),
               size = 1.2, alpha = 0.4) +
    labs(title = dg, subtitle = "Pearson vs Spearman R² per dataset", y = "Tool", x = expression(R^2), fill = "Method") +
    ggpubr::theme_classic2()
}) %>% setNames(unique(r2_per_dataset$dataset_group))

r2_vs_spearman_plots
```

A metric compressed into a narrow observed range (e.g. cLISI often sits within ~0.75-1) has far less room to vary than one spanning the full 0-1 range like ISC Local (which mirrors F1's own range). R² values are therefore not directly comparable across ISC metrics with very different observed ranges: a narrow-range metric needs a much tighter (and more leverage-sensitive) fit to reach the same R² as a full-range one. Check `isc_range`/`isc_sd` next to `r2` before concluding a narrow-range metric with a high R² is actually a better predictor.

```{r r2_range_restriction}
range_vs_r2 <- r2_per_dataset %>%
  select(dataset_group, dataset_id, isc_label, r2, r2_spearman, isc_min, isc_max, isc_range, isc_sd)

range_vs_r2 %>% arrange(isc_range)

ggplot(range_vs_r2, aes(x = isc_range, y = r2, color = isc_label)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  facet_wrap(~dataset_group) +
  labs(
    title = "R² vs observed ISC range per dataset",
    subtitle = "Narrow-range metrics (left) need a tighter fit to reach the same R² as full-range ones",
    x = "Observed ISC range (max - min)", y = expression(R^2), color = "ISC metric"
  ) +
  ggpubr::theme_classic2()
```

```{r, fig.width=10, fig.height=10}
ds <- "Stephenson_Cambridge_Healthy"
isc_col <- "silhouette...recip_classif.Match"
df <- res_raw$intra_dataset
t <- df %>% 
  filter(dataset_id == ds,
         !classifier %in% excl,
         !is.na(f1) & !is.na(.data[[isc_col]]))

plot(t[[isc_col]], t$f1)
cor(t[[isc_col]], t$f1, use = "complete.obs")^2
cor_stats(t[[isc_col]], t$f1)
```


```{r}
m_isc  <- lm(f1 ~ silhouette...recip_classif.Match,  data = t)
m_lisi <- lm(f1 ~ nsa_cLISI...NA, data = t)

summary(m_isc)$r.squared
summary(m_lisi)$r.squared
cor(t[["f1"]], t[["silhouette...recip_classif.Match"]],  use = "complete.obs")^2
cor(t[["f1"]], t[["nsa_cLISI...NA"]], use = "complete.obs")^2
```


# Mixed-effects model

A single-dataset `lm(f1 ~ isc)` as above ignores that observations are repeated across classifiers and cell types within the same dataset, and ignores classifier identity entirely when pooling across datasets - both can inflate the apparent ISC-F1 relationship. Account for both by fitting one mixed-effects model per ISC metric.

### Main analysis
For each ISC metric, fit a mixed-effects linear model where F1 is predicted by the ISC metric while accounting for classifier and repeated observations from the same dataset.

The basic model is conceptually:

```
lmer(f1 ~ ISC + classifier + (1 | dataset), data = t)
```
where `ISC` is replaced by each metric in `iscs`.

```{r lmer_main_analysis}
lmer_results <- lapply(names(res_raw), function(r){
  lapply(iscs, function(isc_col){
    fit_isc_f1_lmer(res[[r]], isc_col = isc_col, excl = excl)
  }) %>% bind_rows()
}) %>%
  setNames(names(res)) %>%
  bind_rows(.id = "dataset_group") %>%
  mutate(isc_label = factor(isc_label, levels = labs))

lmer_results
```

```{r lmer_main_analysis_forest, fig.width=6, fig.height=5}
lmer_forest_plots <- lapply(unique(lmer_results$dataset_group), function(dg){
  lmer_results %>%
    filter(dataset_group == dg) %>%
    ggplot(aes(x = estimate, y = forcats::fct_reorder(isc_label, estimate), color = pval < 0.05)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.2) +
    geom_point(size = 2) +
    scale_color_manual(values = c("TRUE" = "#0173B2", "FALSE" = "grey60"), name = "p < 0.05") +
    labs(
      title = dg,
      subtitle = "F1 ~ scale(ISC) + classifier + (1 | dataset_id)",
      x = "ISC fixed-effect estimate per 1 SD (95% Wald CI)",
      y = "ISC metric"
    ) +
    ggpubr::theme_classic2()
}) %>% setNames(unique(lmer_results$dataset_group))

lmer_forest_plots
```



# Reference consistency

# Consistency table
Make a geom_point plot showing the mean of F1 in predictions for each consistency in reference, adding error bar plots to show the variability of the different predictions.

Same rationale as the Test-set ISC analysis applies here: report Pearson alongside Spearman (ceiling/leverage effects) and the observed ISC range (restriction-of-range) rather than a single Pearson R².

```{r}
iscs_ref <- c("X2label_silhouette...Pseudobulk.Cosine",
          "silhouette...recip_classif.Match",
          "product",
          "SCCAF",
          "MetaNeighbor_Supervised...NA",
          "nsa_cLISI...NA")

labs_ref <- c("ISC Global",
          "ISC Local",
          "ISC Product",
          "SCCAF",
          "MetaNeighbor Supervised",
          "cLISI") 
```


```{r reference_consistency_overall}

# filter for only references
cons_ref_all <- 
  lapply(names(res), function(r){
    joined <- join_reference_consistency(res[[r]], isc_col = iscs_ref, excl = excl)
    lapply(iscs_ref, function(isc){
      # Overall correlation (Pearson + Spearman), with the observed ISC range for context
      cor_table <- cor_stats(joined[[isc]][[isc]], joined[[isc]]$f1) %>%
        mutate(label = sprintf("R²(P)=%.3f R²(S)=%.3f\np=%.2e, range=%.2f",
                               r2, r2_spearman, pval, isc_range))
      
      cat("\nCorrelation (F1 vs reference consistency", isc, ") overall:\n")
      print(cor_table)
      
      # Plot: F1 vs reference consistency product with facets and correlation stats
      p_ref <- joined[[isc]] %>% 
        ggplot(aes(x = .data[[isc]], y = f1)) +
        geom_point(aes(color = classifier), alpha = 0.7, size = 2, show.legend = FALSE) +
        geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
        geom_label(data = cor_table, 
                   aes(x = Inf, y = -Inf, label = label),
                   fill = "white",
                   alpha = 0.4,
                   hjust = 1.05, vjust = -0.2, size = 3, 
                   inherit.aes = FALSE) +
        labs(
          title = r,
          subtitle = "Train ISC (reference) vs Prediction F1, per cell type",
          x = paste0("Value Train: ", labs_ref[match(isc, iscs_ref)]),
          y = "F1 (predictions)" 
        ) +
        ggpubr::theme_classic2() +
        theme(legend.position = "right")
      
      return(p_ref)
    })
  }) 

cons_ref_all
```

## correlation per classifier

```{r reference_consistency_perclassifier}
# filter for only references
cons_ref <- 
  lapply(iscs_ref, function(isc){
    lapply(names(res), function(r){
      # Summarise prediction F1 per dataset, classifier, and cell type
      joined <- join_reference_consistency(res[[r]], isc_col = isc, excl = excl)[[isc]]
      
      # Correlation per classifier (Pearson + Spearman vs reference ISC)
      cor_table <- joined %>%
        group_by(classifier) %>%
        group_modify(~ cor_stats(.x[[isc]], .x$f1)) %>%
        ungroup() %>%
        mutate(label = sprintf("R²(P)=%.2f R²(S)=%.2f\np=%.1e", r2, r2_spearman, pval))
      
      # Plot: F1 vs reference consistency product with facets and correlation stats
      p_ref <- joined %>% 
        ggplot(aes(x = .data[[isc]], y = f1)) +
        geom_point(aes(color = classifier), alpha = 0.7, size = 2, show.legend = FALSE) +
        geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
        geom_label(data = cor_table, 
                   aes(x = Inf, y = -Inf, label = label),
                   fill = "white",
                   alpha = 0.4,
                   hjust = 1.05, vjust = -0.2, size = 2.8, 
                   inherit.aes = FALSE) +
        labs(
          title = r,
          subtitle = "Train ISC (reference) vs Prediction F1, per cell type",
          x = paste0("Value Train: ", isc),
          y = "F1 (predictions)"
        ) +
        ggpubr::theme_classic2() +
        theme(legend.position = "right") +
        facet_wrap(~classifier)
      
      return(p_ref)
    })
    
  })

cons_ref
```

## scatter and R² boxplot across all ISC metrics

For every metric in `iscs`, join the ground_truth reference value with prediction F1 (per dataset) and look at both the scatter and the per-dataset R² distribution - a metric can look fine in one dataset and be driven by leverage/restricted range in another.


```{r reference_consistency_r2_boxplot}
reference_r2_per_dataset <- lapply(names(res), function(r){
  lapply(iscs_ref, function(isc_col){
    compute_r2_per_dataset_celltype(reference_joined[[r]][[isc_col]],
                                    isc_col = isc_col,
                                    group_cols = c("dataset_id"),
                                    excl = excl)
  }) %>% bind_rows()
}) %>% setNames(names(res)) %>%
  bind_rows(.id = "dataset_group") %>%
  mutate(isc_label = factor(labs_ref[match(isc, iscs_ref)], levels = labs_ref))

reference_r2_boxplots <- lapply(unique(reference_r2_per_dataset$dataset_group), function(dg){
  reference_r2_per_dataset %>%
    filter(dataset_group == dg) %>%
    ggplot(aes(x = r2, y = forcats::fct_reorder(isc_label, r2, .fun = median))) +
    geom_boxplot(aes(fill = isc_label), outlier.shape = NA, alpha = 0.4, show.legend = FALSE) +
    geom_jitter(width = 0, height = 0.1, size = 1.5, alpha = 0.4) +
    labs(
      title = dg,
      subtitle = "R² per dataset (reference ISC vs prediction F1)",
      y = "Tool", x = expression(R^2)
    ) +
    ggpubr::theme_classic2()
}) %>% setNames(unique(reference_r2_per_dataset$dataset_group))

reference_r2_boxplots
```

## mixed-effects model

Same rationale as the Test-set ISC main analysis: pool across datasets and classifiers properly by fitting `f1 ~ ISC + classifier + (1 | dataset_id)` per metric, here using the reference (ground_truth) ISC value joined to each classifier's prediction F1.

```{r reference_lmer_main_analysis}
reference_joined <- lapply(res, function(r) {
  join_reference_consistency(r, isc_col = iscs_ref, excl = excl) 
  })
fml <- list(intra_dataset = ~ classifier + (1 | dataset_id) + (1 | dataset_id:replicate),
            inter_dataset = ~ classifier + (1 | dataset_ref:dataset_test))
reference_lmer_results <- lapply(names(res_raw), function(r){
  lapply(iscs_ref, function(isc_col){
    fit_isc_f1_lmer(reference_joined[[r]][[isc_col]],
                    isc_col = isc_col,
                    excl = excl,
                    model = fml[[r]])
  }) %>% bind_rows()
}) %>%
  setNames(names(res)) %>%
  bind_rows(.id = "dataset_group") %>%
  mutate(isc_label = factor(isc_label, levels = labs))

reference_lmer_results
```

```{r reference_lmer_main_analysis_forest, fig.width=6, fig.height=5}
reference_lmer_forest_plots <- lapply(unique(reference_lmer_results$dataset_group), function(dg){
  reference_lmer_results %>%
    filter(dataset_group == dg) %>%
    ggplot(aes(x = estimate, y = forcats::fct_reorder(isc_label, estimate), color = pval < 0.05)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.2) +
    geom_point(size = 2) +
    scale_color_manual(values = c("TRUE" = "#0173B2", "FALSE" = "grey60"), name = "p < 0.05") +
    labs(
      title = dg,
      subtitle = "F1 ~ scale(Reference ISC) + classifier + (1 | dataset_id)",
      x = "ISC fixed-effect estimate per 1 SD (95% Wald CI)",
      y = "ISC metric"
    ) +
    ggpubr::theme_classic2()
}) %>% setNames(unique(reference_lmer_results$dataset_group))

reference_lmer_forest_plots
```


## categorize ISC

```{r}
excl <- c("ground_truth", "Ensemble", "Random")
ths <- 0.7
# filter for only references
cons_ref_cat <- lapply(names(res), function(r){
  # Summarise prediction F1 per dataset, classifier, and cell type
  class_levs <- names(rankings_palette(res[[r]], "f1"))[!names(rankings_palette(res[[r]], "f1")) %in% excl]
  top_6 <- c(head(class_levs, 3), tail(class_levs,3))
  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier,
                               levels = c(class_levs))) %>% 
    filter(classifier != "ground_truth",
           !is.na(f1),
           classifier %in% top_6) %>% 
    select(cell_type, dataset_id, f1, classifier)
  
  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth",
           !is.na(.data[[isc]])) %>% 
    select(cell_type, dataset_id, .data[[isc]]) %>% 
    mutate(consistency = ifelse(.data[[isc]] > ths,
                              "consistent",
                              "inconsistent"))
  
  # Join reference consistency with prediction F1
  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(.data[[isc]])) %>% 
    group_by(dataset_id, cell_type)
  
  
  
  # Plot: F1 vs reference consistency product with facets and correlation stats
  p_ref <- joined %>% 
    ggplot(aes(x = consistency, y = f1)) +
    geom_jitter(alpha = 0.3, color = "grey", width = 0.15) +
    geom_boxplot(aes(color = classifier),
                 fill = NA,
                 outlier.colour = NA,
                 show.legend = FALSE) +
    scale_color_manual(values = rankings_palette(res[[r]], "f1")) +
    labs(
      title = r,
      subtitle = paste0("F1 in consistent (ISC>", ths,
                        "), and incosistent cell types in train set"),
      x = "ISC category",
      y = "F1 (predictions)"
    ) +
    ggpubr::theme_classic2() +
    facet_wrap(~classifier)
  
  return(p_ref)
})

ggarrange(plotlist = cons_ref_cat)
```


## contigency plot

```{r density_contigency}
ths_f1 <- 0.7
ths_isc <- 0.5

# One contingency plot per ISC metric and per dataset group
contingency_results <- lapply(names(res), function(r){
  joined_r <- join_reference_consistency(res[[r]], excl = excl)
  lapply(iscs_ref, function(isc_col){
    ths_isc <- quantile(joined_r[[isc_col]][[isc_col]], 0.5)
    plot_isc_f1_contingency(joined_r[[isc_col]], isc_col = isc_col,
                            ths_isc = ths_isc, ths_f1 = ths_f1, dataset_label = r)
  }) %>% setNames(iscs_ref)
}) %>% setNames(names(res))

# Render one grid of plots (across all ISC metrics) per dataset group
lapply(names(contingency_results), function(r){
  ggarrange(plotlist = lapply(contingency_results[[r]], `[[`, "plot"))
})
```

## quadrant quantification across all ISC metrics

```{r quadrant_quantification_all_isc}
quadrant_quantification <- lapply(names(contingency_results), function(r){
  lapply(contingency_results[[r]], `[[`, "quadrant_counts") %>% bind_rows()
}) %>%
  setNames(names(contingency_results)) %>%
  bind_rows(.id = "dataset_group") %>%
  mutate(isc_label = factor(isc_label, levels = labs))

quadrant_quantification %>%
  select(dataset_group, isc_label, quadrant, n, prop) %>%
  arrange(dataset_group, isc_label, quadrant)

ggplot(quadrant_quantification, aes(x = isc_label, y = prop, fill = quadrant)) +
  geom_col(position = "stack") +
  geom_text(aes(label = paste0(sprintf("%.0f", 100 * prop), "%")),
            position = position_stack(vjust = 0.5), size = 2.8) +
  facet_wrap(~dataset_group) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Quadrant composition per ISC metric",
    subtitle = sprintf("Reference ISC vs Prediction F1 (thresholds: ISC=%.1f, F1=%.1f)", ths_isc, ths_f1),
    x = "ISC metric", y = "Proportion of cell type / classifier pairs", fill = "Quadrant"
  ) +
  ggpubr::theme_classic2() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r upper_left_quadrant_depletion}
# Test whether upper-left quadrant is depleted vs 25% null expectation
quadrant_stats_plots <- lapply(names(res), function(r){
  class_levs <- names(rankings_palette(res[[r]], "f1"))[!names(rankings_palette(res[[r]], "f1")) %in% excl]

  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier, levels = c(class_levs))) %>%
    filter(classifier != "ground_truth", !is.na(f1)) %>%
    select(cell_type, dataset_id, f1, classifier)

  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth", !is.na(.data[[isc]])) %>%
    select(cell_type, dataset_id, .data[[isc]])

  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(.data[[isc]]))

  quad_tbl <- joined %>%
    mutate(quadrant = case_when(
      .data[[isc]] >= ths_isc & f1 >= ths_f1 ~ "top-right",
      .data[[isc]] <  ths_isc & f1 >= ths_f1 ~ "top-left",
      .data[[isc]] >= ths_isc & f1 <  ths_f1 ~ "bottom-right",
      TRUE                              ~ "bottom-left"
    )) %>%
    count(quadrant, .drop = FALSE)

  all_quadrants <- c("top-left", "top-right", "bottom-left", "bottom-right")
  quad_tbl <- dplyr::full_join(
    tibble::tibble(quadrant = all_quadrants),
    quad_tbl,
    by = "quadrant"
  ) %>%
    mutate(n = ifelse(is.na(n), 0L, n))

  n_total <- sum(quad_tbl$n)
  n_ul <- quad_tbl$n[quad_tbl$quadrant == "top-left"]
  bt <- binom.test(x = n_ul, n = n_total, p = 0.25, alternative = "less")

  quad_tbl <- quad_tbl %>%
    rowwise() %>%
    mutate(
      prop = n / n_total,
      ci_low = binom.test(n, n_total)$conf.int[1],
      ci_high = binom.test(n, n_total)$conf.int[2]
    ) %>%
    ungroup() %>%
    mutate(
      quadrant = factor(quadrant, levels = all_quadrants),
      is_top_left = quadrant == "top-left"
    )

  p_quad <- ggplot(quad_tbl, aes(x = quadrant, y = prop, fill = is_top_left)) +
    geom_col(width = 0.7) +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15) +
    geom_hline(yintercept = 0.25, linetype = "dashed", color = "#00FFFF", linewidth = 0.7) +
    geom_text(aes(label = paste0(sprintf("%.1f", 100 * prop), "%")),
              vjust = -0.4,
              size = 3.6) +
    scale_fill_manual(values = c("TRUE" = "#F94144", "FALSE" = "grey75"), guide = "none") +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    labs(
      title = gsub("_", "-", r),
      subtitle = paste0(
        "Upper-left depletion vs 25% null: p = ",
        format.pval(bt$p.value, digits = 3, eps = 1e-4),
        " (one-sided binomial)"
      ),
      x = "Quadrant",
      y = "Proportion of points"
    ) +
    ggpubr::theme_classic2()

  p_quad
})

ggarrange(plotlist = quadrant_stats_plots)
```

