---
title: Consistency metrics on supervised cell type classification performance
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show

---

# Introduction

This script analyzes the results of the label transfer consistency benchmarking pipeline (label_transfer_task pipeline) in a supervised cell type classification performance.

Building on the benchmarking framework established by Abdelaal et al., 2019 and implemented as part of the Open problems in single-cell analysis (Luecken et al., 2025), we benchmarked 13 classifiers across a diverse collection of datasets (data_processing/config/specs_datasets.csv). We implemented two complementary evaluation schemes: (i) intra-dataset, using sample-aware train-test splits to assess generalization across biological replicates, and (ii) inter-dataset, training on one dataset and testing on an independent dataset to emulate cross-study label transfer

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

```{r libraries}
require(ggplot2)
require(dplyr)
require(tidyr)
require(ggpubr)

# Custom theme
theme_set(theme_bw(base_size = 12))
```

```{r exporting_dir}
fig_dir <- "exported_figs"
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

Load data from results (within dataset splitting) and results_between (prediction between different datasets but same annotation) obtained previously in label_transfer_task pipeline.

```{r load_data}
paths <- 
  c(
    "intra_dataset" = file.path("..", "label_transfer_task", "results", "aggregated", "label_transfer_consistency_detailed.csv"),
    "inter_dataset" = file.path("..", "label_transfer_task", "results_between", "aggregated", "label_transfer_between_consistency_detailed.csv")
  )
res <- lapply(paths, read.csv)
# Preview
lapply(res, head)
```

which ISC metric to use

```{r}
isc <- c("X2label_silhouette...Pseudobulk.Cosine",
         "silhouette...recip_classif.Match",
         "product" )[3]
```


# Generate ranking-based color palette
```{r create_color_palette}
# Calculate rankings to determine color palette
# Step 1: Calculate mean product per dataset and classifier (across cell types)
# Step 2: Calculate geometric mean across datasets
rankings_palette <- lapply(res, function(cons_combined){
  rankings_for_palette <- cons_combined %>%
    filter(!is.na(f1) & !is.na(.data[[isc]]),
           classifier != "ground_truth") %>%
    group_by(dataset_id, classifier) %>%
    summarise(
      mean_product_per_dataset = mean(.data[[isc]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    group_by(classifier) %>%
    summarise(
      geom_mean_product = exp(mean(log(mean_product_per_dataset), na.rm = TRUE)),
      mean_product = mean(mean_product_per_dataset, na.rm = T),
      .groups = "drop"
    ) %>%
    mutate(rank = rank(-geom_mean_product, ties.method = "average")) %>%
    arrange(rank)
  
  print(rankings_for_palette)
  
  # Create a unified, elegant, color-blind friendly palette
  # Integrating highly distinguishable top 3 best and top 3 worst colors
  n_classifiers <- nrow(rankings_for_palette)
  
  # Define the 6 distinct colors for top 3 best and top 3 worst
  colors_top3_best <- c("#0173B2", "#029E73", "#B19CD9")  # Deep blue, forest green, malva/violet
  colors_top3_worst <- c("#DE8F05", "#FFD700", "#882255")  # Orange, gold, purple-red
  
  if (n_classifiers <= 6) {
    # If 6 or fewer classifiers, use the 6 distinct colors
    gradient_colors <- c(colors_top3_best, colors_top3_worst)[1:n_classifiers]
  } else {
    # For more than 6 classifiers, create a palette that:
    # - Uses top 3 best colors for the best performers
    # - Interpolates intermediate colors
    # - Uses top 3 worst colors for the worst performers
    
    # Generate intermediate colors using color interpolation
    # From best (blue) to worst (red)
    intermediate_palette <- colorRampPalette(c("#0173B2", "#E0E0E0", "#882255"))(n_classifiers - 6)
    
    # Combine: top 3 best + intermediate + top 3 worst
    gradient_colors <- c(
      colors_top3_best,
      intermediate_palette[1:(n_classifiers - 6)],
      colors_top3_worst
    )
  }
  
  # Create named palette
  classifier_colors <- setNames(gradient_colors, rankings_for_palette$classifier)
  
  # Identify top 6 classifiers
  top_3_best <- rankings_for_palette$classifier[1:3]
  top_3_worst <- tail(rankings_for_palette$classifier, 3)
  top_6_classifiers <- c(top_3_best, top_3_worst)
  
  # For f1_vs_product_mean plot: highlight top 6, gray out the rest
  classifier_colors_highlight <- classifier_colors
  classifier_colors_highlight[!names(classifier_colors_highlight) %in% top_6_classifiers] <- "#808080"
  
  return(classifier_colors)
  
})
  
  
```



# Per dataset/classifier
Show mean F1 vs mean product 
Add in the plot R2 and p-val
```{r f1_vs_product_mean}

# Calculate mean F1 and product per dataset and classifier
scatter_mean_cor <- lapply(names(res), function(r){
  top_6 <- c(head(names(rankings_palette[[r]]), 3),
             tail(names(rankings_palette[[r]]), 3))
  mean_metrics <- res[[r]] %>%
    filter(!is.na(f1) & !is.na(.data[[isc]])) %>%
    filter(classifier != "ground_truth") %>%
    mutate(classifier = factor(classifier,
                               levels = c(names(rankings_palette[[r]])))) %>% 
    group_by(dataset_id, classifier) %>%
    summarise(
      mean_f1 = mean(f1, na.rm = TRUE),
      mean_product = mean(.data[[isc]], na.rm = TRUE),
      .groups = "drop"
    )
  
  # Add highlight status to data (for filtering in legend)
  mean_metrics <- mean_metrics %>%
    mutate(is_highlighted = classifier %in% top_6)
  
  # Calculate correlation
  cor_test_mean <- cor.test(mean_metrics$mean_f1, mean_metrics$mean_product)
  r2_mean <- cor_test_mean$estimate^2
  pval_mean <- cor_test_mean$p.value
  
  # Create plot
  p2 <- ggplot(mean_metrics, aes(y = mean_f1, x = mean_product)) +
    # First layer: gray dots for non-highlighted classifiers (drawn first, so in back)
    geom_point(data = mean_metrics %>% filter(!is_highlighted), 
               aes(color = classifier), alpha = 0.4, size = 1.5) +
    # Second layer: colored dots for highlighted classifiers (drawn last, so in front)
    geom_point(data = mean_metrics %>% filter(is_highlighted), 
               aes(color = classifier), alpha = 0.7, size = 1.5) +
    scale_x_continuous(limits = c(-0.1, 1)) +
    scale_y_continuous(limits = c(-0.1, 1)) +
    geom_smooth(method = "lm", color = "black", linetype = "dashed", se = TRUE) +
    scale_color_manual(values = rankings_palette[[r]]) +
    labs(
      title = r,
      subtitle = sprintf("R² = %.3f, p-value = %.2e", r2_mean, pval_mean),
      y = "F1 Score Test",
      x = paste0("ISC Test (", isc, ")"),
      color = "Classifier"
    ) +
    ggpubr::theme_classic2() +
    theme(
      legend.position = "right",
      legend.title = element_blank()
    )
  
  # Custom legend: only show highlighted classifiers
  # Get unique highlighted classifiers from the data
  highlighted_classifiers <- unique(mean_metrics$classifier[mean_metrics$is_highlighted])
  p2 <- p2 + 
    scale_color_manual(
      values = rankings_palette[[r]][names(rankings_palette[[r]]) %in% top_6],
      breaks = names(rankings_palette[[r]][names(rankings_palette[[r]]) %in% top_6])
    )
  
  return(p2)
  
})


ggarrange(plotlist = scatter_mean_cor)
```

## Export plots

```{r}
pdf_file <- file.path(fig_dir, paste0("LabelProjection_perClassifier_dataset_", "mean_scatter.pdf"))

pdf(pdf_file, width = 5, height = 4)
for(a in scatter_mean_cor) print(a)
dev.off()
```

# Per classifier
Show mean F1 vs mean product 
Add in the plot R2 and p-val
```{r f1_vs_product_perClassifier}

# Create custom axis transformation: 25% of space for [-0.1, 0.7], 75% for [0.7, 1]
compressed_axis_trans <- scales::trans_new(
  name = "compressed",
  transform = function(x) {
    ifelse(x <= 0.7,
           (x + 0.1) / 3.2,           # [-0.1, 0.7] → [0, 0.25]
           0.25 + (x - 0.7) * 2.5)    # [0.7, 1] → [0.25, 1]
  },
  inverse = function(x) {
    ifelse(x <= 0.25,
           x * 3.2 - 0.1,             # [0, 0.25] → [-0.1, 0.7]
           (x - 0.25) / 2.5 + 0.7)    # [0.25, 1] → [0.7, 1]
  }
)

# Calculate mean F1 and product per dataset and classifier
scatter_mean_cor <- lapply(names(res), function(r){
  mean_metrics <- res[[r]] %>%
    filter(!is.na(f1) & !is.na(.data[[isc]])) %>%
    filter(classifier != "ground_truth") %>%
    mutate(classifier = factor(classifier,
                               levels = c(names(rankings_palette[[1]])))) %>% 
    group_by(classifier) %>%
    summarise(
      mean_f1 = mean(f1, na.rm = TRUE),
      mean_product = mean(.data[[isc]], na.rm = TRUE),
      .groups = "drop"
    )
  
  # Calculate correlation
  cor_test_mean <- cor.test(mean_metrics$mean_f1, mean_metrics$mean_product)
  r2_mean <- cor_test_mean$estimate^2
  pval_mean <- cor_test_mean$p.value
  
  # Create plot
  p2 <- ggplot(mean_metrics,
               aes(y = mean_f1, x = mean_product)) +
    geom_smooth(method = "lm",
                color = "black",
                linetype = "dashed",
                se = TRUE,
                alpha = 0.2) +
    geom_point(aes(color = classifier),
               size = 2,
               show.legend = F) +
    ggrepel::geom_label_repel(aes(label = classifier,
                                  color = classifier),
                              size = 2,
                              show.legend = F,
                              max.overlaps = 20) +
    scale_color_manual(values = rankings_palette[[1]]) +
    scale_x_continuous(limits = c(-0.1, 0.9), trans = compressed_axis_trans) +
    scale_y_continuous(limits = c(-0.1, 0.9), trans = compressed_axis_trans) +
    
    scale_color_manual(values = rankings_palette[[1]]) +
    labs(
      title = gsub("_", "-", r),
      subtitle = sprintf("R² = %.3f, p-value = %.2e", r2_mean, pval_mean),
      y = "F1 Score Test",
      x = paste0("ISC Test (", isc, ")"),
      color = "Classifier"
    ) +
    ggpubr::theme_classic2()
  
  return(p2)
  
})


ggarrange(plotlist = scatter_mean_cor)
```

```{r}
# Save as PDF
pdf_file <- file.path(fig_dir, paste0("LabelProjection_perClassifier_", "mean_scatter.pdf"))

pdf(pdf_file, width = 4, height = 4)
for(a in scatter_mean_cor) print(a)
dev.off()
```


# Plot multiple ISC
Plot only correlation lines for 1) product (as earlier), 2) global ISC, 3) lobal ISC

```{r isc_names}
iscs <- c("X2label_silhouette...Pseudobulk.Cosine",
         "silhouette...recip_classif.Match",
         "product" )
isc_colors <- c(
  "ISC Global" = "#0173B2",   # blue
  "ISC Local" = "#DE8F05",    # orange
  "ISC Product" = "#029E73"   # green
)
```

```{r corr_different_isc}
# Calculate mean F1 and product per dataset and classifier
smooth_cor <- lapply(names(res), function(r){
  mean_metrics <- res[[r]] %>%
    filter(!is.na(f1) & !is.na(.data[[isc]])) %>%
    filter(classifier != "ground_truth") %>%
    mutate(classifier = factor(classifier,
                               levels = c(names(rankings_palette[[r]])))) %>% 
    group_by(dataset_id, classifier) %>%
    summarise(
      mean_f1 = mean(f1, na.rm = TRUE),
      across(all_of(iscs), ~ mean(.x, na.rm = TRUE), .names = "{.col}"),
      .groups = "drop"
    ) %>% 
    pivot_longer(-c(dataset_id, classifier, mean_f1),
                 names_to = "metric",
                 values_to = "score") %>% 
    mutate(metric = factor(metric,
                           levels = iscs,
                           labels = c("ISC Global",
                                      "ISC Local",
                                      "ISC Product"))
    )
  
  corrs <- mean_metrics %>% 
    group_by(metric) %>% 
    summarize(
      r2_mean = signif(cor(mean_f1, score, use = "complete.obs")^2, 2),
      pval_mean = formatC(cor.test(mean_f1, score)$p.value, format = "e", digits = 1),
      .groups = "drop"
    )
  
    # Custom legend labels
  legend_labels <- setNames(
    sprintf("%s\nR²=%.2f\np=%.1e", corrs$metric, corrs$r2_mean, as.numeric(corrs$pval_mean)),
    corrs$metric
  )
  # Create plot
  p2 <- ggplot(mean_metrics, aes(y = mean_f1, x = score,
                                 color = metric,
                                 fill = metric)) +
    geom_smooth(
      method = "lm",
      se = TRUE,
      linetype = "solid",
      alpha = 0.4,
      aes(fill = metric),
      show.legend = T
    ) +
    scale_color_manual(values = isc_colors, labels = legend_labels, name = "ISC metric") +
    scale_fill_manual(values = isc_colors, guide = "none") +
    labs(
      title = r,
      y = "F1 Score Test",
      x = "ISC Test",
    ) +
    guides(fill = "none") +
    ggpubr::theme_classic2() +
    theme(
      legend.position = "bottom",
      legend.title = element_blank()
    )

  
  return(p2)
  
})


ggarrange(plotlist = smooth_cor)
```
```{r}
pdf_file <- file.path(fig_dir, paste0("LabelProjection_multipleISC_corr", ".pdf"))

pdf(pdf_file, width = 4, height = 5)
for(a in smooth_cor) print(a)
dev.off()
```



# Compare ranking of classifiers either using F1 or product and how they agree in the ranking

```{r ranking_comparison}
# Calculate overall rankings
# Step 1: Calculate mean metrics per dataset and classifier (across cell types)
# Step 2: Calculate geometric mean across datasets

scatter_ranking <- lapply(names(res), function(r){
rankings <- res[[r]] %>%
  filter(!is.na(f1) & !is.na(product),
         classifier != "ground_truth") %>%
  group_by(dataset_id, classifier) %>%
  summarise(
    mean_f1_per_dataset = mean(f1, na.rm = TRUE),
    mean_product_per_dataset = mean(product, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(classifier) %>%
  summarise(
    geom_mean_f1 = exp(mean(log(mean_f1_per_dataset), na.rm = TRUE)),
    geom_mean_product = exp(mean(log(mean_product_per_dataset), na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    rank_f1 = rank(-geom_mean_f1, ties.method = "average"),
    rank_product = rank(-geom_mean_product, ties.method = "average"),
    rank_difference = abs(rank_f1 - rank_product)
  ) %>%
  mutate(classifier = factor(classifier,
                             levels = c(names(rankings_palette[[1]])))) %>%
  arrange(rank_product)

print(rankings)

# Spearman correlation between rankings
rank_cor <- cor.test(rankings$rank_f1, rankings$rank_product, method = "spearman")

cat(sprintf("\nSpearman correlation between F1 and Product rankings: %.3f (p = %.2e)\n", 
            rank_cor$estimate, rank_cor$p.value))

# Visualize ranking comparison
rankings_long <- rankings %>%
  select(classifier, rank_f1, rank_product) %>%
  pivot_longer(cols = c(rank_f1, rank_product), names_to = "metric", values_to = "rank") %>%
  mutate(metric = recode(metric, 
                         rank_f1 = "F1 Rank", 
                         rank_product = "Product Rank"))

# Ranking agreement table
p6 <- ggplot(rankings, aes(x = rank_product, y = rank_f1)) +
  geom_point(aes(color = classifier),
             size = 4,) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  ggrepel::geom_label_repel(aes(label = classifier,
                                color = classifier),
                           size = 5) +
  scale_color_manual(values = rankings_palette[[1]]) +
  scale_x_reverse(breaks = 1:nrow(rankings)) +
  scale_y_reverse(breaks = 1:nrow(rankings)) +
  labs(
    title = "Ranking Agreement: ISC vs F1",
    subtitle = r,
    x = "ISC Rank",
    y = "F1 Rank",
    color = "Classifier"
  ) +
  coord_fixed() +
  ggpubr::theme_classic2() +
  theme(legend.position = "none") 

return(p6)

})

ggarrange(plotlist = scatter_ranking)
```

# ranking mean
Per-dataset ranking of classifiers based on product score, then average rank across datasets

```{r ranking_per_dataset}
# Parameter: use geometric mean (TRUE) or arithmetic mean (FALSE)
use_geom_mean <- TRUE

per_dataset_ranking <- lapply(names(res), function(r){
  # Step 1: Calculate mean metrics per dataset and classifier (across cell types)
  dataset_metrics_mean <- res[[r]] %>%
    filter(!is.na(f1) & !is.na(product),
           classifier != "ground_truth") %>%
    group_by(dataset_id, classifier) %>%
    summarise(
      mean_f1 = mean(f1, na.rm = TRUE),
      mean_product = mean(product, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Step 2: Compute ranking of classifiers per dataset for both metrics
  dataset_rankings <- dataset_metrics_mean %>%
    group_by(dataset_id) %>%
    mutate(
      rank_f1 = rank(-mean_f1, ties.method = "average"),
      rank_product = rank(-mean_product, ties.method = "average")
    ) %>%
    ungroup()
  
  # Step 3: Calculate mean rank per classifier across datasets
  # Use either arithmetic mean or geometric mean based on parameter
  mean_rank_per_classifier <- dataset_rankings %>%
    group_by(classifier) %>%
    summarise(
      mean_rank_f1 = if (use_geom_mean) exp(mean(log(rank_f1), na.rm = TRUE)) else mean(rank_f1, na.rm = TRUE),
      mean_rank_product = if (use_geom_mean) exp(mean(log(rank_product), na.rm = TRUE)) else mean(rank_product, na.rm = TRUE),
      sd_rank_f1 = sd(rank_f1, na.rm = TRUE),
      sd_rank_product = sd(rank_product, na.rm = TRUE),
      n_datasets = n(),
      .groups = "drop"
    ) %>%
    mutate(classifier = factor(classifier,
                               levels = c(names(rankings_palette[[r]])))) %>%
    arrange(mean_rank_product)
  
  mean_type <- if (use_geom_mean) "Geometric Mean" else "Arithmetic Mean"
  cat(sprintf("\nMean rank per classifier across datasets (%s):\n", mean_type))
  print(mean_rank_per_classifier)
  
  # Calculate correlation between F1 rank and product rank
  rank_cor_dataset <- cor.test(mean_rank_per_classifier$mean_rank_f1, mean_rank_per_classifier$mean_rank_product, method = "spearman")
  
  cat(sprintf("\nSpearman correlation between F1 and Product rankings (per-dataset): %.3f (p = %.2e)\n", 
              rank_cor_dataset$estimate, rank_cor_dataset$p.value))
  
  # Visualize ranking comparison
  rankings_long_dataset <- mean_rank_per_classifier %>%
    select(classifier, mean_rank_f1, mean_rank_product) %>%
    pivot_longer(cols = c(mean_rank_f1, mean_rank_product), names_to = "metric", values_to = "rank") %>%
    mutate(metric = recode(metric, 
                           mean_rank_f1 = "F1 Rank", 
                           mean_rank_product = "Product Rank"))
  
  p_dataset_rank <- ggplot(rankings_long_dataset, aes(x = metric, y = rank, group = classifier, color = classifier)) +
    geom_line(size = 1.2, alpha = 0.7) +
    geom_point(size = 3) +
    scale_y_reverse() +
    scale_color_manual(values = rankings_palette[[1]]) +
    labs(
      title = "Classifier Ranking per Dataset: F1 vs. Consistency Score",
      subtitle = sprintf("Spearman ρ = %.3f (p = %.2e) [%s]", rank_cor_dataset$estimate, rank_cor_dataset$p.value, mean_type),
      x = "",
      y = "Mean Rank (1 = Best)",
      color = "Classifier"
    ) +
    theme(legend.position = "right")
  
  print(p_dataset_rank)
  
  # Ranking agreement table per dataset
  p_dataset_rank_agreement <- ggplot(mean_rank_per_classifier, aes(x = mean_rank_product, y = mean_rank_f1)) +
    geom_point(aes(color = classifier),
               size = 4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
    ggrepel::geom_label_repel(aes(label = classifier,
                                  color = classifier),
                              size = 5) +
    scale_color_manual(values = rankings_palette[[1]]) +
    scale_x_reverse() +
    scale_y_reverse() +
    labs(
      title = r,
      subtitle = sprintf("Ranking Agreement per Dataset: ISC vs F1 [%s]", mean_type),
      x = "Mean ISC Rank",
      y = "Mean F1 Rank",
      color = "Classifier"
    ) +
    coord_fixed() +
    ggpubr::theme_classic2() +
    theme(legend.position = "none") 
  
  return(p_dataset_rank_agreement)
})

ggarrange(plotlist = per_dataset_ranking)
```


## Export plots
```{r}
# Save as PDF
pdf_file <- file.path(fig_dir, paste0("LabelProjection_", "ranking.pdf"))
ggsave(
  pdf_file,
  plot = p6,
  width = 5,
  height = 5,
  device = "pdf"
)

```





# Reference consistency


Make a geom_point plot showing the mean of F1 in predictions for each consistency in reference, adding error bar plots to show the variability of the different predictions.

## correlation

```{r reference_consistency}
excl <- c("ground_truth", "Ensemble", "Random")
# filter for only references
cons_ref <- lapply(names(res), function(r){
  # Summarise prediction F1 per dataset, classifier, and cell type
  class_levs <- names(rankings_palette[[1]])[!names(rankings_palette[[1]]) %in% excl]
  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier,
                               levels = c(class_levs))) %>% 
    filter(classifier != "ground_truth",
           !is.na(f1),
           ) %>% 
    select(cell_type, dataset_id, f1, classifier)
  
  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth",
           !is.na(product)) %>% 
    select(cell_type, dataset_id, product)
  
  # Join reference consistency with prediction F1
  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(product)) %>% 
    group_by(dataset_id, cell_type)
  
  # Correlation per classifier (mean F1 vs reference product)
  cor_table <- joined %>%
    group_by(classifier) %>%
    summarise(
      r = cor(f1, product, use = "complete.obs"),
      r2 = r^2,
      p_value = cor.test(f1, product)$p.value,
      n_pairs = sum(!is.na(f1) & !is.na(product)),
      .groups = "drop"
    ) %>%
    ungroup() %>% 
    # Apply FDR correction for multiple testing
    mutate(
      p_value_adjusted = p.adjust(p_value, method = "BH"),
      label = sprintf("R² = %.3f\np = %.2e", r2, p_value)
    )
  
  cat("\nCorrelation (F1 vs reference consistency product) per classifier:\n")
  print(cor_table)
  
  # Plot: F1 vs reference consistency product with facets and correlation stats
  p_ref <- joined %>% 
    ggplot(aes(x = product, y = f1)) +
    geom_point(aes(color = classifier), alpha = 0.7, size = 2, show.legend = FALSE) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    scale_color_manual(values = rankings_palette[[1]]) +
    geom_label(data = cor_table, 
               aes(x = Inf, y = -Inf, label = label),
               fill = "white",
               alpha = 0.4,
               hjust = 1.1, vjust = -0.5, size = 3.5, 
               inherit.aes = FALSE) +
    labs(
      title = r,
      subtitle = "Train ISC vs Prediction F1, per cell type",
      x = "ISC Train",
      y = "F1 (predictions)"
    ) +
    ggpubr::theme_classic2() +
    theme(legend.position = "right") +
    facet_wrap(~classifier)
  
  return(p_ref)
})

ggarrange(plotlist = cons_ref)
```

```{r}
pdf_file <- file.path(fig_dir, paste0("LabelProjection_referenceISC_scatter", ".pdf"))

pdf(pdf_file, width = 8*1.1, height = 6*1.1)
for(a in cons_ref) print(a)
dev.off()
```

## categorize ISC

```{r}
excl <- c("ground_truth", "Ensemble", "Random")
ths <- 0.7
# filter for only references
cons_ref_cat <- lapply(names(res), function(r){
  # Summarise prediction F1 per dataset, classifier, and cell type
  class_levs <- names(rankings_palette[[r]])[!names(rankings_palette[[r]]) %in% excl]
  top_6 <- c(head(class_levs, 3), tail(class_levs,3))
  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier,
                               levels = c(class_levs))) %>% 
    filter(classifier != "ground_truth",
           !is.na(f1),
           classifier %in% top_6) %>% 
    select(cell_type, dataset_id, f1, classifier)
  
  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth",
           !is.na(product)) %>% 
    select(cell_type, dataset_id, product) %>% 
    mutate(consistency = ifelse(product > ths,
                              "consistent",
                              "inconsistent"))
  
  # Join reference consistency with prediction F1
  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(product)) %>% 
    group_by(dataset_id, cell_type)
  
  
  
  # Plot: F1 vs reference consistency product with facets and correlation stats
  p_ref <- joined %>% 
    ggplot(aes(x = consistency, y = f1)) +
    geom_jitter(alpha = 0.3, color = "grey", width = 0.15) +
    geom_boxplot(aes(color = classifier),
                 fill = NA,
                 outlier.colour = NA,
                 show.legend = FALSE) +
    scale_color_manual(values = rankings_palette[[1]]) +
    labs(
      title = r,
      subtitle = paste0("F1 in consistent (ISC>", ths,
                        "), and incosistent cell types in train set"),
      x = "ISC category",
      y = "F1 (predictions)"
    ) +
    ggpubr::theme_classic2() +
    facet_wrap(~classifier)
  
  return(p_ref)
})

ggarrange(plotlist = cons_ref_cat)
```


## Export plots
```{r}
# Save as PDF
pdf_file <- file.path(fig_dir, paste0("LabelProjection_", "TrainISC.pdf"))
ggsave(
  pdf_file,
  plot = p_ref,
  width = 7,
  height = 5,
  device = "pdf"
)

```

## contigency plot

```{r density_contigency}
ths_f1 <- 0.7
ths_isc <- 0.5
xcl <- c("ground_truth", "Ensemble", "Random")
# filter for only references
cons_ref_contingency <- lapply(names(res), function(r){
  # Summarise prediction F1 per dataset, classifier, and cell type
  class_levs <- names(rankings_palette[[1]])[!names(rankings_palette[[1]]) %in% excl]
  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier,
                               levels = c(class_levs))) %>% 
    filter(classifier != "ground_truth",
           !is.na(f1),
           ) %>% 
    select(cell_type, dataset_id, f1, classifier)
  
  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth",
           !is.na(product)) %>% 
    select(cell_type, dataset_id, product)
  
  # Join reference consistency with prediction F1
  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(product)) %>% 
    group_by(dataset_id, cell_type)


  
  # Compute quadrant counts (threshold = 0.5 on both axes)
  quadrant_counts <- joined %>%
    ungroup() %>%
    mutate(quadrant = case_when(
      product >= ths_isc & f1 >= ths_f1 ~ "top-right",
      product <  ths_isc & f1 >= ths_f1 ~ "top-left",
      product >= ths_isc & f1 <  ths_f1 ~ "bottom-right",
      TRUE                        ~ "bottom-left"
    )) %>%
    count(quadrant) %>%
    mutate(
      x = case_when(quadrant == "top-right"    ~ 0.75,
                    quadrant == "top-left"      ~ 0.25,
                    quadrant == "bottom-right"  ~ 0.75,
                    quadrant == "bottom-left"   ~ 0.25),
      y = case_when(quadrant == "top-right"    ~ 0.85,
                    quadrant == "top-left"      ~ 0.85,
                    quadrant == "bottom-right"  ~ 0.25,
                    quadrant == "bottom-left"   ~ 0.25),
      #label = n
      label = paste0(sprintf("%.1f", 100 * n / sum(n)), "%")
    )

  # Plot: bin2d density + quadrant percentages
  p_ref <- joined %>%
    ggplot(aes(x = product, y = f1)) +
    geom_bin2d(bins = 50, show.legend = T) +
    scale_fill_viridis_c(option = "magma", trans = "sqrt") +
    geom_vline(xintercept = ths_isc, linetype = "dashed", color = "#00FFFF", linewidth = 0.8) +
    geom_hline(yintercept = ths_f1, linetype = "dashed", color = "#00FFFF", linewidth = 0.8) +
    geom_label(data = quadrant_counts,
              aes(x = x, y = y, label = label),
              inherit.aes = FALSE,
              size = 6,
              fill = "grey90",
              color = "black",
              alpha = 0.9) +
    scale_x_continuous(limits = c(0, 1)) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title = gsub("_", "-", r),
      subtitle = "Train ISC vs Prediction F1, per cell type",
      x = "ISC Train",
      y = "F1 score Test"
    ) +
    ggpubr::theme_classic2() +
    theme(legend.position = "right")
  
  return(p_ref)
})

ggarrange(plotlist = cons_ref_contingency)
```

```{r}
pdf_file <- file.path(fig_dir, paste0("LabelProjection_referenceISC_contigency", ".pdf"))

pdf(pdf_file, width = 4.5, height = 4)
for(a in cons_ref_contingency) print(a)
dev.off()
```


```{r upper_left_quadrant_depletion}
# Test whether upper-left quadrant is depleted vs 25% null expectation
quadrant_stats_plots <- lapply(names(res), function(r){
  class_levs <- names(rankings_palette[[1]])[!names(rankings_palette[[1]]) %in% excl]

  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier, levels = c(class_levs))) %>%
    filter(classifier != "ground_truth", !is.na(f1)) %>%
    select(cell_type, dataset_id, f1, classifier)

  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth", !is.na(product)) %>%
    select(cell_type, dataset_id, product)

  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(product))

  quad_tbl <- joined %>%
    mutate(quadrant = case_when(
      product >= ths_isc & f1 >= ths_f1 ~ "top-right",
      product <  ths_isc & f1 >= ths_f1 ~ "top-left",
      product >= ths_isc & f1 <  ths_f1 ~ "bottom-right",
      TRUE                              ~ "bottom-left"
    )) %>%
    count(quadrant, .drop = FALSE)

  all_quadrants <- c("top-left", "top-right", "bottom-left", "bottom-right")
  quad_tbl <- dplyr::full_join(
    tibble::tibble(quadrant = all_quadrants),
    quad_tbl,
    by = "quadrant"
  ) %>%
    mutate(n = ifelse(is.na(n), 0L, n))

  n_total <- sum(quad_tbl$n)
  n_ul <- quad_tbl$n[quad_tbl$quadrant == "top-left"]
  bt <- binom.test(x = n_ul, n = n_total, p = 0.25, alternative = "less")

  quad_tbl <- quad_tbl %>%
    rowwise() %>%
    mutate(
      prop = n / n_total,
      ci_low = binom.test(n, n_total)$conf.int[1],
      ci_high = binom.test(n, n_total)$conf.int[2]
    ) %>%
    ungroup() %>%
    mutate(
      quadrant = factor(quadrant, levels = all_quadrants),
      is_top_left = quadrant == "top-left"
    )

  p_quad <- ggplot(quad_tbl, aes(x = quadrant, y = prop, fill = is_top_left)) +
    geom_col(width = 0.7) +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15) +
    geom_hline(yintercept = 0.25, linetype = "dashed", color = "#00FFFF", linewidth = 0.7) +
    geom_text(aes(label = paste0(sprintf("%.1f", 100 * prop), "%")),
              vjust = -0.4,
              size = 3.6) +
    scale_fill_manual(values = c("TRUE" = "#F94144", "FALSE" = "grey75"), guide = "none") +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    labs(
      title = gsub("_", "-", r),
      subtitle = paste0(
        "Upper-left depletion vs 25% null: p = ",
        format.pval(bt$p.value, digits = 3, eps = 1e-4),
        " (one-sided binomial)"
      ),
      x = "Quadrant",
      y = "Proportion of points"
    ) +
    ggpubr::theme_classic2()

  p_quad
})

ggarrange(plotlist = quadrant_stats_plots)
```

