---
title: "Figure 4: Consistency metrics on supervised cell type classification performance"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# Introduction

This script analyzes the results of the label transfer consistency benchmarking pipeline (`label_transfer_task` pipeline) in a supervised cell type classification performance. 
Exported plots are included in the manuscript as Main Figure 4 and Extended Data Figure 3. 

Building on the benchmarking framework established by Abdelaal et al., 2019 and implemented as part of the Open problems in single-cell analysis (Luecken et al., 2025), we benchmarked 13 classifiers across a diverse collection of datasets (`data_processing/config/specs_datasets.csv`). We implemented two complementary evaluation schemes: (i) intra-dataset, using sample-aware train-test splits to assess generalization across biological replicates, and (ii) inter-dataset, training on one dataset and testing on an independent dataset to emulate cross-study label transfer

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

```{r libraries}
require(ggplot2)
require(dplyr)
require(tidyr)
require(ggpubr)

# Custom theme
theme_set(theme_bw(base_size = 12))
```

```{r exporting_dir}
fig_dir <- file.path(
  "~/OneDrive - Université de Genève/CSI/LabPapers/Intersample_Consistency/Figures/plots/Figure4/"
)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

Load data from results (within dataset splitting) and results_between (prediction between different datasets but same annotation) obtained previously in label_transfer_task pipeline.

Add function to collapse replicates
```{r}
iscs <- c("X2label_silhouette...Pseudobulk.Cosine",
          "silhouette...recip_classif.Match",
          "product",
          "SCCAF",
          "MetaNeighbor_Supervised...NA",
          "nsa_cLISI...NA",
          "hotelling",
          "cramer")

labs <- c("ISC Global",
          "ISC Local",
          "ISC Product",
          "SCCAF",
          "MetaNeighbor Supervised",
          "cLISI",
          "scDiagnostics-Hotelling",
          "scDiagnostics-Cramer") 

collapse_replicates <- function(x) {
  x %>%
    group_by(dataset_id, classifier, cell_type) %>%
    summarise(
      across(
        c(f1, accuracy, all_of(iscs)),
        ~ mean(.x, na.rm = TRUE)
      ),
      n_replicates = n_distinct(replicate),
      .groups = "drop"
    )
}
```


```{r load_data}
paths <- 
  c(
    "intra_dataset" = file.path("..", "label_transfer_task", "results", "aggregated", "label_transfer_consistency_detailed.csv"),
    "inter_dataset" = file.path("..", "label_transfer_task", "results_between", "aggregated", "label_transfer_between_consistency_detailed.csv")
  )
res0 <- lapply(paths, read.csv)

res <- lapply(res0, function(x){
  x$product <- x[["X2label_silhouette...Pseudobulk.Cosine"]] * x[["silhouette...recip_classif.Match"]]
  x <- collapse_replicates(x)
  return(x)
})
# Preview
lapply(res, head)
```

# Helper functions

Modular helpers reused by the sections below. Each takes the ISC metric column (`isc_col`) as a parameter so analyses can be looped over all `iscs`. `excl` controls which classifiers (e.g. "Random", "Ensemble") are dropped before computing correlations/rankings.

```{r isc_helpers}
# Set to FALSE to keep "Random" and "Ensemble" classifiers in correlation/ranking analyses
source("helpers/supervised_classification.R")
exclude_random_ensemble <- TRUE
excl <- if (exclude_random_ensemble) c("ground_truth", "Random", "Ensemble") else "ground_truth"
```

# Per classifier
Show mean F1 vs mean ISC, looped over every metric in `iscs`.
```{r f1_vs_product_perClassifier}
perClassifier_results <- lapply(iscs, function(isc_col){
  isc_label <- labs[match(isc_col, iscs)]
  lapply(names(res), function(r){
    plot_isc_vs_f1_per_classifier(
      df = res[[r]],
      isc_col = isc_col,
      palette = "f1",
      excl = excl,
      dataset_label = r
    )
  }) %>% setNames(names(res))
}) %>% setNames(iscs)

# Preview for the default metric (product)
lapply(perClassifier_results, function(e){
  lapply(e, function(i){
    print(i[["plot"]])
  })
})
```


# Plot multiple ISC
Overlay correlation lines for every ISC metric in `iscs` on a single plot per dataset.

```{r isc_names}
isc_colors <- c(
  "ISC Global" = "#0173B2",             # blue
  "ISC Local" = "#D55E00",              # orange
  "ISC Product" = "#029E73",            # green
  "SCCAF" = "#CC78BC",                  # purple
  "MetaNeighbor Supervised" = "brown4", # brown
  "cLISI" = "#949494",                  # grey
  "scDiagnostics-Hotelling" = "#56B4E9",# sky blue
  "scDiagnostics-Cramer" = "#CA9161"    # tan
)
```

```{r corr_different_isc, fig.height=10, fig.width=5}
multi_isc_results <- lapply(names(res), function(r){
  plot_multi_isc_corr(res[[r]], iscs = iscs, labs = labs, isc_colors = isc_colors,
                       excl = excl, dataset_label = r)
}) %>% setNames(names(res))

multi_isc_results
```


# Compare ranking of classifiers either using F1 or an ISC metric, looped over `iscs`

```{r ranking_comparison}
ranking_comparison_results <- lapply(iscs, function(isc_col){
  lapply(names(res), function(r){
    plot_ranking_comparison(res[[r]], isc_col = isc_col,
                             excl = excl, dataset_label = r)
  }) %>% setNames(names(res))
}) %>% setNames(iscs)

ranking_comparison_results
```

# ranking mean
Per-dataset ranking of classifiers based on each ISC metric, then average rank across datasets, looped over `iscs`.

```{r ranking_per_dataset}
# Parameter: use geometric mean (TRUE) or arithmetic mean (FALSE)
use_geom_mean <- TRUE

ranking_per_dataset_results <- lapply(iscs, function(isc_col){
  lapply(names(res), function(r){
    plot_ranking_per_dataset(res[[r]], isc_col = isc_col,
                              excl = excl, use_geom_mean = use_geom_mean,
                             dataset_label = r)
  }) %>% setNames(names(res))
}) %>% setNames(iscs)

ranking_per_dataset_results
```


# ISC vs F1 scatter grouped by cell type / classifier, per ISC metric and dataset

For every ISC metric and dataset (intra/inter), plot mean F1 vs mean ISC grouped (colored) either by `cell_type` or by `classifier`, and collect the R² / p-value of each fit into a single summary table.

```{r isc_vs_f1_grouped_all}
group_vars <- list(celltype = c("classifier", "cell_type"),
                  classifier = "classifier")

grouped_results <- lapply(group_vars, function(gv){
  lapply(iscs, function(isc_col){
    isc_label <- labs[match(isc_col, iscs)]
    lapply(names(res), function(r){
      plot_isc_vs_f1_grouped(
        df = res[[r]],
        isc_col = isc_col,
        isc_label = isc_label,
        group_by = gv,
        excl = excl,
        dataset_label = paste0(r, " - ", paste(gv, collapse = ":")),
        palette = "f1"
      )
    }) %>% setNames(names(res))
  }) %>% setNames(iscs)
}) %>% setNames(names(group_vars))

lapply(grouped_results, function(a){
  lapply(a, function(b){
    lapply(b, function(i){
      print(i[["plot"]])
    })
  })
})

```


```{r}

# Combine R2 / p-value stats for every (group_by x isc x dataset) combination
isc_vs_f1_stats_table <- do.call(rbind, lapply(names(group_vars), function(gv){
  do.call(rbind, lapply(iscs, function(isc_col){
    do.call(rbind, lapply(names(res), function(r){
      grouped_results[[gv]][[isc_col]][[r]]$stats
    }))
  }))
})) %>%
  mutate(isc_label = factor(isc_label, levels = labs)) %>%
  arrange(desc(r2), dataset)
rownames(isc_vs_f1_stats_table) <- NULL

isc_vs_f1_stats_table
```


## plot r2 per dataset

```{r}

```




# Reference consistency


Make a geom_point plot showing the mean of F1 in predictions for each consistency in reference, adding error bar plots to show the variability of the different predictions.

## overall correlation
```{r reference_consistency_overall}
excl <- c("ground_truth", "Ensemble", "Random")
# filter for only references
cons_ref_all <- lapply(names(res), function(r){
  # Summarise prediction F1 per dataset, classifier, and cell type
  class_levs <- names(rankings_palette[[1]])[!names(rankings_palette[[1]]) %in% excl]
  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier,
                               levels = c(class_levs))) %>% 
    filter(classifier != "ground_truth",
           !is.na(f1),
           ) %>% 
    select(cell_type, dataset_id, f1, classifier)
  
  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth",
           !is.na(.data[[isc]])) %>% 
    select(cell_type, dataset_id, .data[[isc]])
  
  # Join reference consistency with prediction F1
  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(.data[[isc]])) %>% 
    group_by(dataset_id, cell_type)
  
  # Correlation per classifier (mean F1 vs reference product)
  cor_table <- joined %>%
    summarise(
      r = cor(f1, .data[[isc]], use = "complete.obs"),
      r2 = r^2,
      p_value = cor.test(f1, .data[[isc]])$p.value,
      n_pairs = sum(!is.na(f1) & !is.na(.data[[isc]])),
      .groups = "drop"
    ) %>%
    ungroup() %>% 
    # Apply FDR correction for multiple testing
    mutate(
      p_value_adjusted = p.adjust(p_value, method = "BH"),
      label = sprintf("R² = %.3f\np = %.2e", r2, p_value)
    )
  
  cat("\nCorrelation (F1 vs reference consistency product) per classifier:\n")
  print(cor_table)
  
  # Plot: F1 vs reference consistency product with facets and correlation stats
  p_ref <- joined %>% 
    ggplot(aes(x = .data[[isc]], y = f1)) +
    geom_point(aes(color = classifier), alpha = 0.7, size = 2, show.legend = FALSE) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    scale_color_manual(values = rankings_palette[[1]]) +
    geom_label(data = cor_table, 
               aes(x = Inf, y = -Inf, label = label),
               fill = "white",
               alpha = 0.4,
               hjust = 1.1, vjust = -0.5, size = 3.5, 
               inherit.aes = FALSE) +
    labs(
      title = r,
      subtitle = "Train ISC vs Prediction F1, per cell type",
      x = paste0("Value Train: ", isc),
      y = "F1 (predictions)"
    ) +
    ggpubr::theme_classic2() +
    theme(legend.position = "right")
  
  return(p_ref)
})

ggarrange(plotlist = cons_ref_all)
```

## correlation per classifier

```{r reference_consistency_perclassifier}
excl <- c("ground_truth", "Ensemble", "Random")
# filter for only references
cons_ref <- lapply(names(res), function(r){
  # Summarise prediction F1 per dataset, classifier, and cell type
  class_levs <- names(rankings_palette[[1]])[!names(rankings_palette[[1]]) %in% excl]
  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier,
                               levels = c(class_levs))) %>% 
    filter(classifier != "ground_truth",
           !is.na(f1),
           ) %>% 
    select(cell_type, dataset_id, f1, classifier)
  
  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth",
           !is.na(.data[[isc]])) %>% 
    select(cell_type, dataset_id, .data[[isc]])
  
  # Join reference consistency with prediction F1
  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(.data[[isc]])) %>% 
    group_by(dataset_id, cell_type)
  
  # Correlation per classifier (mean F1 vs reference product)
  cor_table <- joined %>%
    group_by(classifier) %>%
    summarise(
      r = cor(f1, .data[[isc]], use = "complete.obs"),
      r2 = r^2,
      p_value = cor.test(f1, .data[[isc]])$p.value,
      n_pairs = sum(!is.na(f1) & !is.na(.data[[isc]])),
      .groups = "drop"
    ) %>%
    ungroup() %>% 
    # Apply FDR correction for multiple testing
    mutate(
      p_value_adjusted = p.adjust(p_value, method = "BH"),
      label = sprintf("R² = %.3f\np = %.2e", r2, p_value)
    )
  
  cat("\nCorrelation (F1 vs reference consistency product) per classifier:\n")
  print(cor_table)
  
  # Plot: F1 vs reference consistency product with facets and correlation stats
  p_ref <- joined %>% 
    ggplot(aes(x = .data[[isc]], y = f1)) +
    geom_point(aes(color = classifier), alpha = 0.7, size = 2, show.legend = FALSE) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    scale_color_manual(values = rankings_palette[[1]]) +
    geom_label(data = cor_table, 
               aes(x = Inf, y = -Inf, label = label),
               fill = "white",
               alpha = 0.4,
               hjust = 1.1, vjust = -0.5, size = 3.5, 
               inherit.aes = FALSE) +
    labs(
      title = r,
      subtitle = "Train ISC vs Prediction F1, per cell type",
      x = paste0("Value Train: ", isc),
      y = "F1 (predictions)"
    ) +
    ggpubr::theme_classic2() +
    theme(legend.position = "right") +
    facet_wrap(~classifier)
  
  return(p_ref)
})

ggarrange(plotlist = cons_ref)
```

```{r}
pdf_file <- file.path(fig_dir, paste0("LabelProjection_referenceISC_scatter", ".pdf"))

pdf(pdf_file, width = 8*1.1, height = 6*1.1)
for(a in cons_ref) print(a)
dev.off()
```

## categorize ISC

```{r}
excl <- c("ground_truth", "Ensemble", "Random")
ths <- 0.7
# filter for only references
cons_ref_cat <- lapply(names(res), function(r){
  # Summarise prediction F1 per dataset, classifier, and cell type
  class_levs <- names(rankings_palette[[r]])[!names(rankings_palette[[r]]) %in% excl]
  top_6 <- c(head(class_levs, 3), tail(class_levs,3))
  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier,
                               levels = c(class_levs))) %>% 
    filter(classifier != "ground_truth",
           !is.na(f1),
           classifier %in% top_6) %>% 
    select(cell_type, dataset_id, f1, classifier)
  
  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth",
           !is.na(.data[[isc]])) %>% 
    select(cell_type, dataset_id, .data[[isc]]) %>% 
    mutate(consistency = ifelse(.data[[isc]] > ths,
                              "consistent",
                              "inconsistent"))
  
  # Join reference consistency with prediction F1
  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(.data[[isc]])) %>% 
    group_by(dataset_id, cell_type)
  
  
  
  # Plot: F1 vs reference consistency product with facets and correlation stats
  p_ref <- joined %>% 
    ggplot(aes(x = consistency, y = f1)) +
    geom_jitter(alpha = 0.3, color = "grey", width = 0.15) +
    geom_boxplot(aes(color = classifier),
                 fill = NA,
                 outlier.colour = NA,
                 show.legend = FALSE) +
    scale_color_manual(values = rankings_palette[[1]]) +
    labs(
      title = r,
      subtitle = paste0("F1 in consistent (ISC>", ths,
                        "), and incosistent cell types in train set"),
      x = "ISC category",
      y = "F1 (predictions)"
    ) +
    ggpubr::theme_classic2() +
    facet_wrap(~classifier)
  
  return(p_ref)
})

ggarrange(plotlist = cons_ref_cat)
```


## Export plots
```{r}
# Save as PDF
pdf_file <- file.path(fig_dir, paste0("LabelProjection_", "TrainISC.pdf"))
ggsave(
  pdf_file,
  plot = p_ref,
  width = 7,
  height = 5,
  device = "pdf"
)

```

## contigency plot

```{r density_contigency}
ths_f1 <- 0.7
ths_isc <- 0.5
xcl <- c("ground_truth", "Ensemble", "Random")
# filter for only references
cons_ref_contingency <- lapply(names(res), function(r){
  # Summarise prediction F1 per dataset, classifier, and cell type
  class_levs <- names(rankings_palette[[1]])[!names(rankings_palette[[1]]) %in% excl]
  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier,
                               levels = c(class_levs))) %>% 
    filter(classifier != "ground_truth",
           !is.na(f1),
           ) %>% 
    select(cell_type, dataset_id, f1, classifier)
  
  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth",
           !is.na(.data[[isc]])) %>% 
    select(cell_type, dataset_id, .data[[isc]])
  
  # Join reference consistency with prediction F1
  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(.data[[isc]])) %>% 
    group_by(dataset_id, cell_type)


  
  # Compute quadrant counts (threshold = 0.5 on both axes)
  quadrant_counts <- joined %>%
    ungroup() %>%
    mutate(quadrant = case_when(
      .data[[isc]] >= ths_isc & f1 >= ths_f1 ~ "top-right",
      .data[[isc]] <  ths_isc & f1 >= ths_f1 ~ "top-left",
      .data[[isc]] >= ths_isc & f1 <  ths_f1 ~ "bottom-right",
      TRUE                        ~ "bottom-left"
    )) %>%
    count(quadrant) %>%
    mutate(
      x = case_when(quadrant == "top-right"    ~ 0.75,
                    quadrant == "top-left"      ~ 0.25,
                    quadrant == "bottom-right"  ~ 0.75,
                    quadrant == "bottom-left"   ~ 0.25),
      y = case_when(quadrant == "top-right"    ~ 0.85,
                    quadrant == "top-left"      ~ 0.85,
                    quadrant == "bottom-right"  ~ 0.25,
                    quadrant == "bottom-left"   ~ 0.25),
      #label = n
      label = paste0(sprintf("%.1f", 100 * n / sum(n)), "%")
    )

  # Plot: bin2d density + quadrant percentages
  p_ref <- joined %>%
    ggplot(aes(x = .data[[isc]], y = f1)) +
    geom_bin2d(bins = 50, show.legend = T) +
    scale_fill_viridis_c(option = "magma", trans = "sqrt") +
    geom_vline(xintercept = ths_isc, linetype = "dashed", color = "#00FFFF", linewidth = 0.8) +
    geom_hline(yintercept = ths_f1, linetype = "dashed", color = "#00FFFF", linewidth = 0.8) +
    geom_label(data = quadrant_counts,
              aes(x = x, y = y, label = label),
              inherit.aes = FALSE,
              size = 6,
              fill = "grey90",
              color = "black",
              alpha = 0.9) +
    scale_x_continuous(limits = c(0, 1)) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title = gsub("_", "-", r),
      subtitle = "Train ISC vs Prediction F1, per cell type",
      x = "ISC Train",
      y = "F1 score Test"
    ) +
    ggpubr::theme_classic2() +
    theme(legend.position = "right")
  
  return(p_ref)
})

ggarrange(plotlist = cons_ref_contingency)
```

```{r}
pdf_file <- file.path(fig_dir, paste0("LabelProjection_referenceISC_contigency", ".pdf"))

pdf(pdf_file, width = 4.5, height = 4)
for(a in cons_ref_contingency) print(a)
dev.off()
```


```{r upper_left_quadrant_depletion}
# Test whether upper-left quadrant is depleted vs 25% null expectation
quadrant_stats_plots <- lapply(names(res), function(r){
  class_levs <- names(rankings_palette[[1]])[!names(rankings_palette[[1]]) %in% excl]

  pred_summary <- res[[r]] %>%
    mutate(classifier = factor(classifier, levels = c(class_levs))) %>%
    filter(classifier != "ground_truth", !is.na(f1)) %>%
    select(cell_type, dataset_id, f1, classifier)

  cons_ref <- res[[r]] %>%
    filter(classifier == "ground_truth", !is.na(.data[[isc]])) %>%
    select(cell_type, dataset_id, .data[[isc]])

  joined <- pred_summary %>%
    left_join(cons_ref, by = c("dataset_id", "cell_type")) %>%
    filter(!is.na(.data[[isc]]))

  quad_tbl <- joined %>%
    mutate(quadrant = case_when(
      .data[[isc]] >= ths_isc & f1 >= ths_f1 ~ "top-right",
      .data[[isc]] <  ths_isc & f1 >= ths_f1 ~ "top-left",
      .data[[isc]] >= ths_isc & f1 <  ths_f1 ~ "bottom-right",
      TRUE                              ~ "bottom-left"
    )) %>%
    count(quadrant, .drop = FALSE)

  all_quadrants <- c("top-left", "top-right", "bottom-left", "bottom-right")
  quad_tbl <- dplyr::full_join(
    tibble::tibble(quadrant = all_quadrants),
    quad_tbl,
    by = "quadrant"
  ) %>%
    mutate(n = ifelse(is.na(n), 0L, n))

  n_total <- sum(quad_tbl$n)
  n_ul <- quad_tbl$n[quad_tbl$quadrant == "top-left"]
  bt <- binom.test(x = n_ul, n = n_total, p = 0.25, alternative = "less")

  quad_tbl <- quad_tbl %>%
    rowwise() %>%
    mutate(
      prop = n / n_total,
      ci_low = binom.test(n, n_total)$conf.int[1],
      ci_high = binom.test(n, n_total)$conf.int[2]
    ) %>%
    ungroup() %>%
    mutate(
      quadrant = factor(quadrant, levels = all_quadrants),
      is_top_left = quadrant == "top-left"
    )

  p_quad <- ggplot(quad_tbl, aes(x = quadrant, y = prop, fill = is_top_left)) +
    geom_col(width = 0.7) +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15) +
    geom_hline(yintercept = 0.25, linetype = "dashed", color = "#00FFFF", linewidth = 0.7) +
    geom_text(aes(label = paste0(sprintf("%.1f", 100 * prop), "%")),
              vjust = -0.4,
              size = 3.6) +
    scale_fill_manual(values = c("TRUE" = "#F94144", "FALSE" = "grey75"), guide = "none") +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    labs(
      title = gsub("_", "-", r),
      subtitle = paste0(
        "Upper-left depletion vs 25% null: p = ",
        format.pval(bt$p.value, digits = 3, eps = 1e-4),
        " (one-sided binomial)"
      ),
      x = "Quadrant",
      y = "Proportion of points"
    ) +
    ggpubr::theme_classic2()

  p_quad
})

ggarrange(plotlist = quadrant_stats_plots)
```

