---
title: "Figure 6 analysis: ISC-guided refinement of atlas-level cell type annotation"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# Introduction

To demonstrate the utility of ISC for identifying and resolving problematic annotations at atlas scale, we applied it to the Human Lung Cell Atlas (HLCA), comprising ~500,000 cells from 486 individuals across 49 datasets with consensus labels curated by six experts (Sikkema et al., 2023).

This scripts follow the ISC analysis of HLCA and the visualization of their results.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

```{r libraries}
# Load packages
library(dplyr)
library(ggplot2)
library(tidyr)
library(BiocParallel)
library(RColorBrewer)
library(Seurat)
library(scTypeEval)

# Custom theme
theme_set(theme_bw(base_size = 12))
```

```{r exporting_dir}
fig_dir <- "exported_figs/6_Figure"
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

Define which ISC paramters
```{r set_ISC}
#black list from scTypeEval package
data(black_list)
bl <- list(black_list$TCR, black_list$Immunoglobulins, black_list$Ygenes) %>% unlist()
# define used ISC
mety <- c("local" = "silhouette | recip_classif:Match",
          "global" = "2label_silhouette | Pseudobulk:Cosine")
# Define dissimilarities to use
dm <- c("Pseudobulk:Cosine",
        "recip_classif:Match")
# min samples to consider a cell type
min_samples <- 5
```


Function to clean names
```{r}
clean_ident_labels <- function(x) {
  x |>
    gsub("\\.\\(", "\n(", x = _) |>   # ".(" -> newline
    gsub("\\.", " ", x = _) |>        # "." -> space
    gsub("\\bcell\\b", "", x = _) |>  # remove standalone "cell"
    gsub(" +", " ", x = _) |>         # collapse multiple spaces
    trimws()
}
```



# Reannotation: Lung Atlas

## load

Obtain raw files from Zenodo (10.5281/zenodo.18921437).

```{r}
lpath <- "data/raw/Sikkema_2023_37291214_b351804c-293e-4aeb-9c4c-043db67f4540.rds"

object <- load_single_cell_object(lpath)
```

## Dissimilarity
```{r}
ident <- "cell_type" # annotation column to use
sample = "sample" # column identifying sample of origin

sub_meta = object[["metadata"]] %>% 
  filter(!is.na(.data[[ident]]))
sub_count_matrix = object[["counts"]][,rownames(sub_meta)]

#create sceval object
sc <- create_scTypeEval(matrix = sub_count_matrix,
                        metadata = sub_meta,
                        black.list = bl)

sc <- run_processing_data(sc,
                          ident = ident,
                          sample = sample,
                          # enforce more cell type by lower threshold of samples
                          min_samples = min_samples,
                          verbose = F)

sc <- run_hvg(sc,
              ngenes = 2000,
              ncores = 4, 
              verbose = F)
# remove single-cell slots processed
#sc@data$`single-cell` <- NULL
# emtpy counts slot to make it light
sc@counts <- Matrix::sparseMatrix(i = integer(0),
                                  j = integer(0),
                                  x = numeric(0),
                                  dims = c(0, 0))

sc <- run_pca(sc,
              verbose = F)

for(mdiss in dm){
  cat("#### Running ", mdiss, " ####\n")
  sc <- run_dissimilarity(sc,
                          method = mdiss,
                          ncores = 8,
                          verbose = F)
}
```

save cache
```{r}
dir.create("cache", showWarnings = F)
saveRDS(sc, "cache/sceval_lungAtlas_original.rds")
```

Plot RMM
Explore consistencies
```{r}
phs <- plot_heatmap(sc,
                    sort.consistency = "silhouette",
                    sort.similarity = "Pseudobulk:Cosine")
```


## Consistency

```{r}
cons0 <- get_consistency(sc,
                         consistency_metric = c("silhouette",
                                                "2label_silhouette")
)

cons <- 
  cons0 %>% 
  mutate(method.type = paste(consistency_metric,
                             dissimilarity_method,
                             sep = " | ")) %>% 
  filter(method.type %in% mety) %>% 
  mutate(method.type = factor(method.type,
                              levels = mety,
                              labels = names(mety))) %>% 
  select(method.type, measure, celltype) %>% 
  pivot_wider(names_from = "method.type",
              values_from = "measure") %>% 
  mutate(measure = global * local
  ) %>% 
  filter(!is.na(measure))

consis <- cons %>%
  arrange(measure) %>%
  mutate(celltype = clean_ident_labels(celltype),
         celltype = factor(celltype, levels = celltype)
         )
```


## plotting
### barplot
```{r}
pl <- consis %>% 
  ggplot(aes(y = reorder(celltype, measure),
             x = measure,
             fill = measure)) +
  geom_col(show.legend = FALSE,
           color = "grey50") +
  scale_fill_gradient(
    low  = "white",
    high = "#2166AC"   # deep blue
  ) +
  labs(y = "Author's annotation",
       x = "Integrated ISC score") +
  ggpubr::theme_classic2() +
  theme(axis.title = element_text(size = 22),
        axis.ticks = element_blank())

pl
```

### Fig 6A - scatter

Joint distribution of per-cell-type ISCL (silhouette on reciprocal-classification match) and ISCG (2-label silhouette on pseudobulk cosine similarity) across atlas labels (“cell_type”). Each point represents one annotated cell type. 

```{r}

linewidth <- 0.5
alpha_line <- 0.6
# Empiric threshold obtain in Figure 5, and 5_Figure.Rmd script.
q_local <- 0.5649059
q_global <- 0.7551685



pl_scatter <- consis$original %>%
  ggplot(aes(x = local,
             y = global)) +
  # threshold line of figure 5
  geom_vline(
    xintercept = q_local,
    linetype = "dashed",
    linewidth = linewidth,
    alpha = alpha_line
  ) +
  geom_hline(
    yintercept = q_global,
    linetype = "dashed",
    linewidth = linewidth,
    alpha = alpha_line
  ) +
  # Points
  geom_point(aes(fill = measure),
             shape = 21,
             size = 3.2,
             stroke = 0.4,
             color = "black",
             alpha = 0.9,
             show.legend = FALSE) +
  
  # Clean blue gradient (Nature-friendly, perceptually smooth)
  scale_fill_gradient(
    low  = "#F7FBFF",
    high = "#08306B"
  ) +
  
  # Repelled labels (better spacing control)
  ggrepel::geom_text_repel(
    aes(label = clean_ident_labels(celltype)),
    size = 3.5,
    max.overlaps = 8,
    box.padding = 0.4,
    point.padding = 0.3,
    segment.size = 0.3,
    segment.color = "grey50",
    min.segment.length = 0,
    seed = 123
  ) +
  
  labs(
    x = "ISC score (Local)",
    y = "ISC score (Global)"
  ) +
  
  coord_cartesian(clip = "off") +
  
  theme_classic(base_size = 12) +
  theme(
    axis.title = element_text(size = 20),
    axis.text  = element_text(size = 11),
    axis.line  = element_line(linewidth = 0.4),
    axis.ticks = element_blank(),
    plot.margin = margin(10, 20, 10, 10)
  )

pl_scatter

```


### export plot

```{r}
fpath <- file.path(fig_dir, paste0("LungAtlas_consistency_Scatter", ".pdf"))
pdf(fpath, width = 5.2, height = 5)
pl_scatter
dev.off()
```

## Reannotation

Within respiratory epithelial populations, club, multiciliated epithelial, and bronchial goblet cells exhibited particularly low ISC. Inspection of these populations revealed strong structuring by tissue of origin (nose, airway and lung parenchyma) (Fig. 6b), accompanied by differential expression of canonical markers consistent with anatomical specialization (Fig. 6c). Epithelial-associated genes such as ELF3 and FXYD3 were enriched in nose-derived cells, whereas SCGB1A1 showed higher expression in airway and lung compartments. These results suggested that transcriptionally and anatomically distinct subpopulations were aggregated under single labels, reducing within-label consistency across samples.
Guided by these observations, we implemented targeted, hypothesis-driven refinements by introducing tissue-resolved sublabels for affected epithelial populations (e.g., separating club cells by nasal, airway, and lung origin). 

```{r}
md <- sub_meta %>% 
  tibble::rownames_to_column("cellid") %>% 
  mutate(annot2 = case_when(
    cell_type == "club cell" & tissue == "respiratory airway" ~ "club cell (airway)",
    cell_type == "club cell" & tissue == "nose" ~ "club cell (nasal)",
    cell_type == "club cell" & tissue == "lung parenchyma" ~ "club cell (lung)",
    ##############################################################################
    cell_type == "multi-ciliated epithelial cell" &
      ann_level_4 == "Deuterosomal" ~
      "multi-ciliated epithelial cell (Deuterosomal)",
    cell_type == "multi-ciliated epithelial cell" &
      ann_level_4 != "Deuterosomal" ~
      "multi-ciliated epithelial cell (Non-Deuterosomal)",
    ##############################################################################
    cell_type %in% c("nasal mucosa goblet cell","bronchial goblet cell") &
      tissue == "respiratory airway"~
      NA,
    cell_type %in% c("nasal mucosa goblet cell","bronchial goblet cell") &
      tissue == "nose" ~
      "mucosa goblet cell (nose)",
    cell_type %in% c("nasal mucosa goblet cell","bronchial goblet cell") &
      tissue == "lung parenchyma" ~
      "mucosa goblet cell (lung)",
    ##############################################################################
    cell_type == "tracheobronchial goblet cell" &
      tissue == "respiratory airway" ~
      "tracheobronchial goblet cell(airway)",
    cell_type == "tracheobronchial goblet cell" &
      tissue == "lung parenchyma" ~
      "tracheobronchial goblet cell (lung)",
    cell_type == "tracheobronchial goblet cell" &
      tissue == "nose" ~
      "tracheobronchial goblet cell(nose)",
    TRUE ~ cell_type  # Default case
  )
  ) %>% 
  tibble::column_to_rownames("cellid")
```

## Consistency reannotation

```{r}
ident <- "annot2"# new proposed annotation
sc <- create_scTypeEval(matrix = sub_count_matrix,
                        metadata = md,
                        black.list = bl)

sc <- run_processing_data(sc,
                         ident = ident,
                         sample = sample,
                         # enforce more cell type by lower threshold of samples
                         min.samples = min.samples,
                         verbose = F)

sc <- run_hvg(sc,
              ngenes = 2000,
              ncores = 4, 
              verbose = F)
# remove single-cell slots processed
#sc@data$`single-cell` <- NULL
# emtpy counts slot to make it light
sc@counts <- Matrix::sparseMatrix(i = integer(0),
                                  j = integer(0),
                                  x = numeric(0),
                                  dims = c(0, 0))

sc <- run_pca(sc,
              verbose = F)

for(mdiss in dm){
  cat("#### Running ", mdiss, " ####\n")
  sc <- run_dissimilarity(sc,
                          method = mdiss,
                          ncores = 8,
                          verbose = F)
}
```

save cache
```{r}
saveRDS(sc, "cache/sceval_lungAtlas_reannot.rds")
```

## Consistency

```{r}
cons0 <- get_consistency(sc,
                         consistency_metric = c("silhouette",
                                                "2label_silhouette")
)

cons <- 
  cons0 %>% 
  mutate(method.type = paste(consistency_metric,
                             dissimilarity_method,
                             sep = " | ")) %>% 
  filter(method.type %in% mety) %>% 
  mutate(method.type = factor(method.type,
                              levels = mety,
                              labels = names(mety))) %>% 
  select(method.type, measure, celltype) %>% 
  pivot_wider(names_from = "method.type",
              values_from = "measure") %>% 
  mutate(measure = global * local
  ) %>% 
  filter(!is.na(measure))

consis2 <- cons %>%
  arrange(measure) %>%
  mutate(celltype = factor(celltype, levels = celltype))
```

## plotting
### Boxplot
```{r}
pl2 <- consis2 %>% 
  ggplot(aes(y = reorder(celltype, measure),
             x = measure,
             fill = measure)) +
  geom_col(show.legend = FALSE,
           color = "grey50") +
  scale_fill_gradient(
    low  = "white",
    high = "#2166AC"   # deep blue
  ) +
  labs(y = "Author's annotation",
       x = "ISC score") +
  ggpubr::theme_classic2() +
  theme(axis.title = element_text(size = 22),
        axis.ticks = element_blank())

pl2
```



## plot improvement

Make a list of consistency before and after reannotation

```{r}
consis <- list(original = consis,
               reannot = consis2)
```


```{r}
dfp <- rbind(consis$original %>% mutate(ident = "original"),
             consis$reannot %>%
               mutate(ident = "reannot")
             ) %>% 
  select(celltype, score = measure, ident) %>% 
  mutate(celltype = scTypeEval:::purge_label(celltype))

dict <- md %>% 
  mutate(cell_type = scTypeEval:::purge_label(cell_type), # starndardize cell type labels
         annot2 = scTypeEval:::purge_label(annot2)
  ) %>% 
  distinct(cell_type, annot2) %>% 
  dplyr::rename("celltype" = "annot2")

spl <- split(dfp, dfp$ident)

spl$reannot <- left_join(spl$reannot, dict, by = "celltype") %>% 
  rename("newannot" = "celltype",
         "celltype" = "cell_type",
         "score2" = "score")

dfp2 <- left_join(spl$original, spl$reannot,
                  by = "celltype") %>% 
  mutate(
    scoref = score2 - score,
    #scoref = log2(score2/score),
    is.newannot = ifelse(celltype == newannot,
                         "no", "yes")
  )
dfp2 <- dfp2[,!grepl("ident", names(dfp2))]

dd <- dfp2 %>% 
  select(-scoref) %>% 
  dplyr::rename("Reannotation" = "score2",
                "Original" = "score") %>% 
  pivot_longer(-c(celltype, newannot, is.newannot),
               names_to = "annotation",
               values_to = "score")

dd %>%
  group_by(annotation) %>% 
  summarize(m = mean(score))
```

### Figure 6E - Grouped Barplots

```{r}
nm_cols <- c(
  "#0072B2",  # blue
  "#D55E00"   # vermillion
)

limits <- c(min(dd$score) * 0.9,
            max(dd$score) * 1.05)

titles_dict <- c("yes" = "Reannotated cell types",
                 "no" = "Unchanged cell types")

pl3 <- lapply(c("yes", "no"), function(i) {
  title <- titles_dict[i]
  dd_sub <- dd %>%
    filter(is.newannot == i) %>%
    mutate(g = paste(celltype, newannot, sep = "__"))
  
  ggplot(dd_sub, aes(annotation, score, fill = annotation)) +
    
    # Boxplot
    geom_boxplot(
      width = 0.6,
      outlier.shape = NA,
      alpha = 0.35,
      color = "black",
      linewidth = 0.4
    ) +
    
    # Connecting lines
    geom_line(
      aes(group = g),
      color = "grey40",
      alpha = 0.35,
      linewidth = 0.4
    ) +
    
    # Points
    geom_point(
      shape = 21,
      size = 2.2,
      stroke = 0.3,
      alpha = 0.8,
      color = "black"
    ) +
    
    scale_fill_manual(values = nm_cols) +
    coord_cartesian(ylim = limits) +
    
    labs(
      y = "Integrated ISC score",
      x = NULL,
      title = title
    ) +
    
    theme_classic(base_size = 12) +
    theme(
      legend.position = "none",
      axis.title.y = element_text(margin = margin(r = 8),
                                  size = 18),
      axis.line = element_line(linewidth = 0.4),
      axis.ticks = element_line(linewidth = 0.4),
      plot.margin = margin(5, 10, 5, 5)
    )
})

pl3 <- ggpubr::ggarrange(plotlist = pl3, nrow = 1)

pl3
```

### export plot

```{r}
fpath <- file.path(fig_dir, paste0("LungAtlas_reannotation", ".pdf"))
pdf(fpath, width = 5, height = 3)
pl3
dev.off()
```

## Reload cache

Reload cache (if needed)
```{r}
lpath <- list(original = "cache/sceval_lungAtlas_original.rds",
              reannot = "cache/sceval_lungAtlas_reannot.rds")
sceval <- lapply(lpath,
                 readRDS)
```

```{r}
# list of cells with confusion problems with club cells

confusion_cells <- c(
  ## worst
  "bronchial goblet cell",
  "club cell",
  "tracheobronchial goblet cell",
  "multi-ciliated epithelial cell",
  "nasal mucosa goblet cell"
)

dict <- sceval$reannot@metadata %>% 
  filter(cell_type %in% confusion_cells) %>% 
  mutate(original = scTypeEval:::purge_label(cell_type),
         reannot = scTypeEval:::purge_label(annot2)
  ) %>% 
  distinct(original, reannot, tissue)
```

# Figure 6D - Reannotation heatmaps

Build RMM(~heatmap) but for only a subset of cells.

RMM using reciprocal classification match for selected low-ISC cell types before and after.

## auxiliary function

```{r custom_heatmap}
scTypeEval <- sceval$reannot
dissimilarity.slot = "RecipClassif:Match"
da = "RecipClassif:Match"
sort.consistency = "silhouette"
sort.similarity = "Pseudobulk:Cosine"
low.color = "black"
high.color = "white"
hclust.method = "ward.D2"
verbose = T
celltype.keep = dict$reannot

custom_heatmap <- function(scTypeEval,
                           dissimilarity.slot = "RecipClassif:Match",
                           sort.similarity = "Pseudobulk:Cosine",
                           sort.consistency = "silhouette",
                           low.color = "black",
                           high.color = "white",
                           hclust.method = "ward.D2",
                           celltype.keep = NULL,
                           verbose = TRUE){
  
  # check if dissimilarity object are present in object
  diss.assays <- scTypeEval:::.check_dissimilarityAssays(scTypeEval, slot = dissimilarity.slot)

  
  # if not indicated just order cell types based on alphabetical
  if(is.null(sort.similarity) && is.null(sort.consistency)){
    if(verbose){message("No ordering (sort.similarity or sort.consistency) indicated,
                          sorting cell type by alphabetical order.")}
  }
  
  # Compute similarity of cell types if indicated
  if(!is.null(sort.similarity)){
    # get dissimilarity matrix
    diss.assays.cluster <- scTypeEval:::.check_dissimilarityAssays(scTypeEval,
                                                                   slot = sort.similarity)
    assay_cluster <- scTypeEval@dissimilarity[[diss.assays.cluster]]
    d_mat_cluster <- as.matrix(assay_cluster@dissimilarity)
    # get similarities between cell types
    if(verbose){message("Computing cell type similarity based on ", diss.assays.cluster, ".")}
    ident_levels <- levels(assay_cluster@ident[[1]])
    mat <- matrix(0, nrow = length(ident_levels),
                  ncol = length(ident_levels),
                  dimnames = list(ident_levels, ident_levels))
    for (i in ident_levels) {
      for (j in ident_levels) {
        sub <- d_mat_cluster[assay_cluster@ident[[1]] == i,
                             assay_cluster@ident[[1]] == j,
                             drop = FALSE]
        mat[i, j] <- mean(sub)
      }
    }
    hc <- stats::hclust(as.dist(mat),
                        method = hclust.method)
    ident_order <- hc$labels[hc$order]
  } else {
    ident_order <- NULL
  }
  
  pls <- lapply(diss.assays,
                function(da){
                  assay <- scTypeEval@dissimilarity[[da]]
                  title <- paste(da, names(assay@ident), sep = " - ")
                  
                  d_mat_plot <- as.matrix(assay@dissimilarity)
                  
                  annot_ident <- assay@ident[[1]]
                  
                  #  order by consistency if indicated
                  if (!is.null(sort.consistency)) {
                    
                    if(verbose){message("Computing consistency metric for ", sort.consistency, ".\n")}
                    
                    consis <- get.Consistency(scTypeEval,
                                              dissimilarity.slot = da,
                                              consistency_metric = sort.consistency,
                                              verbose = FALSE) |>
                      filter(!is.na(measure)) |>
                      dplyr::arrange(measure) |>
                      dplyr::pull(measure, name = celltype)
                    
                    if(verbose){message("Consistency computed.")}
                    
                    ident_order <- names(consis)  
                  } 
                  
                  if(is.null(ident_order)){
                    # order alphabetically
                    ident_order <- sort(levels(annot_ident))
                  }
                  
                  # reorder cells within cell type: first by ident order, then cluster within ident
                  if(!is.null(sort.similarity)){
                    ordering <- unlist(lapply(ident_order, function(iden) {
                      cells <- names(annot_ident)[annot_ident == iden]
                      if (length(cells) > 2) {
                        hc <- stats::hclust(as.dist(d_mat_cluster[cells, cells]),
                                            method = hclust.method)
                        cells[hc$order]
                      } else {
                        cells
                      }
                    }))
                  } else {
                    ordering <- unlist(lapply(ident_order, function(iden) {
                      cells <- names(annot_ident)[annot_ident == iden] |>
                        sort()
                      return(cells)
                    }))
                  }
                  
                  # reorder matrix + annotation
                  d_mat_plot <- d_mat_plot[ordering, ordering, drop = FALSE]
                  
                  # convert matrix to tidy df
                  df <- as.data.frame(as.table(d_mat_plot))
                  colnames(df) <- c("row", "col", "value")
                  
                  if(!is.null(celltype.keep)){
                    celltype.keep <- scTypeEval:::purge_label(celltype.keep)
                    # filter for cell type
                    df <- df %>%
                      filter(
                        stringr::str_extract(row, "(?<=_)[^_]+") %in% celltype.keep,
                        stringr::str_extract(col, "(?<=_)[^_]+") %in% celltype.keep
                      ) %>% 
                      mutate(value = ifelse(value > 0,
                                            1, 0))
                    ident_order <- ident_order[ident_order %in% celltype.keep]
                    annot_ident <- annot_ident[annot_ident %in% celltype.keep]
                    ordering <- ordering[stringr::str_extract(ordering, "(?<=_)[^_]+") %in%
                                           celltype.keep]
                  }
                  
                
                  # ident boundaries + label positions
                  counts <- as.numeric(table(as.character(annot_ident))[ident_order])
                  sep_lines <- cumsum(counts)
                  ident_labels <- ident_order
                  
                  # adjust labels
                  ident_labels <- clean_ident_labels(ident_labels)
                  
                  # midpoints for axis ticks
                  x_ticks <- sep_lines - counts/2 + 0.5
                  y_ticks <- length(ordering) - (sep_lines - counts/2 + 0.5) + 1
                  
                  
                  # build plot
                  p <- ggplot2::ggplot(df, ggplot2::aes(x = col, y = row, fill = value)) +
                    ggplot2::geom_tile() +
                    ggplot2::scale_fill_gradient(low = low.color,
                                                 high = high.color) +
                    ggplot2::scale_x_discrete(
                      breaks = ordering[x_ticks],
                      labels = ident_labels,
                      expand = c(0, 0)
                    ) +
                    ggplot2::scale_y_discrete(
                      breaks = rev(ordering)[y_ticks],
                      labels = ident_labels,
                      expand = c(0, 0)
                    ) +
                    ggplot2::theme_minimal(base_size = 12) +
                    ggplot2::theme(
                      axis.text.x = ggplot2::element_blank(),
                      axis.text.y = ggplot2::element_text(hjust = 1),
                      axis.ticks = ggplot2::element_blank(),
                      axis.title = ggplot2::element_blank(),
                      legend.position = "none",
                      panel.grid = ggplot2::element_blank()
                    ) +
                    ggplot2::geom_hline(yintercept = sep_lines + 0.5,
                                        color = "grey70",
                                        linewidth = 0.4) +
                    ggplot2::geom_vline(xintercept = sep_lines + 0.5,
                                        color = "grey70",
                                        linewidth = 0.4) +
                    ggplot2::labs(title = title)
                  
                  return(p)
                })
  
  names(pls) <- diss.assays
  
  if(length(pls) == 1){
    pls <- pls[[1]]
  }
  
  return(pls)
  
}
```

## Plots production
```{r}
nm_cols <- c(
  "original" = "#0072B2",  # blue
  "reannot" = "#D55E00"   # vermillion
)

focus_heatmap <- lapply(names(sceval),
                        function(s){
                          custom_heatmap(sceval[[s]],
                                         celltype.keep = dict[[s]],
                                         low.color = nm_cols[s]) +
                            ggtitle(s)
                        })

# tune params
focus_heatmap 
```

### export focus heatmaps
```{r}
w <- 7
h <- 5

fpath <- file.path(fig_dir, paste0("LungAtlas_consistency_QRfocus_original", ".pdf"))
pdf(fpath, width = w, height = h)
focus_heatmap[[1]] +
  ggtitle("Original")
dev.off()

fpath <- file.path(fig_dir, paste0("LungAtlas_consistency_QRfocus_Reannotation", ".pdf"))
pdf(fpath, width = w, height = h)
focus_heatmap[[2]] +
  ggtitle("Reannotation")
dev.off()

```

# PCAs
Let's produce embeddings for pca grouping by cell type and sample only the cells of interest (reanntoated):

utilities
```{r}
g = c("FXYD3", "EPCAM",
      "ELF3","SCGB1A1")
library(STACAS)
data("EnsemblGeneTable.Hs")

# scaling of gene expression
scale_to_range <- function(x, new_min = -1, new_max = 1) {
  old_min <- min(x, na.rm = TRUE)
  old_max <- max(x, na.rm = TRUE)
  (x - old_min) / (old_max - old_min) * (new_max - new_min) + new_min
}


get_normExpression <- function(scTypeEval,
                               genes = c("FXYD3", "EPCAM",
                                         "ELF3","SCGB1A1"),
                               assay = "pseudobulk"){
  #Function to translate Ensemble_id (HCLA) to gene symbols.
  genes_dict <- EnsemblGeneTable.Hs %>% 
    dplyr::filter(`Gene name` %in% genes) %>% 
    dplyr::distinct(`Gene stable ID`, `Gene name`)
  
  mat <- scTypeEval@data[[assay]]@matrix[
    genes_dict$`Gene stable ID`, 
  ] %>% 
    t()
  
  # scale expression
  mat <- apply(mat, 2, function(e) scale_to_range(e, -2, 2)) %>% 
    as.data.frame()
  
  # ---- rename columns using match ----
  idx <- match(colnames(mat), genes_dict$`Gene stable ID`)
  colnames(mat) <- genes_dict$`Gene name`[idx]
  
  return(mat)
}
```

Get embeddings and gene expression in the same dataframe
```{r}
get_embeddings <- function(sceval, reduction, dims = c(1,2)){
  embeds <- sceval@reductions[[reduction]]@embeddings
  idents <- sceval@reductions[[reduction]]@ident[[1]]
  samples <- sceval@reductions[[reduction]]@sample
  # compute variance of PCs
  vrs <- scTypeEval:::var_PCA(t(embeds))[dims]
  vrs <- round(vrs*100, 2)
  # filter idents
  idents_keep <- idents %in% scTypeEval:::purge_label(confusion_cells)
  
  celltypes <- idents[idents_keep]
  sample <- samples[idents_keep]
  
  md <- sceval@metadata %>% 
    mutate(sample = scTypeEval:::purge_label(sample),
           celltype = scTypeEval:::purge_label(cell_type)) %>% 
    filter(celltype %in% celltypes) %>% 
    dplyr::select(sample, tissue)
  
  embeds_df <- embeds[dims, idents_keep] %>% 
    t() %>% as.data.frame() %>% 
    mutate(celltype = clean_ident_labels(celltypes),
           sample = sample) %>% 
    left_join(., md, by= "sample") %>% 
    distinct()
  
  # add genes expression of genes
  gene_expr <- get_normExpression(sceval,
                                  assay = reduction)
  
  embeds_df <- cbind(embeds_df, gene_expr)
  
  return(list(mat = embeds[, idents_keep],
              df = embeds_df,
              variance = vrs)
         )
}
```



## Recalculate embeddings

Get PCA embeddings focuins only on the cells of interest.

```{r}
ident <- "cell_type"
sample = "sample"

sub_meta = object[["metadata"]] %>% 
  filter(!is.na(.data[[ident]]),
         .data[[ident]] %in% confusion_cells)
sub_count_matrix = object[["counts"]][,rownames(sub_meta)]

sc <- create_scTypeEval(matrix = sub_count_matrix,
                        metadata = sub_meta,
                        black.list = sub_meta)

sc <- run_processing_data(sc,
                         ident = ident,
                         sample = sample,
                         # enforce more cell type by lower threshold of samples
                         min.samples = min.samples,
                         verbose = F)

sc <- run_hvg(sc,
              ngenes = 2000,
              ncores = 6, 
              verbose = F)
# emtpy counts slot to make it light
sc@counts <- Matrix::sparseMatrix(i = integer(0),
                                  j = integer(0),
                                  x = numeric(0),
                                  dims = c(0, 0))

sc <- run_pca(sc,
              verbose = F)

plot.PCA(sc)
```

```{r}
embeds_part <- lapply(names(sc@reductions),
                       function(x){
                         get_embeddings(sc, x)
                       })
names(embeds_part) <- names(sc@reductions)
```

Coloring

```{r}
celltype_cols <- c(
  "club"                      = "#0072B2",  # blue
  "nasal mucosa goblet"       = "#009E73",  # bluish green
  "tracheobronchial goblet"   = "#D55E00",  # vermillion
  "multi ciliated epithelial" = "#CC79A7",  # purple
  "bronchial goblet"          = "grey15" 
)


tissue_cols <- c(
  "nose"               = "#4C72B0",  # blue
  "respiratory airway" = "#E69F00",  # orange
  "lung parenchyma"    = "#CC79A7"  # bluish green
)
```

# Figure 6B - Low dimensional representations

### scatter celltypes

Function to produce scatter plot based on datafram with columns
1 - Embedding 1
2- Embedding2
3 - Coloring

```{r}
helper.plot.scatter <- function(df,
                                show.legend = FALSE,
                                label = TRUE,
                                dot.size = 5,
                                label.size = 8,
                                color.by = "celltype",
                                label.by = "celltype",
                                vars = NULL,
                                keep.square = TRUE) {
   # Replace dots with spaces in the ident column
   
  pl <- ggplot2::ggplot() +
    ggplot2::geom_point(
      data = df,
      ggplot2::aes(
        x = .data[[names(df)[1]]],  # First column
        y = .data[[names(df)[2]]],  # Second column
        color = .data[[color.by]]
      ),
      alpha = 0.8,
      size = dot.size,
      show.legend = show.legend) +
    ggpubr::theme_classic2() +
    ggplot2::theme(legend.text = ggplot2::element_text(size = 12, face = "bold"),
                   axis.title = ggplot2::element_text(size = 22),
                   axis.text = ggplot2::element_blank(),
                   axis.ticks = ggplot2::element_blank(),
                   axis.line = ggplot2::element_blank()
    )

  if (keep.square) {
    pl <- pl +
      ggplot2::coord_fixed() +
      ggplot2::theme(aspect.ratio = 1)
  }
  
   if(label){
      # Calculate centroids and format labels
      cnt <- df |>
         dplyr::group_by(.data[[label.by]]) |>
         dplyr::summarize(
            x = mean(.data[[names(df)[1]]]),
            y = mean(.data[[names(df)[2]]]),
         )
      # Always show labels for mentioned cell types
      pl <- pl +
        ggrepel::geom_label_repel(
          data = cnt,
          ggplot2::aes(
            x = x,
            y = y,
            label = .data[[label.by]],
            color = .data[[label.by]],
          ),
          alpha = 0.9,
          size = label.size,
          max.overlaps = Inf,
          show.legend = FALSE
        )
   }
  
  if(!is.null(vars)){
    labs <- paste0(names(vars), " (", vars, "%)")
    
    pl <- 
      pl +
      ggplot2::labs(x = labs[1],
                    y = labs[2])
  }
   
   return(pl)
}
```

### pca
```{r}
pca_celltype <- helper.plot.scatter(df = embeds_part$pseudobulk$df,
                                    vars = embeds_part$pseudobulk$variance) +
  scale_color_manual(values = celltype_cols)
pca_tissue <- helper.plot.scatter(df = embeds_part$pseudobulk$df,
                                  label = F,
                                  color.by = "tissue",
                                  show.legend = T,
                                  vars = embeds_part$pseudobulk$variance)+
  scale_color_manual(values = tissue_cols)

pca_celltype; pca_tissue
```

### umap

Same but with umap
```{r}
# obtain umap embeddings from 30 PCs PCA
um <- uwot::umap(embeds_part$`single-cell`[["mat"]] %>% t(),
                 seed = 22
                 # metric = "cosine",
                 # min_dist = 0.3,
                 # n_neighbors = 10
)

df <- um |>
  as.data.frame()

names(df) <- c("umap_1", "umap_2")

df <- cbind(df, embeds_part$`single-cell`$df[,-c(1,2)])

embeds_part$`single-cell`[["umap"]] <- df

```


```{r}
dt <- 0.5
umap_celltype <- helper.plot.scatter(df = embeds_part$`single-cell`$umap,
                                     dot.size = dt) +
  scale_color_manual(values = celltype_cols)
umap_tissue <- helper.plot.scatter(df = embeds_part$`single-cell`$umap,
                                  label = F,
                                  dot.size = dt,
                                  color.by = "tissue",
                                  show.legend = T)+
  scale_color_manual(values = tissue_cols)

umap_celltype; umap_tissue
```



## expression
Show coloring as scaled expression of gene interest in low embeddings
### pca
```{r}
# PCA
pca_expr <- lapply(g, function(gene){
  helper.plot.scatter(df = embeds_part$pseudobulk$df,
                      vars = embeds_part$pseudobulk$variance,
                      label = F,
                      show.legend = T,
                      color.by = gene) +
    scale_color_gradient2(
      low = "grey",
      mid = "#F5C6A5",   # soft warm beige (NOT white)
      high = "red3",
      midpoint = 0
    ) +
    theme(legend.position = "top")
}
)

# UMAP
umap_expr <- lapply(g, function(gene){
  helper.plot.scatter(df = embeds_part$`single-cell`$umap,
                      label = F,
                      show.legend = F,
                      dot.size = dt,
                      color.by = gene) +
     scale_color_gradient2(
      low = "grey",
      mid = "#F5C6A5",   # soft warm beige (NOT white)
      high = "red3",
      midpoint = 0
    ) +
    theme(legend.position = "top")
}
)

g; umap_expr
```


# export plots scatters

```{r}

sanitize_name <- function(x) {
  gsub("[^A-Za-z0-9_-]", "_", x)
}

w <- 5
h <- 5

# PCA/UMAP scatter plots
plots_to_export_nl <- list(
  "LungAtlas_reannotation_PCA_celltype" = pca_celltype,
  "LungAtlas_reannotation_UMAP_celltype" = umap_celltype
)

plots_to_export_legend<- list(
  "LungAtlas_reannotation_PCA_tissue" = pca_tissue,
  "LungAtlas_reannotation_UMAP_tissue" = umap_tissue
)


for (nm in names(plots_to_export_nl)) {
  fpath <- file.path(fig_dir, paste0(nm, ".pdf"))
  pdf(fpath, width = w, height = h)
  print(plots_to_export[[nm]])
  dev.off()
}

for (nm in names(plots_to_export_legend)) {
  fpath <- file.path(fig_dir, paste0(nm, ".pdf"))
  pdf(fpath, width = 7, height = h)
  print(plots_to_export[[nm]])
  dev.off()
}

# Expression feature plots (PCA)
names(pca_expr) <- g
for (gene in names(pca_expr)) {
  fpath <- file.path(fig_dir, paste0("LungAtlas_reannotation_PCA_", sanitize_name(gene), ".pdf"))
  pdf(fpath, width = w, height = 5.5)
  print(pca_expr[[gene]])
  dev.off()
}

# Expression feature plots (UMAP)
names(umap_expr) <- g
for (gene in names(umap_expr)) {
  fpath <- file.path(fig_dir, paste0("LungAtlas_reannotation_UMAP_", sanitize_name(gene), ".pdf"))
  pdf(fpath, width = w, height = 5.5)
  print(umap_expr[[gene]])
  dev.off()
}

```


# Figure 6F and Extended Data Figure 5 - Improve consistency scatter plot

Get consistency if not loaded

```{r}
consis <- lapply(sceval, function(sc){
  cons0 <- get_consistency(sc,
                           consistency_metric = c("silhouette",
                                                  "2label_silhouette")
  )
  
  cons <- 
    cons0 %>% 
    mutate(method.type = paste(consistency_metric,
                               dissimilarity_method,
                               sep = " | ")) %>% 
    filter(method.type %in% mety) %>% 
    mutate(method.type = factor(method.type,
                                levels = mety,
                                labels = names(mety))) %>% 
    select(method.type, measure, celltype) %>% 
    pivot_wider(names_from = "method.type",
                values_from = "measure") %>% 
    mutate(measure = global * local
    ) %>% 
    filter(!is.na(measure))
  
  cons %>%
    arrange(measure)
    
})
```

```{r}
dfp <- rbind(consis$original %>% mutate(ident = "original"),
             consis$reannot %>%
               mutate(ident = "reannot",
                      celltype = recode(celltype,
                          "club.cell.(awirway)" = "club.cell.(airway)"))
             ) %>%
  mutate(celltype = scTypeEval:::purge_label(celltype))

dict <- md %>% 
  mutate(cell_type = scTypeEval:::purge_label(cell_type),
         annot2 = scTypeEval:::purge_label(annot2)
  ) %>% 
  distinct(cell_type, annot2) %>% 
  dplyr::rename("celltype" = "annot2")

dfp2 <- left_join(dfp, dict, by = "celltype") %>% 
  mutate(pair_id = coalesce(cell_type, celltype)) %>%
  filter(!(ident == "reannot" & celltype == cell_type)) %>% 
  mutate(highlight = ifelse(is.na(cell_type) | ident == "reannot",
         "yes", "no"),
         celltype = clean_ident_labels(celltype),
         cell_type = clean_ident_labels(cell_type)
         )


```

## scatter improvement
```{r}
linewidth <- 0.5
alpha_line <- 0.6
# empirical threshold found in 5_Figure.Rmd
q_local <- 0.5649059
q_global <- 0.7551685

cols <- c("original"  = "grey20", "reannot" = "red4")
vals <- c("yes" = 1, "no" = 0.3)
imp_scatter <- dfp2 %>%
  ggplot(aes(x = local,
             y = global,
             alpha = highlight)) +
    # threshold line of figure 5
  geom_vline(
    xintercept = q_local,
    linetype = "dashed",
    linewidth = linewidth,
    alpha = alpha_line
  ) +
  geom_hline(
    yintercept = q_global,
    linetype = "dashed",
    linewidth = linewidth,
    alpha = alpha_line
  ) +
  
  # Points
  geom_point(aes(fill = ident),
             shape = 21,
             size = 3.2,
             stroke = 0.4,
             color = "black",
             show.legend = T) +
  
  scale_fill_manual(values = cols) +
  
  # Repelled labels (better spacing control)
  ggrepel::geom_text_repel(
    aes(label = ifelse(highlight == "yes",
                       celltype, NA),
        color = ident),
    size = 3.5,
    max.overlaps = 20,
    box.padding = 0.4,
    point.padding = 0.3,
    segment.size = 0.3,
    segment.color = "grey15",
    min.segment.length = 0,
    seed = 123,
    show.legend = F
  ) +
  scale_color_manual(values = cols) +
  scale_alpha_manual(values = vals) +
  guides(alpha = "none") +
  
  labs(
    x = "ISC score (Local)",
    y = "ISC score (Global)"
  ) +
  
  coord_cartesian(clip = "off") +
  
  theme_classic(base_size = 12) +
  theme(
    axis.title = element_text(size = 20),
    axis.text  = element_text(size = 11),
    axis.line  = element_line(linewidth = 0.4),
    axis.ticks = element_blank(),
    plot.margin = margin(10, 20, 10, 10)
  )

imp_scatter
```

```{r}
fpath <- file.path(fig_dir, paste0("LungAtlas_consistency_Scatter_improvement", ".pdf"))
pdf(fpath, width = 6*1.2, height = 5*1.2)
imp_scatter
dev.off()
```










