---
title: "Heart Atlas analysis"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# Introduction

To analyze heart single-cell atlas annotation with ISC

https://www.heartcellatlas.org/v1.html


# 0. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

```{r libraries}
# Load packages
library(dplyr)
library(ggplot2)
library(tidyr)
library(BiocParallel)
library(Seurat)
library(scTypeEval)
source("helpers/consistency_exploration.R")
# Custom theme
theme_set(theme_bw(base_size = 12))
```

Define which ISC paramters
```{r set_ISC}
#black list from scTypeEval package
data(black_list)
bl <- list(black_list$TCR, black_list$Immunoglobulins, black_list$Ygenes) %>% unlist()
# define used ISC
mety <- c("local" = "silhouette | recip_classif:Match",
          "global" = "2label_silhouette | Pseudobulk:Cosine")
# Define dissimilarities to use
dm <- c("Pseudobulk:Cosine",
        "recip_classif:Match")
# min samples to consider a cell type
min_samples <- 3
min_cells <- 5
```


# 1. Load dataset

```{r}
lpath <- "~/Documents/Projects/ISC_benchmark_reproducibility/data/hca_subsampled_20k.h5ad"

object <- load_single_cell_object(lpath)

ident <- "cell_type" # annotation column to use
sample = "donor" # column identifying sample of origin

n_distinct(object$metadata[[sample]])
unique(object$metadata[[ident]])
table(is.na(object$metadata[[ident]]))
```


# 2. Run scTypeEval

## 2.1 Processing

```{r}
#create sceval object
sc <- create_scTypeEval(matrix = object$counts,
                        metadata = object$metadata,
                        black_list = bl)

sc <- run_processing_data(sc,
                          ident = ident,
                          sample = sample,
                          # enforce more cell type by lower threshold of samples
                          min_samples = min_samples,
                          min_cells = min_cells,
                          verbose = F)

sc <- run_hvg(sc,
              ngenes = 2000,
              ncores = 4, 
              verbose = F)

sc <- run_pca(sc,
              verbose = F)

for(mdiss in dm){
  cat("#### Running ", mdiss, " ####\n")
  sc <- run_dissimilarity(sc,
                          method = mdiss,
                          ncores = 6,
                          verbose = F)
}
```

## 2.2 Explore plots

### RMC plot
```{r}
plot_heatmap(sc, 
             dissimilarity_slot = "recip_classif:Match",
             sort_consistency = "silhouette",
             sort_similarity = "Pseudobulk:Cosine")
```
### PCA

```{r}
plot_pca(sc)
```


# 3. Consistency

```{r}
cons0 <- get_consistency(sc,
                         consistency_metric = c("silhouette",
                                                "2label_silhouette")
)

cons <- 
  cons0 %>% 
  mutate(method.type = paste(consistency_metric,
                             dissimilarity_method,
                             sep = " | ")) %>% 
  filter(method.type %in% mety) %>% 
  mutate(method.type = factor(method.type,
                              levels = mety,
                              labels = names(mety))) %>% 
  select(method.type, measure, celltype) %>% 
  pivot_wider(names_from = "method.type",
              values_from = "measure") %>% 
  mutate(measure = global * local
  ) %>% 
  filter(!is.na(measure))

consis <- cons %>%
  arrange(measure)
```

## 3.1. plotting
### barplot
```{r}
q_local <- 0.45
q_global <- 0.74
consistency_barplot(consis, x_threshold = q_local)
```
### scatter

Joint distribution of per-cell-type ISCL (silhouette on reciprocal-classification match) and ISCG (2-label silhouette on pseudobulk cosine similarity) across atlas labels (“cell_type”). Each point represents one annotated cell type. 

```{r}
consistency_scatter(consis)
```
## 4. SCCAF
```{r}
repo_root <- if (dir.exists("sample_agnostic_utils")) {
  getwd()
} else {
  normalizePath("..", mustWork = TRUE)
}
source(file.path(repo_root, "sample_agnostic_utils", "run_python_pipelines.R"))

sccaf_results <- run_sample_agnostic_python_pipelines(
  scTypeEval = sc,
  pipelines = list(
    sccaf = pipeline_spec_sccaf(cluster_key = ident, n = 100)
  ),
  file_prefix = basename(tempfile("heart_")),
  continue_on_error = FALSE,
  cleanup = TRUE
)$sccaf

sccaf_results
```

### barplot
```{r}
consistency_barplot(sccaf_results,
                    xval = "f1_score",
                    xlabel = "SCCAF F1")
```



## 5. MetaNeighbor

```{r}
library(MetaNeighbor)

meta_neighbor_results <- get_consistency(
  scTypeEval = sc,
  dissimilarity_slot = NA_character_,
  consistency_metric = "MetaNeighbor_Supervised"
)

meta_neighbor_results
```
### barplot
```{r}
consistency_barplot(meta_neighbor_results,
                    xval = "measure",
                    x_threshold = 0.5)
```


