---
title: "ISC-guided refinement of stressed T cells"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# Introduction

To demonstrate the utility of ISC for identifying and resolving problematic annotations on [Chu et al., 2023](https://www.nature.com/articles/s41591-023-02371-y) which identifies T-stressed cell subsets.


# 0. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

```{r libraries}
# Load packages
library(dplyr)
library(ggplot2)
library(tidyr)
library(BiocParallel)
library(Seurat)
library(scTypeEval)

# Custom theme
theme_set(theme_bw(base_size = 12))
```

Define which ISC paramters
```{r set_ISC}
#black list from scTypeEval package
data(black_list)
bl <- list(black_list$TCR, black_list$Immunoglobulins, black_list$Ygenes) %>% unlist()
# define used ISC
mety <- c("local" = "silhouette | recip_classif:Match",
          "global" = "2label_silhouette | Pseudobulk:Cosine")
# Define dissimilarities to use
dm <- c("Pseudobulk:Cosine",
        "recip_classif:Match")
# min samples to consider a cell type
min_samples <- 3
min_cells <- 10
```


# 1. Load dataset

```{r}
lpath <- "~/Documents/Projects/Pan-cancer_human_TILs/cache/Yost_Stressed_annotation.rds"

object <- load_single_cell_object(lpath)
object$metadata %>%
   distinct(cell.type, Celltype, Cluster,
            Cellsubtype, celltype)

ident <- "celltype" # annotation column to use
sample = "sample" # column identifying sample of origin

n_distinct(object$metadata[[sample]])
unique(object$metadata[[ident]])
table(is.na(object$metadata[[ident]]))
```


# 2. Run scTypeEval

## 2.1 Processing

```{r}
#create sceval object
sc <- create_scTypeEval(matrix = object$counts,
                        metadata = object$metadata,
                        black_list = bl)

sc <- run_processing_data(sc,
                          ident = ident,
                          sample = sample,
                          # enforce more cell type by lower threshold of samples
                          min_samples = min_samples,
                          min_cells = min_cells,
                          verbose = F)

sc <- run_hvg(sc,
              ngenes = 2000,
              ncores = 4, 
              verbose = F)

sc <- run_pca(sc,
              verbose = F)

for(mdiss in dm){
  cat("#### Running ", mdiss, " ####\n")
  sc <- run_dissimilarity(sc,
                          method = mdiss,
                          ncores = 6,
                          verbose = F)
}
```

## 2.2 Explore plots

### RMC plot
```{r}
plot_heatmap(sc, 
             dissimilarity_slot = "recip_classif:Match",
             sort_consistency = "silhouette",
             sort_similarity = "Pseudobulk:Cosine")
```
### PCA

```{r}
plot_pca(sc)
```


# 3. Consistency

```{r}
cons0 <- get_consistency(sc,
                         consistency_metric = c("silhouette",
                                                "2label_silhouette")
)

cons <- 
  cons0 %>% 
  mutate(method.type = paste(consistency_metric,
                             dissimilarity_method,
                             sep = " | ")) %>% 
  filter(method.type %in% mety) %>% 
  mutate(method.type = factor(method.type,
                              levels = mety,
                              labels = names(mety))) %>% 
  select(method.type, measure, celltype) %>% 
  pivot_wider(names_from = "method.type",
              values_from = "measure") %>% 
  mutate(measure = global * local
  ) %>% 
  filter(!is.na(measure))

consis <- cons %>%
  arrange(measure)
```

## 3.1. plotting
### barplot
```{r}
yval <- "local"
pl <- consis %>% 
  ggplot(aes(y = reorder(celltype, .data[[yval]]),
             x = .data[[yval]],
             fill = .data[[yval]])) +
  geom_col(show.legend = FALSE,
           color = "grey50") +
  scale_fill_gradient(
    low  = "white",
    high = "#2166AC"   # deep blue
  ) +
  labs(y = "Author's annotation",
       x = "ISC score") +
  ggpubr::theme_classic2() +
  theme(axis.title = element_text(size = 22),
        axis.ticks = element_blank())

pl
```
### scatter

Joint distribution of per-cell-type ISCL (silhouette on reciprocal-classification match) and ISCG (2-label silhouette on pseudobulk cosine similarity) across atlas labels (“cell_type”). Each point represents one annotated cell type. 

```{r}
# empirical threshold found in 5_Figure.Rmd
q_local <- 0.45
q_global <- 0.74

linewidth <- 0.5
alpha_line <- 0.6

pl_scatter <- consis %>%
  ggplot(aes(x = local,
             y = global)) +
  # threshold line of figure 5
  geom_vline(
    xintercept = q_local,
    linetype = "dashed",
    linewidth = linewidth,
    alpha = alpha_line
  ) +
  geom_hline(
    yintercept = q_global,
    linetype = "dashed",
    linewidth = linewidth,
    alpha = alpha_line
  ) +
  # Points
  geom_point(aes(fill = measure),
             shape = 21,
             size = 3.2,
             stroke = 0.4,
             color = "black",
             alpha = 0.9,
             show.legend = FALSE) +
  
  # Clean blue gradient (Nature-friendly, perceptually smooth)
  scale_fill_gradient(
    low  = "#F7FBFF",
    high = "#08306B"
  ) +
  
  # Repelled labels (better spacing control)
  ggrepel::geom_text_repel(
    aes(label = celltype),
    size = 3.5,
    max.overlaps = 8,
    box.padding = 0.4,
    point.padding = 0.3,
    segment.size = 0.3,
    segment.color = "grey50",
    min.segment.length = 0,
    seed = 123
  ) +
  
  labs(
    x = "ISC score (Local)",
    y = "ISC score (Global)"
  ) +
  
  coord_cartesian(clip = "off") +
  
  theme_classic(base_size = 12) +
  theme(
    axis.title = element_text(size = 20),
    axis.text  = element_text(size = 11),
    axis.line  = element_line(linewidth = 0.4),
    axis.ticks = element_blank(),
    plot.margin = margin(10, 20, 10, 10)
  )

pl_scatter

```
## 4. SCCAF
```{r}
repo_root <- if (dir.exists("sample_agnostic_utils")) {
  getwd()
} else {
  normalizePath("..", mustWork = TRUE)
}
source(file.path(repo_root, "sample_agnostic_utils", "run_python_pipelines.R"))

sccaf_results <- run_sample_agnostic_python_pipelines(
  scTypeEval = sc,
  pipelines = list(
    sccaf = pipeline_spec_sccaf(cluster_key = ident, n = 100)
  ),
  file_prefix = basename(tempfile("stressed_tcells_")),
  continue_on_error = FALSE,
  cleanup = TRUE
)$sccaf

sccaf_results
```

### barplot
```{r}
yval <- "score"
pl <- sccaf_results %>% 
  ggplot(aes(y = reorder(celltype, .data[[yval]]),
             x = .data[[yval]],
             fill = .data[[yval]])) +
  geom_col(show.legend = FALSE,
           color = "grey50") +
  scale_fill_gradient(
    low  = "white",
    high = "#2166AC"   # deep blue
  ) +
  labs(y = "Author's annotation",
       x = "SCCAF F1") +
  ggpubr::theme_classic2() +
  theme(axis.title = element_text(size = 22),
        axis.ticks = element_blank())

pl
```



## 5. MetaNeighbor

```{r}
library(MetaNeighbor)

meta_neighbor_results <- get_consistency(
  scTypeEval = sc,
  dissimilarity_slot = NA_character_,
  consistency_metric = "MetaNeighbor_Supervised"
)

meta_neighbor_results
```
### barplot
```{r}
yval <- "measure"
pl <- meta_neighbor_results %>% 
  ggplot(aes(y = reorder(celltype, .data[[yval]]),
             x = .data[[yval]],
             fill = .data[[yval]])) +
  geom_col(show.legend = FALSE,
           color = "grey50") +
  scale_fill_gradient(
    low  = "white",
    high = "#2166AC"   # deep blue
  ) +
  labs(y = "Author's annotation",
       x = "MetaNeighbor AUROC") +
  ggpubr::theme_classic2() +
  theme(axis.title = element_text(size = 22),
        axis.ticks = element_blank())

pl
```

