---
title: "ISC-guided refinement of stressed T cells"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# Introduction

To demonstrate the utility of ISC for identifying and resolving problematic annotations on [Chu et al., 2023](https://www.nature.com/articles/s41591-023-02371-y) which identifies T-stressed cell subsets.


# 0. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

```{r libraries}
# Load packages
library(dplyr)
library(ggplot2)
library(tidyr)
library(BiocParallel)
library(Seurat)
library(scTypeEval)
source("helpers/consistency_exploration.R")
# Custom theme
theme_set(theme_bw(base_size = 12))
```

Define which ISC paramters
```{r set_ISC}
#black list from scTypeEval package
data(black_list)
bl <- list(black_list$TCR, black_list$Immunoglobulins, black_list$Ygenes) %>% unlist()
# define used ISC
mety <- c("local" = "silhouette | recip_classif:Match",
          "global" = "2label_silhouette | Pseudobulk:Cosine")
# Define dissimilarities to use
dm <- c("Pseudobulk:Cosine",
        "recip_classif:Match")
# min samples to consider a cell type
min_samples <- 3
min_cells <- 5
```


# 1. Load dataset

```{r}
lpath <- "~/Documents/Projects/Pan-cancer_human_TILs/cache/Yost_Stressed_annotation.rds"

object <- load_single_cell_object(lpath)
object$metadata %>%
   distinct(cell.type, Celltype, Cluster,
            Cellsubtype, celltype)

ident <- "celltype" # annotation column to use
sample = "sample" # column identifying sample of origin

n_distinct(object$metadata[[sample]])
unique(object$metadata[[ident]])
table(is.na(object$metadata[[ident]]))
```


# 2. Run scTypeEval

## 2.1 Processing

```{r}
#create sceval object
sc <- create_scTypeEval(matrix = object$counts,
                        metadata = object$metadata,
                        black_list = bl)

sc <- run_processing_data(sc,
                          ident = ident,
                          sample = sample,
                          # enforce more cell type by lower threshold of samples
                          min_samples = min_samples,
                          min_cells = min_cells,
                          verbose = F)

sc <- run_hvg(sc,
              ngenes = 2000,
              ncores = 4, 
              verbose = F)

sc <- run_pca(sc,
              verbose = F)

for(mdiss in dm){
  cat("#### Running ", mdiss, " ####\n")
  sc <- run_dissimilarity(sc,
                          method = mdiss,
                          ncores = 6,
                          verbose = F)
}
```

## 2.2 Explore plots

### RMC plot
```{r}
plot_heatmap(sc, 
             dissimilarity_slot = "recip_classif:Match",
             sort_consistency = "silhouette",
             sort_similarity = "Pseudobulk:Cosine")
```
### PCA

```{r}
plot_pca(sc)
```


# 3. Consistency

```{r}
cons0 <- get_consistency(sc,
                         consistency_metric = c("silhouette",
                                                "2label_silhouette")
)

cons <- 
  cons0 %>% 
  mutate(method.type = paste(consistency_metric,
                             dissimilarity_method,
                             sep = " | ")) %>% 
  filter(method.type %in% mety) %>% 
  mutate(method.type = factor(method.type,
                              levels = mety,
                              labels = names(mety))) %>% 
  select(method.type, measure, celltype) %>% 
  pivot_wider(names_from = "method.type",
              values_from = "measure") %>% 
  mutate(measure = global * local
  ) %>% 
  filter(!is.na(measure))

consis <- cons %>%
  arrange(measure)
```

## 3.1. plotting
### barplot
```{r}
q_local <- 0.45
q_global <- 0.74
consistency_barplot(consis, x_threshold = q_local)
```
### scatter

Joint distribution of per-cell-type ISCL (silhouette on reciprocal-classification match) and ISCG (2-label silhouette on pseudobulk cosine similarity) across atlas labels (“cell_type”). Each point represents one annotated cell type. 

```{r}
consistency_scatter(consis)
```
## 4. SCCAF
```{r}
repo_root <- if (dir.exists("sample_agnostic_utils")) {
  getwd()
} else {
  normalizePath("..", mustWork = TRUE)
}
source(file.path(repo_root, "sample_agnostic_utils", "run_python_pipelines.R"))

sccaf_results <- run_sample_agnostic_python_pipelines(
  scTypeEval = sc,
  pipelines = list(
    sccaf = pipeline_spec_sccaf(cluster_key = ident, n = 100)
  ),
  file_prefix = basename(tempfile("stressed_tcells_")),
  continue_on_error = FALSE,
  cleanup = TRUE
)$sccaf

sccaf_results
```

### barplot
```{r}
consistency_barplot(sccaf_results,
                    xval = "f1_score",
                    xlabel = "SCCAF F1")
```



## 5. MetaNeighbor

```{r}
library(MetaNeighbor)

meta_neighbor_results <- get_consistency(
  scTypeEval = sc,
  dissimilarity_slot = NA_character_,
  consistency_metric = "MetaNeighbor_Supervised"
)

meta_neighbor_results
```
### barplot
```{r}
consistency_barplot(meta_neighbor_results,
                    xval = "measure",
                    x_threshold = 0.5)
```
# 6. Markers expression

## 6.1 umap

### Process for Seurat - CD4 & CD8

```{r}
filter_sctypeval <- FALSE # if false run standard Seurat workflow

if(filter_sctypeval){
  keep <- colnames(sc@data$`single-cell`@matrix)
  md <- object$metadata[keep,]
  counts <- object$counts[,keep]
  hvg <- sc@gene_lists$HVG
} else {
  md <- object$metadata
  counts <- object$counts
  hvg <- NULL
}

seu <- run_seurat_workflow(counts = counts,
                           metadata = md,
                           do.scale = T,
                           hvg = hvg)
```

```{r}
UMAPPlot(seu,
         group.by = ident,
         label = T,
         repel = T)
```
### scTypeEval

```{r}
um <- get_umap_embeddings(sc)
plot_umap_embeddings(um)
```



### CD4

#### Seurat

```{r}
md <- object$metadata %>% 
  filter(Celltype == "CD4")
counts <- object$counts[,rownames(md)]

cd4 <- run_seurat_workflow(counts = counts,
                           metadata = md,
                           do.scale = T)
```

```{r}
UMAPPlot(cd4,
         group.by = ident,
         label = T,
         repel = T)
UMAPPlot(cd4,
         group.by = sample)
```

### CD8

#### Seurat

```{r}
md <- object$metadata %>% 
  filter(Celltype == "CD8")
counts <- object$counts[,rownames(md)]

cd8 <- run_seurat_workflow(counts = counts,
                           metadata = md,
                           do.scale = T)
```

```{r}
UMAPPlot(cd8,
         group.by = ident,
         label = T,
         repel = T)
UMAPPlot(cd8,
         group.by = sample)
```


## 6.2 Gene markers

### Define gene markers

```{r}
path <- "~/Documents/Projects/Pan-cancer_human_TILs/cache"
chu.markers <- list()
chu.markers[["CD4"]] <- read.table(file.path(path, "Chu_cd4_markers.txt"))
chu.markers[["CD8"]] <- read.table(file.path(path, "Chu_cd8_markers.txt"))

str.genes <- c("HSPA1A","HSPA1B", "BAG3", "HSPE1", "HSPB1", "HSPD1", "NR4A1", "NFKB1")
```


```{r}
markers <- list(
    "CD4.Treg"   = c("FOXP3", "CTLA4", "IL2RA", "TNFRSF4"),
    "CD8.Tex"    = c("TOX", "LAG3", "PRDM1", "CD27"),
    "CD4.Th17"   = c("RORC", "RORA", "CCR6", "IL17A"),
    "CD8.t.Teff" = c("GZMH", "FGFBP2", "PRF1", "CD69"),
    "CD4.Tcm"    = c("IL7R", "GPR183", "CD69", "TCF7"),
    "CD8.Tstr"   = str.genes,
    "CD8.Tn"     = c("CCR7", "TCF7", "SELL"),
    "CD8.p-Tex"  = c("PDCD1", "TIGIT", "TOX", "TCF7"),
    "CD8.Teff"   = c("NKG7", "GNLY", "GZMB"),
    "CD4.Tfh"    = c("CXCR5", "ICOS", "BCL6", "PDCD1"),
    "CD4.Tctl"   = c("GZMA", "GZMB", "GNLY", "NKG7", "FGFBP2", "PRF1"),
    "CD4.Tn"     = c("LEF1", "CCR7", "SELL", "IL7R"),
    "CD8.Tisg"   = c("ISG15", "IFIT1", "MX1"),
    "CD4.Tstr"   = str.genes,
    "CD8.Tcm"    = c("CCR7", "TCF7", "SELL", "LEF1", "NOSIP")
)
```


### Heatmap
Get pseudobulk expression per sample and cell type
```{r}
isc_order <- rev(consis$celltype)

genes <- markers[isc_order] %>%
  unlist(use.names = F)
```

```{r}
ps <- get_pseudobulk(sc,
                     order = isc_order,
                     genes = genes)
```

Make heatmap

```{r}
phall <- do_heatmap(ps)
```

### Per CD4 & CD8

```{r}
genes <- markers[grepl("CD8", names(markers))] %>%
  unlist(use.names = F)
ps_cd8 <- get_pseudobulk(sc,
                     order = isc_order,
                     genes = genes,
                     filter = c("Celltype" = "CD8"))
ph_cd8 <- do_heatmap(ps_cd8,
                     cap = 4)
```

```{r}
genes <- markers[grepl("CD4", names(markers))] %>%
  unlist(use.names = F)
ps_cd4 <- get_pseudobulk(sc,
                     order = isc_order,
                     genes = genes,
                     filter = c("Celltype" = "CD4"))
ph_cd4 <- do_heatmap(ps_cd4,
                     cap = 4,
                     cluster_rows = F)
```

## 6.3 UCell scoring

### CD4
```{r}
library(UCell)
sigs <- markers[grepl("CD4", names(markers))] 
cd4 <- AddModuleScore_UCell(cd4,
                            features = sigs,
                            ncores = 7,
                            name = "_score")
```

```{r}
bp_do <- c("CD4.Tn","CD4.Tctl",
           "CD4.Treg", "CD4.Tcm",
           "CD4.Tstr") %>% 
  sort()
pat <- paste(bp_do, collapse = "|") %>% 
  gsub("[.]", "[.]", .)
bp_cd4 <- lapply(bp_do,
                 function(c){
                   sample_boxplot_expression(cd4,
                                             signature_pattern = pat,
                                             ident = ident,
                                             ident_sel = c,
                                             sample = "sample")
                 })

bp_cd4_gg <- ggpubr::ggarrange(plotlist = bp_cd4,
                               nrow = 1)
bp_cd4_gg
```

### CD8
```{r}
library(UCell)
sigs <- markers[grepl("CD8", names(markers))] 
cd8 <- AddModuleScore_UCell(cd8,
                            features = sigs,
                            ncores = 7,
                            name = "_score")
```

```{r}
bp_do <- c("CD8.Tn","CD8.Teff",
           "CD8.Tex", "CD8.Tcm",
           "CD8.Tstr", "CD8.p-Tex") %>% 
  sort()
pat <- paste(bp_do, collapse = "|") %>% 
  gsub("[.]", "[.]", .)
bp_cd8 <- lapply(bp_do,
                 function(c){
                   sample_boxplot_expression(cd8,
                                             signature_pattern = pat,
                                             ident = ident,
                                             ident_sel = c,
                                             sample = "sample")
                 })

bp_cd8_gg <- ggpubr::ggarrange(plotlist = bp_cd8,
                               nrow = 1)
bp_cd8_gg
```



# 7. Self-projection

Project using ProjecTILls each cell type

## CD4

```{r}
sp_do <- c("CD4.Tn","CD4.Tctl",
           "CD4.Treg", "CD4.Tcm",
           "CD4.Tstr")

cd4_pls <- list()

for(c in sp_do){
  if(c %in% names(cd4_pls)){next}
  cat("## Doing ", c, " ## \n ---------------------- \n")
  cd4_pls[[c]] <- self_project_pt(seu = cd4,
                  heldout = c)
}
  
```

# 8. Reannotation

Try firstly only with CD4+ T cells
```{r}
md <- object$metadata %>% 
  filter(Celltype == "CD4")
counts <- object$counts[,rownames(md)]

cd4 <- CreateSeuratObject(counts = counts,
                          meta.data = md)
```


```{r}
sp_do <- c(
           "CD4.Th17",
           "CD4.Treg",
           "CD4.Tstr")

reann <- list()

for(c in sp_do){
  if(c %in% names(reann)){next}
  cat("## Doing ", c, " ## \n ---------------------- \n")
  reann[[c]] <- reannotate_singleR(seu = cd4,
                                   heldout = c,
                                   ncores = 8)
}

```


## Inject new annotation

```{r}
meta <- md %>% 
  tibble::rownames_to_column("cellid")

for(c in names(reann)){
  nc <- paste0("reannot_", c)
  snout <- reann[[c]] %>% 
    select(pruned.labels) %>% 
    tibble::rownames_to_column("cellid")
  
  meta <- meta %>% 
    left_join(., snout, by = "cellid")
  meta[[nc]] <- ifelse(is.na(meta$pruned.labels),
                       meta[[ident]],
                       meta$pruned.labels)
  meta[["pruned.labels"]] <- NULL
}

meta <- meta %>% 
  tibble::column_to_rownames("cellid")
```

## ISC

```{r}
isc_idents <- c(ident, grep("reannot_", names(meta),value = T))
cc <- counts[,rownames(meta)]
scs <- lapply(isc_idents,
              function(i){
                wrapper_scTypeEval(count_matrix = cc,
                                   metadata = meta,
                                   ident = i,
                                   sample = sample,
                                   dissimilarity_method = dm,
                                   ncores = 6,
                                   min_samples = min_samples,
                                   min_cells = min_cells)
              })
names(scs) <- isc_idents
```
#### RCM
```{r}
lapply(scs, function(x){
  plot_heatmap(x, 
             dissimilarity_slot = "recip_classif:Match",
             sort_consistency = "silhouette",
             sort_similarity = "Pseudobulk:Cosine")
})
```

#### ISC-Scatter

```{r}
cons <- lapply(names(scs), function(x){
  cons0 <- get_consistency(scs[[x]],
                           consistency_metric = c("silhouette",
                                                  "2label_silhouette")
  )
  
  cons <- 
    cons0 %>% 
    mutate(method.type = paste(consistency_metric,
                               dissimilarity_method,
                               sep = " | ")) %>% 
    filter(method.type %in% mety) %>% 
    mutate(method.type = factor(method.type,
                                levels = mety,
                                labels = names(mety))) %>% 
    select(method.type, measure, celltype) %>% 
    pivot_wider(names_from = "method.type",
                values_from = "measure") %>% 
    mutate(measure = global * local
    ) %>% 
    filter(!is.na(measure))
  
  consis <- cons %>%
    arrange(measure) %>% 
    mutate(annot = x)
  
  return(consis)
})
names(cons) <- names(scs)
```

```{r}
lapply(names(cons), function(x){
  consistency_scatter(cons[[x]]) +
    labs(title = x)
})
```


### Annotation improvement

```{r}
allcons <- cons %>% 
  data.table::rbindlist()
```

```{r}
yval <- "local"
allcons %>% 
  ggplot(aes(annot, .data[[yval]])) +
  geom_boxplot(
    outlier.color = NA,
    width = 0.3
  ) +
  geom_point(aes(color = celltype),
                 show.legend = F
  ) + 
  ggrepel::geom_text_repel(aes(label = celltype,
                               color = celltype),
                           show.legend = F) + 
  ggpubr::theme_classic2()
```



