---
title: "Supplementary Figure 2A: Internal saturation analysis"
author: "J. Garnica <josep.garnicacaparros at unige.ch>"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

# Introduction

This notebook renders Supplementary Figure 2A from precomputed results produced by the `ISC_benchmark_saturation` pipeline. The notebook does not recompute the saturation analysis locally. Instead, it loads the aggregated pipeline output, summarizes rank-correlation stability across group sizes, and fits a GAM to model how variability changes as more dataset-annotation combinations are included.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 7
)
```

```{r libraries}
library(dplyr)
library(ggplot2)
library(mgcv)
library(scales)
```

```{r exporting_dir}
fig_dir <- "exported_figs/Supp_2A_Figure"
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

# Load saturation outputs

The saturation analysis is generated in `ISC_benchmark_saturation` and saved under `results/isc_benchmark_saturation/`. The aggregated results file is used directly for both plotting and GAM fitting.

```{r load_results}
results_file <- file.path("..", "results", "isc_benchmark_saturation", "ident_saturation_results.rds")
input_file <- file.path("..", "results", "isc_benchmark_saturation", "ident_input_concatenated.rds")

if (!file.exists(results_file)) {
  stop(
    "Missing saturation results file: ", results_file,
    "\nRun the ISC_benchmark_saturation pipeline first."
  )
}

final_df <- readRDS(results_file)

if (!all(c("GroupSize", "Correlation") %in% colnames(final_df))) {
  stop("Saturation results file does not contain the expected columns: GroupSize, Correlation")
}

if (file.exists(input_file)) {
  saturation_input <- readRDS(input_file)
}
```

# Summaries

```{r summarize_results}
summary_df <- final_df %>%
  filter(!is.na(GroupSize), !is.na(Correlation)) %>%
  group_by(GroupSize) %>%
  summarise(
    MedianCorrelation = median(Correlation, na.rm = TRUE),
    IQR = IQR(Correlation, na.rm = TRUE),
    N = dplyr::n(),
    .groups = "drop"
  ) %>%
  arrange(GroupSize)

summary_df
```

# GAM model

```{r gam_model}
gam_model <- mgcv::gam(IQR ~ s(GroupSize, bs = "cs", k = 10), data = summary_df)

pred_grid <- data.frame(GroupSize = seq(min(summary_df$GroupSize), 29, by = 1))
pred_grid$PredictedIQR <- predict(gam_model, newdata = pred_grid)
pred_grid$PredictedIQR <- pmax(pred_grid$PredictedIQR, 0)

pred_grid
```

# Plots

## Distribution of rank correlations by group size

```{r violin_plot}
rank_plot <- final_df %>%
  filter(!is.na(GroupSize), !is.na(Correlation)) %>%
  mutate(GroupSize = factor(GroupSize, levels = sort(unique(GroupSize)))) %>%
  ggplot(aes(GroupSize, Correlation)) +
  geom_violin(fill = "grey90", color = "grey40") +
  stat_summary(
    geom = "crossbar",
    width = 0.2,
    linewidth = 0.4,
    color = "firebrick",
    fun = median,
    aes(group = GroupSize)
  ) +
  labs(
    x = "Number of dataset-annotation combinations",
    y = "Spearman rank correlation"
  ) +
  theme_classic()

rank_plot

ggsave(
  file.path(fig_dir, "Supp2A_rank_correlation_violin.pdf"),
  rank_plot,
  width = 9,
  height = 7
)
```

## Median correlation and variability across group sizes

```{r median_iqr_plot}
p1 <- ggplot(summary_df, aes(x = GroupSize, y = MedianCorrelation)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_ribbon(
    aes(
      ymin = MedianCorrelation - IQR / 2,
      ymax = MedianCorrelation + IQR / 2
    ),
    fill = "steelblue",
    alpha = 0.2
  ) +
  labs(
    title = "Stability of method rankings by group size",
    x = "Group size",
    y = "Median Spearman correlation ± IQR/2"
  ) +
  theme_classic()

p1

ggsave(
  file.path(fig_dir, "Supp2A_median_correlation_trend.pdf"),
  p1,
  width = 9,
  height = 7
)
```

## GAM fit for IQR decay

```{r gam_fit_plot}
p2 <- ggplot(summary_df, aes(x = GroupSize, y = IQR)) +
  geom_point(size = 2.5, color = "steelblue") +
  geom_line(data = pred_grid, aes(x = GroupSize, y = PredictedIQR), color = "firebrick", linewidth = 1) +
  geom_point(
    data = pred_grid %>% filter(GroupSize > max(summary_df$GroupSize)),
    aes(x = GroupSize, y = PredictedIQR),
    shape = 1,
    color = "firebrick",
    size = 2
  ) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "GAM fit of rank-correlation variability across group sizes",
    x = "Group size",
    y = "IQR of Spearman correlation"
  ) +
  theme_classic()

p2

ggsave(
  file.path(fig_dir, "Supp2A_gam_iqr_fit.pdf"),
  p2,
  width = 9,
  height = 7
)
```

# Methods

For each group size, the `ISC_benchmark_saturation` pipeline samples disjoint pairs of dataset subsets, computes weighted method rankings within each subset, and measures agreement between rankings using Spearman correlation. The aggregated output stored in `ident_saturation_results.rds` was used here directly. To quantify how ranking variability changes with group size, we summarized the interquartile range of the correlation distribution at each group size and fitted a generalized additive model using `mgcv`.

# Results

The dispersion of ranking correlations decreases as group size increases, indicating that benchmark rankings become more stable when more dataset-annotation combinations are included. The GAM fit captures this decay in variability and shows an apparent plateau, consistent with diminishing returns beyond intermediate group sizes.